---
title: "Comparing SoupX and DecontX ambient RNA correction"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Introduction

This notebook compares the ambient-RNA corrections produced by [SoupX](soupx.html) and [DecontX](decontx.html) on the same `SC3pv3_GEX_Human_PBMC` dataset. Both source notebooks save their corrected count matrix and per-cell contamination estimate to `output/`; we load both here and compare them at the per-cell, per-gene, and per-(cell × gene) level.

To recap the two methods briefly:

* **SoupX** estimates the soup profile directly from the empty droplets in the raw 10x output and then estimates a per-cell contamination fraction `rho` by looking for genes that should be soup-only in particular clusters.
* **DecontX** treats the contamination distribution for each cell as a weighted combination of the other clusters' profiles. The version run in the companion notebook additionally uses the raw matrix as an empirical anchor for the ambient distribution.

Both methods produce a corrected count matrix of the same shape as the input filtered matrix, plus a per-cell contamination fraction in `[0, 1]`. We can therefore line up the two outputs cell-by-cell and gene-by-gene to see where they agree and where they diverge.

## Packages

```{r packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(Matrix)
  library(Seurat)
  library(ggplot2)
})
```

## Loading the inputs

We need three things: the original filtered count matrix (so we can compute the per-gene and per-cell totals removed by each method), the SoupX corrected matrix, and the DecontX corrected matrix.

```{r load}
original <- Seurat::Read10X_h5("data/SC3pv3_GEX_Human_PBMC_filtered_feature_bc_matrix.h5")

soupx   <- readRDS("output/soupx_corrected.rds")
decontx <- readRDS("output/decontx_corrected.rds")
```

The two corrected matrices should cover the same set of cells, since both notebooks start from the same Cell Ranger filtered output:

```{r dims}
dim(original)
dim(soupx$counts)
dim(decontx$counts)
```

To make per-cell comparisons we re-order both corrected matrices and their metadata to match the column order of the original filtered matrix.

```{r align}
common_cells <- Reduce(
  intersect,
  list(colnames(original), colnames(soupx$counts), colnames(decontx$counts))
)
length(common_cells)

original       <- original[, common_cells]
soupx$counts   <- soupx$counts[, common_cells]
decontx$counts <- decontx$counts[, common_cells]
soupx$meta     <- soupx$meta[common_cells, ]
decontx$meta   <- decontx$meta[common_cells, ]
```

## Per-cell contamination fraction

The most direct comparison is the per-cell contamination fraction. SoupX estimates this via `autoEstCont()` and stores it in `sc$metaData$rho`; DecontX produces it as `colData(sce)$decontX_contamination`. Both are bounded in `[0, 1]`.

```{r rho_summary}
summary(soupx$meta$rho)
summary(decontx$meta$rho)
```

```{r rho_hist, fig.width=8, fig.height=4}
rho_long <- rbind(
  data.frame(method = "SoupX",   rho = soupx$meta$rho),
  data.frame(method = "DecontX", rho = decontx$meta$rho)
)

ggplot(rho_long, aes(rho, fill = method)) +
  geom_histogram(bins = 60, alpha = 0.6, position = "identity") +
  labs(x = "Estimated contamination fraction", y = "Cells") +
  theme_minimal()
```

A scatter plot of the two estimates per cell shows whether the methods agree on which cells are most contaminated. A diagonal trend means they agree on the relative ranking; the absolute level often differs because the two estimators use different priors and different anchors for the ambient distribution.

```{r rho_scatter, fig.width=5, fig.height=5}
ggplot(
  data.frame(soupx = soupx$meta$rho, decontx = decontx$meta$rho),
  aes(soupx, decontx)
) +
  geom_point(size = 0.3, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "SoupX rho", y = "DecontX contamination") +
  theme_minimal()

cor(soupx$meta$rho, decontx$meta$rho, method = "spearman")
```

## Per-cell UMIs removed

The estimated `rho` is what each model believes; the UMIs actually subtracted from each cell are what the model produced. These two are not exactly equal because both `adjustCounts()` and `decontXcounts()` distribute the removed counts across genes in a constrained way. Comparing per-cell UMIs removed is therefore a useful complement to the `rho` comparison.

```{r umi_removed, fig.width=5, fig.height=5}
total_orig    <- Matrix::colSums(original)
removed_soup  <- total_orig - Matrix::colSums(soupx$counts)
removed_decon <- total_orig - Matrix::colSums(decontx$counts)

ggplot(
  data.frame(soupx = removed_soup, decontx = removed_decon),
  aes(soupx, decontx)
) +
  geom_point(size = 0.3, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  labs(x = "SoupX UMIs removed (per cell)",
       y = "DecontX UMIs removed (per cell)") +
  theme_minimal()
```

## Per-gene removal

Summing the corrections across cells gives a per-gene total of how many UMIs each method assigned to soup. The genes at the top of these lists are the methods' best guesses at the most soup-contaminated genes in the dataset.

```{r gene_totals}
gene_totals <- data.frame(
  gene          = rownames(original),
  before        = Matrix::rowSums(original),
  after_soupx   = Matrix::rowSums(soupx$counts),
  after_decontx = Matrix::rowSums(decontx$counts)
)
gene_totals$removed_soupx   <- gene_totals$before - gene_totals$after_soupx
gene_totals$removed_decontx <- gene_totals$before - gene_totals$after_decontx
```

The 20 genes with the largest counts removed by each method:

```{r top_removed_soupx}
head(
  gene_totals[order(-gene_totals$removed_soupx),
              c("gene", "before", "removed_soupx", "removed_decontx")],
  20
)
```

```{r top_removed_decontx}
head(
  gene_totals[order(-gene_totals$removed_decontx),
              c("gene", "before", "removed_soupx", "removed_decontx")],
  20
)
```

A scatter plot at the gene level — log-scaled because a handful of soup genes dominate the totals — shows whether the two methods agree on per-gene removal:

```{r gene_scatter, fig.width=5, fig.height=5}
ggplot(
  gene_totals,
  aes(removed_soupx + 1, removed_decontx + 1)
) +
  geom_point(size = 0.3, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "SoupX UMIs removed per gene (+1)",
       y = "DecontX UMIs removed per gene (+1)") +
  theme_minimal()
```

## Marker genes

The classical ambient-RNA markers in PBMC preps are the haemoglobin genes (`HBB`, `HBA1`, `HBA2`) — released by lysed red blood cells — and immunoglobulin genes (`IGKC`, `IGLC2`, `IGHA1`) — released by lysed plasma cells. Both methods should remove the haemoglobin counts almost entirely; the immunoglobulin counts should be preserved in the B / plasma cell cluster but removed from elsewhere.

```{r markers_total}
markers <- intersect(
  c("HBB", "HBA1", "HBA2", "IGKC", "IGLC2", "IGHA1"),
  rownames(original)
)

data.frame(
  gene            = markers,
  before          = Matrix::rowSums(original[markers, ]),
  after_soupx     = round(Matrix::rowSums(soupx$counts[markers, ])),
  after_decontx   = round(Matrix::rowSums(decontx$counts[markers, ])),
  removed_soupx   = round(Matrix::rowSums(original[markers, ]) -
                            Matrix::rowSums(soupx$counts[markers, ])),
  removed_decontx = round(Matrix::rowSums(original[markers, ]) -
                            Matrix::rowSums(decontx$counts[markers, ]))
)
```

## Agreement at the (cell × gene) level

Finally, we ask how similar the two corrected matrices are at the (cell × gene) level. For each cell we compute the Pearson correlation between the SoupX-corrected and DecontX-corrected expression vectors. A correlation close to 1 means the two methods broadly agree on which counts to remove from that cell; a correlation noticeably below 1 means they disagree on the per-cell distribution of corrections.

We sample 2000 cells to keep the computation cheap.

```{r per_cell_correlation, fig.width=6, fig.height=4}
set.seed(1984)
sample_cells <- sample(common_cells, min(2000, length(common_cells)))

cell_cor <- vapply(sample_cells, function(b) {
  cor(as.numeric(soupx$counts[, b]), as.numeric(decontx$counts[, b]))
}, numeric(1))

ggplot(data.frame(r = cell_cor), aes(r)) +
  geom_histogram(bins = 50, fill = "steelblue") +
  labs(x = "Per-cell Pearson correlation (SoupX vs DecontX)",
       y = "Cells") +
  theme_minimal()

summary(cell_cor)
```

## Summary

* The two methods typically agree on the *identity* of the most-contaminated genes — haemoglobin and immunoglobulin transcripts dominate both top-removed lists for PBMC preps.
* They agree less precisely on the *magnitude* of the per-cell contamination fraction, because they use different priors and different anchors for the ambient distribution.
* For routine PBMC analysis the choice between SoupX and DecontX is rarely critical; for sorted samples or unusual tissue preparations where the empty-droplet population may not represent the full ambient pool, the assumptions matter more.

## Session info

```{r session_info}
sessionInfo()
```
