---
title: "Using the anndataR package"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE)
```

The [anndataR package](https://bioconductor.org/packages/release/bioc/html/anndataR.html) brings the [AnnData](https://anndata.readthedocs.io/) data structure into R:

> anndataR provides a native R implementation of the AnnData data model, enabling R users to read, write, manipulate, and convert .h5ad files without requiring Python dependencies for core operations.

The AnnData format is the standard container for single-cell genomics data in the Python/scverse ecosystem (used by scanpy, scvi-tools, etc.). anndataR bridges the gap between R and Python single-cell workflows by providing bidirectional conversion with both SingleCellExperiment and Seurat objects.

## Installation

Install from Bioconductor.

```{r install_anndatar, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("anndataR")
```

Optional dependencies for reading/writing `.h5ad` files and converting to other formats.

```{r install_optional, eval=FALSE}
BiocManager::install("rhdf5")
BiocManager::install("SingleCellExperiment")
install.packages("Seurat")
install.packages("SeuratObject")
```

## Package

Load package.

```{r load_package}
packageVersion("anndataR")
suppressPackageStartupMessages(library(anndataR))
```

## The AnnData data model

AnnData stores single-cell data in a structured format with nine slots:

| Slot | Description |
|------|-------------|
| `X` | Primary expression matrix (observations x variables, i.e. cells x genes) |
| `obs` | Observation (cell) metadata as a data.frame |
| `var` | Variable (gene) metadata as a data.frame |
| `obs_names` | Character vector of cell identifiers |
| `var_names` | Character vector of gene identifiers |
| `layers` | Named list of alternative matrices (same dimensions as X) |
| `obsm` | Named list of multi-dimensional observation annotations (e.g. embeddings) |
| `varm` | Named list of multi-dimensional variable annotations (e.g. loadings) |
| `obsp` | Named list of pairwise observation matrices (e.g. cell-cell distances) |
| `varp` | Named list of pairwise variable matrices |
| `uns` | Arbitrary unstructured metadata |

A key difference from R conventions: AnnData stores matrices as **observations x variables** (cells x genes), while SingleCellExperiment and Seurat store them as **variables x observations** (genes x cells). anndataR handles this transposition automatically during conversions.

## Creating an AnnData object

Use the `AnnData()` constructor to create an in-memory AnnData object.

```{r create_anndata}
n_obs <- 100
n_vars <- 50

set.seed(1984)
counts <- matrix(rpois(n_obs * n_vars, lambda = 5), nrow = n_obs)
rownames(counts) <- paste0("cell_", seq_len(n_obs))
colnames(counts) <- paste0("gene_", seq_len(n_vars))

adata <- AnnData(
  X = counts,
  obs = data.frame(
    row.names = paste0("cell_", seq_len(n_obs)),
    cell_type = factor(sample(c("T cell", "B cell", "Monocyte"), n_obs, replace = TRUE)),
    total_counts = rowSums(counts),
    n_genes = rowSums(counts > 0)
  ),
  var = data.frame(
    row.names = paste0("gene_", seq_len(n_vars)),
    gene_name = paste0("Gene", seq_len(n_vars)),
    highly_variable = sample(c(TRUE, FALSE), n_vars, replace = TRUE)
  )
)

adata
```

## Exploring the object

Access the dimensions and slot keys.

```{r explore_dims}
dim(adata)
adata$n_obs
adata$n_vars
```

Observation names and variable names.

```{r explore_names}
head(adata$obs_names)
head(adata$var_names)
```

Cell metadata stored in `obs`.

```{r explore_obs}
head(adata$obs)
table(adata$obs$cell_type)
```

Gene metadata stored in `var`.

```{r explore_var}
head(adata$var)
sum(adata$var$highly_variable)
```

The expression matrix stored in `X` has cells as rows and genes as columns.

```{r explore_x}
dim(adata$X)
adata$X[1:5, 1:5]
```

## Adding layers

Layers store alternative representations of the data with the same dimensions as `X`. A common use case is storing raw counts alongside normalised values.

```{r add_layers}
log_norm <- log1p(sweep(counts, 1, rowSums(counts), "/") * 1e4)

adata$layers <- list(
  log_norm = log_norm
)

adata$layers_keys
adata$layers[["log_norm"]][1:5, 1:5]
```

## Adding embeddings

Dimensionality reductions are stored in `obsm` (one entry per cell). Let's compute a simple PCA and store it.

```{r add_embeddings}
pca_result <- prcomp(log_norm, center = TRUE, scale. = TRUE, rank. = 20)

adata$obsm <- list(
  X_pca = pca_result$x
)

adata$obsm_keys
dim(adata$obsm[["X_pca"]])
```

Visualise the first two principal components.

```{r plot_pca}
pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  cell_type = adata$obs$cell_type
)

plot(
  pca_df$PC1, pca_df$PC2,
  col = as.integer(pca_df$cell_type),
  pch = 16,
  xlab = "PC1",
  ylab = "PC2",
  main = "PCA of simulated data"
)
legend("topright", levels(pca_df$cell_type), col = seq_along(levels(pca_df$cell_type)), pch = 16)
```

Since the data is randomly generated, we do not expect the cell types to separate.

## Adding variable loadings

Gene loadings from PCA can be stored in `varm`.

```{r add_varm}
adata$varm <- list(
  PCs = pca_result$rotation
)

adata$varm_keys
dim(adata$varm[["PCs"]])
```

## Unstructured metadata

The `uns` slot stores arbitrary metadata such as colour palettes, analysis parameters, or summary statistics.

```{r add_uns}
adata$uns <- list(
  analysis_date = Sys.Date(),
  cell_type_colours = c("T cell" = "steelblue", "B cell" = "tomato", "Monocyte" = "forestgreen"),
  pca_variance = summary(pca_result)$importance[2, 1:5]
)

adata$uns_keys
adata$uns[["cell_type_colours"]]
```

## Subsetting

AnnData objects support subsetting with `[` using logical, numeric, or character indices. Subsetting creates a **view** that references the parent object without copying data.

```{r subset_logical}
t_cells <- adata[adata$obs$cell_type == "T cell", ]
t_cells
```

```{r subset_numeric}
small <- adata[1:10, 1:5]
dim(small)
small$X
```

```{r subset_character}
selected <- adata[c("cell_1", "cell_2"), c("gene_1", "gene_10", "gene_50")]
dim(selected)
selected$obs
```

Subset to highly variable genes.

```{r subset_hv_genes}
hv_genes <- adata$var_names[adata$var$highly_variable]
adata_hv <- adata[, hv_genes]
dim(adata_hv)
```

## Reading h5ad files

anndataR reads (and writes) `.h5ad` files using Bioconductor's {rhdf5} package natively, without requiring Python.

```{r read_h5ad}
pbmc3k <- read_h5ad("data/pbmc3k.h5ad")
pbmc3k
```

## Converting to SingleCellExperiment

anndataR provides direct conversion to [SingleCellExperiment](https://bioconductor.org/packages/SingleCellExperiment/) objects, which are widely used in Bioconductor single-cell workflows.

```{r to_sce}
suppressPackageStartupMessages(library(SingleCellExperiment))

sce <- pbmc3k$as_SingleCellExperiment()
sce
```

Note that the matrix is transposed: SingleCellExperiment stores genes as rows and cells as columns.

```{r sce_dims}
dim(sce)
assayNames(sce)
head(colData(sce))
```

## Converting to Seurat object

```{r to_seurat}
suppressPackageStartupMessages(library(Seurat))

seurat_obj <- pbmc3k$as_Seurat()
seurat_obj
```

## Summary

anndataR provides a native R implementation of the AnnData data model that:

* Reads and writes `.h5ad` files without Python via {rhdf5}
* Converts bidirectionally with SingleCellExperiment and Seurat
* Supports in-memory, HDF5-backed, and reticulate-based backends
* Creates memory-efficient views when subsetting
* Replaces older packages (`anndata`, `zellkonverter`, `h5ad`) with a single unified solution

This makes it straightforward to share single-cell datasets between R and Python workflows.

## Further reading

* [anndataR documentation](https://anndatar.data-intuitive.com/)
* [Getting started vignette](https://anndatar.data-intuitive.com/articles/getting-started.html)
* [SingleCellExperiment conversion](https://anndatar.data-intuitive.com/articles/usage_singlecellexperiment.html)
* [Seurat conversion](https://anndatar.data-intuitive.com/articles/usage_seurat.html)
* [GitHub repository](https://github.com/scverse/anndataR)
