---
title: "Checking out Azimuth"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r load_libraries, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(Azimuth)
})
```

## Background

Azimuth provides a reference and a query cell is placed individually on this reference, and then labelled by a weighted vote using the reference cells it lands near.

## Installation

Install as per the [official instructions](https://github.com/satijalab/azimuth#installation).

```{r install_azimuth, eval=FALSE}
if (!requireNamespace('remotes', quietly = TRUE)) {
  install.packages('remotes')
}
remotes::install_github('satijalab/azimuth', ref = 'master')
```

## Azimuth reference

We will examine the [Human PBMC](https://zenodo.org/records/4546839) reference that has been downloaded using [zenodo_get](https://davetang.org/muse/2024/04/12/downloading-data-from-zenodo-using-zenodo_get/).

```
zenodo_get -r 4546839
```

An Azimuth model holds two files:

* `ref.Rds`, a Seurat object carrying the reference cells, their labels and a **precomputed** dimensional reduction
* `idx.annoy`, an approximate nearest neighbor index over that reduction.

```{r azimuth_model}
azimuth_model <- paste0("data/azimuth_pbmc/")
stopifnot(dir.exists(azimuth_model))
```

`RunAzimuth()` calls `Azimuth:::LoadReference()` internally; we will call the same function directly. It accepts a local directory or a URL, and returns a list of two Seurat objects:

* `$map` used for the annotation, and
* `$plot` a lighter copy used only for drawing the reference UMAP.

```{r azimuth_obj, warning=FALSE, message=FALSE}
azimuth_obj <- Azimuth:::LoadReference(path = azimuth_model)
azimuth_obj
```

`refAssay` is the default assay.

```{r ref_assay_name}
ref_assay_name <- SeuratObject::DefaultAssay(azimuth_obj$map)
ref_assay_name
```

We already saw above that there are 5,000 features.

```{r all_features}
all_features <- rownames(azimuth_obj$map)
length(all_features)
```

`refDR` is the reduction the reference was built with. Its feature loadings define the coordinate system query cells are projected into, so its rownames are the genes that can influence where a query cell lands.

```{r refdr}
refdr_loadings <- SeuratObject::Loadings(azimuth_obj$map[["refDR"]])
refdr_features <- rownames(refdr_loadings)
dim(refdr_loadings)
```

`SCTAssay`.

```{r assay_obj}
assay_obj <- azimuth_obj$map[[ref_assay_name]]
assay_obj
all(row.names(assay_obj) == all_features)
```

The labels Azimuth can transfer are simply the reference's metadata columns.

```{r celltypes}
head(azimuth_obj$map@meta.data)
```

Neighbour names.

```{r neighbour_names}
neighbour_names <- names(methods::slot(azimuth_obj$map, "neighbors"))
neighbour_names
```
