---
title: "Checking out Azimuth"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r load_libraries, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(Azimuth)
})
```

## Background

Azimuth provides a reference and a query cell is placed individually on this reference, and then labelled by a weighted vote using the reference cells it lands near.

1. **Load the reference**, a Seurat object of already annotated cells trimmed down to the 5,000 genes it was built from, together with the coordinate system those genes define and a prebuilt nearest neighbour index over it.
2. **Check the gene identifiers** so that query and reference use the same species and the same flavour of gene name.
3. **Normalise the query using the reference's model.** The reference ships an SCTransform model: a per-gene regularised negative binomial regression of counts on sequencing depth. The query's own counts are turned into Pearson residuals using the reference's coefficients, i.e. nothing is refit on the query, so a residual means the same thing in both datasets.
4. **Project the query into the reference's space** by multiplying those residuals by the reference's feature loadings. No dimensional reduction is computed on the query; it borrows the reference's.
5. **Find anchors**, pairs of query and reference cells that are mutual nearest neighbours in that shared space. These are the links along which everything else is carried.
6. **Transfer the labels**: each query cell's label is a distance-weighted vote over the reference cells it is anchored to, producing a score for every possible label, with the highest scoring one becoming the prediction.
7. **Place the query on the reference UMAP** using the reference's saved UMAP model, so query and reference cells can be drawn on the same axes.
8. **Score the mapping** to flag query cells that the reference describes poorly, which matters because every query cell is given a label whether or not the reference contains a sensible one for it.

## Installation

Install as per the [official instructions](https://github.com/satijalab/azimuth#installation).

```{r install_azimuth, eval=FALSE}
if (!requireNamespace('remotes', quietly = TRUE)) {
  install.packages('remotes')
}
remotes::install_github('satijalab/azimuth', ref = 'master')
```

## Azimuth reference

We will examine the [Human PBMC](https://zenodo.org/records/4546839) reference that has been downloaded using [zenodo_get](https://davetang.org/muse/2024/04/12/downloading-data-from-zenodo-using-zenodo_get/).

```
zenodo_get -r 4546839
```

An Azimuth model holds two files:

* `ref.Rds`, a Seurat object carrying the reference cells, their labels and a **precomputed** dimensional reduction
* `idx.annoy`, an approximate nearest neighbor index over that reduction.

```{r azimuth_model}
azimuth_model <- paste0("data/azimuth_pbmc/")
stopifnot(dir.exists(azimuth_model))
```

`RunAzimuth()` calls `Azimuth:::LoadReference()` internally; we will call the same function directly. It accepts a local directory or a URL, and returns a list of two Seurat objects:

* `$map` used for the annotation, and
* `$plot` a lighter copy used only for drawing the reference UMAP.

```{r azimuth_obj, warning=FALSE, message=FALSE}
azimuth_obj <- Azimuth:::LoadReference(path = azimuth_model)
azimuth_obj
```

`refAssay` is the default assay.

```{r ref_assay_name}
ref_assay_name <- SeuratObject::DefaultAssay(azimuth_obj$map)
ref_assay_name
```

We already saw above that there are 5,000 features.

```{r all_features}
all_features <- rownames(azimuth_obj$map)
length(all_features)
```

`refDR` is the reduction the reference was built with. Its feature loadings define the coordinate system query cells are projected into, so its rownames are the genes that can influence where a query cell lands.

```{r refdr}
refdr_loadings <- SeuratObject::Loadings(azimuth_obj$map[["refDR"]])
refdr_features <- rownames(refdr_loadings)
dim(refdr_loadings)
```

`SCTAssay`.

```{r assay_obj}
assay_obj <- azimuth_obj$map[[ref_assay_name]]
assay_obj
all(row.names(assay_obj) == all_features)
```

The labels Azimuth can transfer are simply the reference's metadata columns.

```{r celltypes}
head(azimuth_obj$map@meta.data)
```

Neighbour names.

```{r neighbour_names}
neighbour_names <- names(methods::slot(azimuth_obj$map, "neighbors"))
neighbour_names
```

## pbmc3k

Directly import from [my server](https://davetang.github.io/muse/pbmc3k.html).

```{r pbmc3k}
pbmc3k <- readRDS(url("https://davetang.org/file/pbmc3k_seurat.rds", "rb"))
pbmc3k
```

`RunAzimuth()` is a wrapper around the entire mapping pipeline, so a single call does everything from loading the reference to writing the predicted labels back into the query object. The messages it prints are a running commentary of those steps.

```{r run_azimuth}
pbmc3k_anno <- Azimuth::RunAzimuth(query = pbmc3k, reference = azimuth_model)
pbmc3k_anno
```

* The first three warnings come from `LoadReference()`, the function we called ourselves above.
* `Overwriting miscellanous data for model` is bookkeeping as the stored SCT model is attached to `refAssay`
* `Adding a dimensional reduction (refUMAP) without the associated assay being present` appears once for `$map` and once for `$plot` because the trimmed objects keep the UMAP coordinates but not the assay the UMAP was originally computed from.

* `detected inputs from HUMAN with id type Gene.name` and the matching line for the reference are Azimuth checking that both objects use the same species and the same flavour of gene identifier.

* `Normalizing query using reference SCT model` is the key step for comparability: rather than fitting a fresh SCTransform model on `pbmc3k`, Azimuth computes the query's Pearson residuals using the gene-level regression parameters that were fit on the reference, putting query and reference values on a common scale.
* `No layers found matching search pattern provided` is Seurat v5 layer bookkeeping. The feature message tells us that of the reference's 5,000 SCT features, 952 are absent from `pbmc3k`, so the mapping proceeds on the 4,048 shared genes; these are the genes from `refdr_features` that can actually influence where a query cell lands.

* `Projecting cell embeddings` multiplies the query residuals by the `refDR` feature loadings. No PCA is computed on the query; it simply inherits the reference's coordinate system.

* `Finding query neighbors` / `Finding neighborhoods` / `Finding anchors` is `FindTransferAnchors()` using the `idx.annoy` index to look up reference neighbours quickly. An anchor is a query–reference cell pair that are mutual nearest neighbours in `refDR` space, and 3,015 of them were found for the 2,700 query cells.
* `Finding integration vectors` and the progress bar for `integration vector weights` then decide how much each anchor contributes to each query cell, weighted by distance.

* `Predicting cell labels` appears three times because the PBMC reference carries three annotation levels: `celltype.l1`, `celltype.l2` and `celltype.l3`. For each level, the labels of the anchor reference cells become a weighted vote, giving every query cell a score for every possible label.

* `Integrating dataset 2 with reference dataset` corrects the projected query coordinates towards the reference using the same anchors, producing the `integrated_dr` reduction.

* `Computing nearest neighbors` and `Running UMAP projection` do not compute a new UMAP. They feed the query neighbours into the reference's saved uwot model so that the query cells land in the reference's existing 2D coordinates, giving the `ref.umap` reduction. This is what makes query and reference plottable on the same axes.

The final block of projecting the reference PCA onto the query, projecting the query back into its original PCA space, then comparing neighbourhoods is `MappingScore()`. It asks whether each query cell keeps the same neighbours after a round trip through the reference space. A cell that does is well described by the reference; a low score hints that the cell is a type the reference does not contain, which is worth knowing because every query cell gets a label whether or not the reference has a sensible one for it.

The returned object has 32,833 features across 4 assays. `RNA` is still the active assay with the original 32,738 genes and only a `counts` layer, i.e. the SCT normalisation used internally is not handed back. The extra 95 features are the rows of `prediction.score.celltype.l1`, `prediction.score.celltype.l2` and `prediction.score.celltype.l3`, one row per candidate label at that level plus a `max` row. The labels themselves are in the metadata as `predicted.celltype.l1`, `predicted.celltype.l2` and `predicted.celltype.l3` with their matching `.score` columns, alongside `mapping.score`.
