---
title: "Seurat analysis using memory versus on disk"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

BPCells is an [R package](https://github.com/bnprks/BPCells) that allows for computationally efficient single-cell analysis. It utilizes bit-packing compression to store counts matrices on disk and C++ code to cache operations. 

```{r install_bpcells, eval=FALSE}
remotes::install_github("bnprks/BPCells/r")
```

```{r load_libraries}
suppressPackageStartupMessages(library(BPCells))
suppressPackageStartupMessages(library(Seurat))
```

# Load Data

Download the Peripheral Blood Mononuclear Cells (PBMCs) 2,700 cells dataset.

```{r download_pbmc3k, eval=FALSE}
mkdir -p data/pbmc3k && cd data/pbmc3k
wget -c https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz
tar -xzf pbmc3k_filtered_gene_bc_matrices.tar.gz
```

Create Seurat object.

```{r seurat_obj_mem}
work_dir <- rprojroot::find_rstudio_root_file()
hdf5_file <- paste0(work_dir, "/data/5k_Human_Donor1_PBMC_3p_gem-x_5k_Human_Donor1_PBMC_3p_gem-x_count_sample_filtered_feature_bc_matrix.h5")
stopifnot(file.exists(hdf5_file))

seurat_obj_mem <- CreateSeuratObject(
  counts = Seurat::Read10X_h5(hdf5_file),
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

class(seurat_obj_mem@assays$RNA$counts)
```

Create Seurat object using {BPCells}.

```{r seurat_obj_bpcells}
seurat_obj_bpcells <- CreateSeuratObject(
  counts = BPCells::open_matrix_10x_hdf5(hdf5_file),
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

seurat_obj_bpcells@assays$RNA$counts
```

## Seurat workflow

Seurat version 4 workflow.

```{r seurat_workflow}
options(future.globals.maxSize = 1.5 * 1024^3)

fixed_PrepDR5 <- function(object, features = NULL, layer = 'scale.data', verbose = TRUE) {
  layer <- layer[1L]
  olayer <- layer
  layer <- SeuratObject::Layers(object = object, search = layer)
  if (is.null(layer)) {
    abort(paste0("No layer matching pattern '", olayer, "' not found. Please run ScaleData and retry"))
  }
  data.use <- SeuratObject::LayerData(object = object, layer = layer)
  features <- features %||% VariableFeatures(object = object)
  if (!length(x = features)) {
    stop("No variable features, run FindVariableFeatures() or provide a vector of features", call. = FALSE)
  }
  if (is(data.use, "IterableMatrix")) {
    features.var <- BPCells::matrix_stats(matrix=data.use, row_stats="variance")$row_stats["variance",]
  } else {
    features.var <- apply(X = data.use, MARGIN = 1L, FUN = var)
  }
  features.keep <- features[features.var > 0]
  if (!length(x = features.keep)) {
    stop("None of the requested features have any variance", call. = FALSE)
  } else if (length(x = features.keep) < length(x = features)) {
    exclude <- setdiff(x = features, y = features.keep)
    if (isTRUE(x = verbose)) {
      warning(
        "The following ",
        length(x = exclude),
        " features requested have zero variance; running reduction without them: ",
        paste(exclude, collapse = ', '),
        call. = FALSE,
        immediate. = TRUE
      )
    }
  }
  features <- features.keep
  features <- features[!is.na(x = features)]
  features.use <- features[features %in% rownames(data.use)]
  if(!isTRUE(all.equal(features, features.use))) {
    missing_features <- setdiff(features, features.use)
    if(length(missing_features) > 0) {
    warning_message <- paste("The following features were not available: ",
                             paste(missing_features, collapse = ", "),
                             ".", sep = "")
    warning(warning_message, immediate. = TRUE)
    }
  }
  data.use <- data.use[features.use, ]
  return(data.use)
}

assignInNamespace('PrepDR5', fixed_PrepDR5, 'Seurat')

seurat_workflow <- function(obj, debug_flag = FALSE){
  obj <- NormalizeData(obj, normalization.method = "LogNormalize", verbose = debug_flag)
  obj <- FindVariableFeatures(obj, selection.method = 'vst', nfeatures = 2000, verbose = debug_flag)
  obj <- ScaleData(obj, verbose = debug_flag)
  obj <- RunPCA(obj, verbose = debug_flag)
  obj <- RunUMAP(obj, dims = 1:30, verbose = debug_flag)
  obj <- FindNeighbors(obj, dims = 1:30, verbose = debug_flag)
  obj <- FindClusters(obj, resolution = 0.5, verbose = debug_flag)
  obj
}
```

Run workflow for in memory.

```{r run_seurat_workflow_mem}
start_time <- Sys.time()
seurat_obj_mem <- seurat_workflow(seurat_obj_mem)
end_time <- Sys.time()
end_time - start_time
```

Run workflow for on disk.

```{r run_seurat_workflow_bpcells}
start_time <- Sys.time()
seurat_obj_bpcells <- seurat_workflow(seurat_obj_bpcells)
end_time <- Sys.time()
end_time - start_time
```

## Results

Compare clustering.

```{r compare_clustering}
stopifnot(all(row.names(seurat_obj_bpcells@meta.data) == row.names(seurat_obj_mem@meta.data)))

table(
  seurat_obj_bpcells@meta.data$seurat_clusters,
  seurat_obj_mem@meta.data$seurat_clusters
)
```
