---
title: "Seurat analysis using memory versus on disk"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

BPCells is an [R package](https://github.com/bnprks/BPCells) that allows for computationally efficient single-cell analysis. It utilizes bit-packing compression to store counts matrices on disk and C++ code to cache operations. 

```{r install_bpcells, eval=FALSE}
remotes::install_github("bnprks/BPCells/r")
```

```{r load_libraries}
suppressPackageStartupMessages(library(BPCells))
suppressPackageStartupMessages(library(Seurat))
```

# PBMC3k

Download the Peripheral Blood Mononuclear Cells (PBMCs) 2,700 cells dataset.

```{r download_pbmc3k, eval=FALSE}
mkdir -p data/pbmc3k && cd data/pbmc3k
wget -c https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz
tar -xzf pbmc3k_filtered_gene_bc_matrices.tar.gz
```

Create Seurat object.

```{r seurat_obj_mem}
work_dir <- rprojroot::find_rstudio_root_file()
hdf5_file <- paste0(work_dir, "/data/5k_Human_Donor1_PBMC_3p_gem-x_5k_Human_Donor1_PBMC_3p_gem-x_count_sample_filtered_feature_bc_matrix.h5")
stopifnot(file.exists(hdf5_file))

seurat_obj_mem <- CreateSeuratObject(
  counts = Seurat::Read10X_h5(hdf5_file),
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

class(seurat_obj_mem@assays$RNA$counts)
```

Create Seurat object using {BPCells}.

```{r seurat_obj_bpcells}
seurat_obj_bpcells <- CreateSeuratObject(
  counts = BPCells::open_matrix_10x_hdf5(hdf5_file),
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

seurat_obj_bpcells@assays$RNA$counts
```

## Seurat workflow

Seurat version 4 workflow.

```{r seurat_workflow}
options(future.globals.maxSize = 1.5 * 1024^3)

fixed_PrepDR5 <- function(object, features = NULL, layer = 'scale.data', verbose = TRUE) {
  layer <- layer[1L]
  olayer <- layer
  layer <- SeuratObject::Layers(object = object, search = layer)
  if (is.null(layer)) {
    abort(paste0("No layer matching pattern '", olayer, "' not found. Please run ScaleData and retry"))
  }
  data.use <- SeuratObject::LayerData(object = object, layer = layer)
  features <- features %||% VariableFeatures(object = object)
  if (!length(x = features)) {
    stop("No variable features, run FindVariableFeatures() or provide a vector of features", call. = FALSE)
  }
  if (is(data.use, "IterableMatrix")) {
    features.var <- BPCells::matrix_stats(matrix=data.use, row_stats="variance")$row_stats["variance",]
  } else {
    features.var <- apply(X = data.use, MARGIN = 1L, FUN = var)
  }
  features.keep <- features[features.var > 0]
  if (!length(x = features.keep)) {
    stop("None of the requested features have any variance", call. = FALSE)
  } else if (length(x = features.keep) < length(x = features)) {
    exclude <- setdiff(x = features, y = features.keep)
    if (isTRUE(x = verbose)) {
      warning(
        "The following ",
        length(x = exclude),
        " features requested have zero variance; running reduction without them: ",
        paste(exclude, collapse = ', '),
        call. = FALSE,
        immediate. = TRUE
      )
    }
  }
  features <- features.keep
  features <- features[!is.na(x = features)]
  features.use <- features[features %in% rownames(data.use)]
  if(!isTRUE(all.equal(features, features.use))) {
    missing_features <- setdiff(features, features.use)
    if(length(missing_features) > 0) {
    warning_message <- paste("The following features were not available: ",
                             paste(missing_features, collapse = ", "),
                             ".", sep = "")
    warning(warning_message, immediate. = TRUE)
    }
  }
  data.use <- data.use[features.use, ]
  return(data.use)
}

assignInNamespace('PrepDR5', fixed_PrepDR5, 'Seurat')

seurat_workflow <- function(obj, debug_flag = FALSE){
  obj <- NormalizeData(obj, normalization.method = "LogNormalize", verbose = debug_flag)
  obj <- FindVariableFeatures(obj, selection.method = 'vst', nfeatures = 2000, verbose = debug_flag)
  obj <- ScaleData(obj, verbose = debug_flag)
  obj <- RunPCA(obj, verbose = debug_flag)
  obj <- RunUMAP(obj, dims = 1:30, verbose = debug_flag)
  obj <- FindNeighbors(obj, dims = 1:30, verbose = debug_flag)
  obj <- FindClusters(obj, resolution = 0.5, verbose = debug_flag)
  obj
}
```

Run workflow for in memory.

```{r run_seurat_workflow_mem}
start_time <- Sys.time()
seurat_obj_mem <- seurat_workflow(seurat_obj_mem)
end_time <- Sys.time()
end_time - start_time
```

Run workflow for on disk.

```{r run_seurat_workflow_bpcells}
start_time <- Sys.time()
seurat_obj_bpcells <- seurat_workflow(seurat_obj_bpcells)
end_time <- Sys.time()
end_time - start_time
```

## Results

Compare clustering.

```{r compare_clustering}
stopifnot(all(row.names(seurat_obj_bpcells@meta.data) == row.names(seurat_obj_mem@meta.data)))

table(
  seurat_obj_bpcells@meta.data$seurat_clusters,
  seurat_obj_mem@meta.data$seurat_clusters
)
```

# Tabula Sapiens

[Download](https://cellxgene.cziscience.com/collections/e5f58829-1a66-40b5-a624-9046778e74f5) from Tabula Sapiens:

> Tabula Sapiens is a benchmark, first-draft human cell atlas of over 1.1M cells from 28 organs of 24 normal human subjects. This work is the product of the Tabula Sapiens Consortium. Taking the organs from the same individual controls for genetic background, age, environment, and epigenetic effects, and allows detailed analysis and comparison of cell types that are shared between tissues.

```{r ts_heart}
ts_heart_url <- "https://datasets.cellxgene.cziscience.com/97516b79-8d08-46a6-b329-5d0a25b0be98.h5ad"
ts_heart_file <- paste0('data/', basename(ts_heart_url))
if(file.exists(ts_heart_file)){
  message(paste0(ts_heart_file, " already exists"))
} else {
  options(timeout = 1000)
  download.file(url = ts_heart_url, destfile = ts_heart_file)
}
```

Run workflow for on disk.

```{r run_seurat_workflow_ts_heart}
start_time <- Sys.time()

ts_heart <- CreateSeuratObject(
  counts = BPCells::open_matrix_anndata_hdf5(ts_heart_file),
  min.cells = 3,
  min.features = 200,
  project = "ts_heart"
)

ts_heart <- seurat_workflow(ts_heart)
end_time <- Sys.time()
end_time - start_time
```

UMAP.

```{r ts_heart_umap}
DimPlot(ts_heart)
```
