---
title: "BPCells"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

BPCells is an [R package](https://github.com/bnprks/BPCells) that allows for computationally efficient single-cell analysis. It utilizes bit-packing compression to store counts matrices on disk and C++ code to cache operations. 

```{r install_bpcells, eval=FALSE}
remotes::install_github("bnprks/BPCells/r")
```

## Getting started

Load packages.

```{r load_libraries}
suppressPackageStartupMessages(library(BPCells))
suppressPackageStartupMessages(library(Matrix))
```

Write matrix to disk using BPCells.

```{r write_matrix_dir}
set.seed(1984)
bpcells_dir <- 'bpcells_matrix'
if(dir.exists(bpcells_dir)){
  unlink(bpcells_dir, recursive = TRUE)
}
write_matrix_dir(
  mat = rsparsematrix(50000, 50000, density = 0.01),
  dir = bpcells_dir
)
```

Open the BPCells matrix from disk.

```{r open_matrix_dir}
bp_mat <- open_matrix_dir(bpcells_dir)
bp_mat
```

Calculate row and column sums (lazily, disk-backed).

```{r sums}
row_sums <- rowSums(bp_mat)
col_sums <- colSums(bp_mat)

head(row_sums)

dense_row <- as.matrix(bp_mat[1, ])
sum(dense_row)
```

## Modifying matrix path

Following the [example](https://github.com/bnprks/BPCells/issues/247#issuecomment-2813863978) by Ben Parks:

```{r m1}
my_dir <- file.path(tempdir(), "data")
m1 <- matrix(1:12, nrow=3) |> 
  as("IterableMatrix") |>
  write_matrix_dir(file.path(my_dir, "m1"), overwrite = TRUE) |>
  log1p()
m1
```

`all_matrix_inputs()` strips away any queued operations, i.e., `Transform lop1p` is gone. We can use `inputs` to modify the path. Note that the queued operations in `m1` are intact.

```{r all_matrix_inputs}
inputs <- all_matrix_inputs(m1)
str(inputs)
m1
```

Create another matrix.

```{r m2}
m2 <- matrix(1:21, nrow=3) |> 
  as("IterableMatrix") |>
  write_matrix_dir(file.path(my_dir, "m2"), overwrite = TRUE)
m2
```

Modify path.

```{r modify_path}
inputs[[1]]@dir <- file.path(my_dir, "m2")

all_matrix_inputs(m1) <- inputs
m1
```

Check that it is using `m2`, which has 7 columns.

```{r colmeans_m1}
Matrix::colMeans(m1)
```

### Relative paths

It seems that `write_matrix_dir()` uses full paths by default.

```{r m3}
m3 <- matrix(1:12, nrow=3) |> 
  as("IterableMatrix") |>
  write_matrix_dir("m3", overwrite = TRUE) |>
  log1p()
m3
```

Will it work if I modify it to a relative path?

```{r m3_inputs}
m3_inputs <- all_matrix_inputs(m3)
m3_inputs[[1]]@dir <- file.path("m3")
all_matrix_inputs(m3) <- m3_inputs
m3
```

Calculate column means.

```{r colmeans_m3}
Matrix::colMeans(m3)
```
