---
title: "Pokedex for Cell Types"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> The celldex package provides convenient access to several cell type reference datasets. Most of these references are derived from bulk RNA-seq or microarray data of cell populations that (hopefully) consist of a pure cell type after sorting and/or culturing. The aim is to provide a common resource for further analysis like cell type annotation of single cell expression data or deconvolution of cell type proportions in bulk expression datasets.

## Installation

Install [celldex](https://bioconductor.org/packages/3.20/data/experiment/html/celldex.html).

```{r install_celldex, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("celldex")
```

## References

We can examine the available references using the `surveyReferences()` function. This returns a DataFrame of the reference's name and version, along with additional information like the title, description, species, number of samples, available labels, and so on.

```{r survey_references}
suppressPackageStartupMessages(library(celldex))
surveyReferences()
```
