---
title: "Getting started with CELLxGENE Discover Census"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# https://stackoverflow.com/questions/30237310/setting-work-directory-in-knitr-using-opts-chunksetroot-dir-doesnt-wor
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Preparation

Following the [documentation](https://chanzuckerberg.github.io/cellxgene-census/r/index.html) for the R package {cellxgene.census} which is part of CZ CELLxGENE Discover Census. 

{cellxgene.census} provides an API to efficiently access the cloud-hosted Census single-cell data from R. In just a few seconds users can access any slice of Census data using cell or gene filters across hundreds of single-cell datasets.

Census data can be fetched in an iterative fashion for bigger-than-memory slices of data, or quickly exported to basic R structures, as well as {Seurat} or {SingleCellExperiment} objects for downstream analysis.

Install from the [R-universe platform](https://ropensci.org/r-universe/). If installing using Ubuntu/Debian, you may need to install the following libraries via APT:

* libxml2-dev
* libssl-dev
* libcurl4-openssl-dev

In addition you must have at least `cmake` v3.21. Install the {tiledbsoma} dependency first, since this takes some time to compile and install, it's better to fail first.

```{r install_tiledbsoma, eval=FALSE}
install.packages('tiledbsoma', repos = c('https://tiledb-inc.r-universe.dev',
                                         'https://cloud.r-project.org'))
```

Now install {cellxgene.census}, which should be a breeze now.

```{r install_cellxgene_census, eval=FALSE}
install.packages(
  "cellxgene.census",
  repos=c('https://chanzuckerberg.r-universe.dev', 'https://cloud.r-project.org')
)
```

Now install {Seurat}.

```{r install_seurat, eval=FALSE}
install.packages("Seurat")
```

## Querying the metadata

[Querying and fetching the single-cell data and cell/gene metadata](https://chanzuckerberg.github.io/cellxgene-census/r/articles/census_query_extract.html)

```{r load_cellxgene_census}
library("cellxgene.census")
packageVersion("cellxgene.census")
```

The human gene metadata of the Census, for RNA assays, is located at `census$get("census_data")$get("homo_sapiens")$obs`. The mouse cell metadata is at `census$get("census_data")$get("mus_musculus").obs`.

To learn what metadata columns are available for fetching and filtering we can directly look at the keys of the cell metadata.

```{r my_keys}
census <- open_soma()
my_keys <- census$get("census_data")$get("homo_sapiens")$obs$colnames()
my_keys
```

`soma_joinid` is a special `SOMADataFrame` column that is used for join operations. All of the keys can be used to fetch specific columns or specific rows matching a condition. For the latter we need to know the values we are looking for _a priori_.

For example let's see what are the possible values available for `sex`. To this we can load all cell metadata but fetching only for the column sex; `column_names` are character vector indicating what metadata columns to fetch.

```{r sex_values}
census$get("census_data")$get("homo_sapiens")$obs$read(column_names = "sex")$concat() |>
  as.data.frame() |>
  unique()
```

With this information we can fetch all cell metadata for a specific sex value, for example "unknown"; the `value_filter` is an R expression with selection conditions to fetch rows.

```{r sex_unknown}
census$get("census_data")$get("homo_sapiens")$obs$read(value_filter = "sex == 'unknown'")$concat() |>
  as.data.frame() -> sex_unknown

head(sex_unknown)
```

We can use both `column_names` and `value_filter` to perform specific queries. For example fetching the disease column for the cell_type "B cell" in the `tissue_general` "lung".

```{r cell_metadata_b_cell}
cell_metadata_b_cell <- census$get("census_data")$get("homo_sapiens")$obs$read(
  value_filter = "cell_type == 'B cell' & tissue_general == 'lung'",
  column_names = "disease"
)

cell_metadata_b_cell <- as.data.frame(cell_metadata_b_cell$concat())

table(cell_metadata_b_cell)
```

### Querying expression data as Seurat

Use `get_seurat()` to perform the same type of filtering but returning a Seurat object.

* `obs_column_names` — character vector indicating the columns to select for cell metadata.
* `obs_value_filter` — expression with selection conditions to fetch cells meeting a criteria.
* `var_column_names` — character vector indicating the columns to select for gene metadata.
* `var_value_filter` — expression with selection conditions to fetch genes meeting a criteria.

```{r get_seurat, eval=FALSE}
library("Seurat")

seurat_obj <- get_seurat(
  census, "Homo sapiens",
  obs_column_names = c("cell_type", "tissue_general", "disease", "sex"),
  var_value_filter = "feature_id %in% c('ENSG00000161798', 'ENSG00000188229')",
  obs_value_filter = "cell_type == 'B cell' & tissue_general == 'lung' & disease == 'COVID-19'"
)

saveRDS(object = seurat_obj, file = "data/lung_bcell.rds")
```

Seurat object.

```{r seurat_obj}
seurat_obj <- readRDS(file = "data/lung_bcell.rds")
seurat_obj
```

### Close the census

After use, the census object should be closed to release memory and other resources. This also closes all SOMA objects accessed via the top-level census. Closing can be automated using `on.exit(census$close(), add = TRUE)` immediately after `census <- open_soma()`.

```{r census_close}
census$close()
```
