---
title: "Chi-squared Tests"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
})
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The Chi-squared test checks whether two categorical variables are related. For example:

* Is smoking related to cancer?
* Is gender related to favorite ice cream flavor?
* Is a treatment related to an outcome?

A Chi-squared distribution is a probability distribution that arises when a test statistic is built from squared deviations. Squaring measures how large the differences are, treats positive and negative deviations equally, and leads to a distribution with well-understood mathematical properties (the Chi-squared distribution).

## Is smoking associated with lung cancer?

Create a contingency table

```{r eg1}
data <- matrix(
  c(60, 40, 30, 70),
  nrow = 2,
  byrow = TRUE
)

colnames(data) <- c("Cancer", "NoCancer")
rownames(data) <- c("Smoker", "NonSmoker")

data

chisq.test(data)
```

The test statistic is `X-squared = 16.99` and it measures how different your observed table is from what you would expect if the two variables were not related.

* Bigger number means bigger difference from no relationship
* Smaller number means closer to no relationship

The degrees of freedom for a 2×2 table is always 1.

A p-value answers the question:

* If there were no real association between the variables, how likely is it to see a difference as extreme as the one we observed?

A p-value of 3.758e-05 means that there's only a 0.0037% chance that this big a difference happened by random chance.

```{r dchisq}
x2 <- chisq.test(data)$statistic
df <- chisq.test(data)$parameter

x <- seq(0, x2+3, length = 500)

plot(
  x,
  dchisq(x, df),
  type = "l",
  lwd = 2,
  ylab = "Density",
  xlab = "Chi-squared value",
  main = paste0("Chi-squared distribution (df = ", df, ")")
)

abline(v = x2, col = "red", lwd = 2)
```

The Chi-square test statistic always follows a Chi-square distribution if the null hypothesis is true. Therefore by comparing our obtained value (16.99) to the known curve (df = 1), we can calculate the probability of getting such an extreme value by chance. That probability is the p-value.
