---
title: "Getting started with clusterProfiler"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggarchery))
suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(org.Hs.eg.db))
```

## Getting started

Install.

```{r install_cluster_profiler, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("clusterProfiler")
BiocManager::install("org.Hs.eg.db")
install.packages('ggarchery')
```

## Example data

Use [An example differential gene expression results table](https://davetang.org/muse/2024/10/31/an-example-differential-gene-expression-results-table/).

```{r de_res}
de_res <- read_csv("https://raw.githubusercontent.com/davetang/muse/refs/heads/main/data/13970886_edger_res.csv", show_col_types = FALSE)
head(de_res)
```

## Gene IDs

The {clusterProfiler} package uses the `enrichGO()` function for performing a [Gene Ontology over-representation test](https://yulab-smu.top/biomedical-knowledge-mining-book/clusterprofiler-go.html#clusterprofiler-go-ora). The input for `gene` is a vector of Entrez Gene IDs.

Our example data uses Ensembl gene IDs, so we need to convert these into Entrez Gene IDs. We can use the {org.Hs.eg.db} package provided by [Bioconductor](https://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html), which provides genome wide annotation for human, primarily based on mapping using Entrez Gene identifiers. 

```{r ensembl_to_entrez}
ensembl_to_entrez <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = de_res$ensembl_gene_id,
  columns = c("ENSEMBL", "ENTREZID"), 
  keytype = "ENSEMBL"
)

de_res |>
  dplyr::inner_join(ensembl_to_entrez, by = dplyr::join_by(ensembl_gene_id == ENSEMBL)) |>
  dplyr::select(ensembl_gene_id, ENTREZID, dplyr::everything()) -> de_res

head(de_res)
```

## clusterProfiler

The example data contains results of a comparison between cancer samples and normal samples; positive fold change indicates that genes were expressed higher in cancer. There are a lot of genes up-regulated in cancer.

```{r up_regulated_in_cancer}
de_res |>
  dplyr::filter(logFC > 0, adjusted_pvalue < 0.05) |>
  nrow()
```

Instead of examining the list of up-regulated genes individually, which would take a long time, we can see whether these genes have common Gene Ontology terms more often than expected.

First we'll get the top 500 most significantly up-regulated genes.

```{r sig_up_genes}
de_res |>
  dplyr::filter(logFC > 0, adjusted_pvalue < 0.05) |>
  dplyr::slice_min(order_by = adjusted_pvalue, n = 500) |>
  dplyr::filter(!is.na(ENTREZID)) |>
  dplyr::pull(ENTREZID) -> sig_up_genes
```

Create the background, i.e., the universe, so we can test for over-representation.

```{r the_universe}
de_res |>
  dplyr::filter(!is.na(ENTREZID)) |>
  dplyr::pull(ENTREZID) -> the_universe
```

Gene Ontology terms are grouped into three ontologies:

1. Molecular Function (MF)
2. Biological Process (BP), and
3. Cellular Component (CC)

Here, we'll perform an over-representation analysis using biological processes.

```{r ego_bp}
ego_bp <- enrichGO(
  gene          = sig_up_genes,
  universe      = the_universe,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.01,
  qvalueCutoff  = 0.05,
  minGSSize     = 10,
  maxGSSize     = 500,
  readable      = TRUE
)

head(ego_bp)
```

Bar plot showing each enriched GO term coloured by the adjusted p-value.

```{r barplot}
barplot(ego_bp, showCategory=10)
```

Dot plot showing each enriched GO term with associated statistics.

```{r dotplot}
dotplot(ego_bp, showCategory=10)
```

Heat plot showing the enriched GO terms on the y-axis and the genes on the x-axis. Genes with the associated GO term are highlighted.

```{r heatplot}
heatplot(ego_bp, showCategory=10)
```

`goplot` shows the gene ontology graph with the enriched GO terms highlighted.

```{r goplot, fig.height=10, fig.width=12}
goplot(ego_bp)
```

Another nice feature of {clusterProfiler} is that you can plot multiple gene lists together. We can create a list of down-regulated genes.

```{r sig_down_genes}
de_res |>
  dplyr::filter(logFC < 0, adjusted_pvalue < 0.05) |>
  dplyr::slice_min(order_by = adjusted_pvalue, n = 500) |>
  dplyr::filter(!is.na(ENTREZID)) |>
  dplyr::pull(ENTREZID) -> sig_down_genes
```

Perform GO enrichment on two gene lists.

```{r ego_bp_both}
my_gene_list <- list(
  up_gene = sig_up_genes,
  down_gene = sig_down_genes
)

ego_bp_both <- compareCluster(
  geneCluster = my_gene_list,
  fun = "enrichGO",
  universe = the_universe,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pvalueCutoff = 0.01,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  minGSSize = 10,
  maxGSSize = 500,
  readable = TRUE
)

head(as.data.frame(ego_bp_both))
```

Dot plot with enriched GO terms by gene list.

```{r dotplot_gene_list, fig.width=12, fig.height=12}
dotplot(ego_bp_both, showCategory = 15)
```
