---
title: "Making a heatmap in R with the ComplexHeatmap package"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ComplexHeatmap)
library(circlize)
```

## Background

[ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)
is an R/Bioconductor package developed by Zuguang Gu that provides a highly
flexible framework for drawing heatmaps. While simpler packages like `pheatmap`
or `heatmap2` are fine for quick visualisations, ComplexHeatmap excels when
you need:

* **Multiple heatmaps** placed side-by-side with a shared colour legend.
* **Rich annotations** along rows and/or columns — categorical, continuous,
  bar plots, box plots, density plots, text, and more.
* **Heatmap splitting** (slicing) by k-means clusters or a grouping factor.
* **UpSet plots** as an alternative to Venn diagrams.
* Fine-grained control over colours, dendrograms, legends, and layout.

The comprehensive reference book is freely available at
<https://jokergoo.github.io/ComplexHeatmap-reference/book/>.

## Installation

Install the stable version from
[Bioconductor](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html).

```{r install, eval=FALSE}
BiocManager::install("ComplexHeatmap")
# circlize is used for colorRamp2(), a flexible colour mapping helper
install.packages("circlize")
```

## Data

Download example RNA-seq count data (TagSeq format). We keep only genes with
a high total count so the heatmap is not too crowded.

```{r example_data}
example_file <- "https://davetang.org/file/TagSeqExample.tab"
data <- read.delim(example_file, header = TRUE, row.names = "gene")
data_subset <- as.matrix(data[rowSums(data) > 50000, ])
dim(data_subset)
head(data_subset)
```

The six columns are biological samples; each row is a gene. Values are raw
counts.

## Z-score scaling

### Why scale?

Raw counts vary enormously between genes — a highly-expressed gene can have
counts hundreds of times larger than a lowly-expressed one. If we colour the
heatmap on the raw scale, the colour gradient is dominated by the most abundant
genes and most rows look uniformly blue (low) regardless of their actual
expression pattern.

**Z-score row scaling** transforms each row so that it has mean 0 and standard
deviation 1:

$$z_i = \frac{x_i - \bar{x}}{\text{sd}(x)}$$

After this transformation the colour of each cell reflects *how far that
sample is from the gene's own mean*, not the absolute expression level.
This makes cross-gene patterns — e.g. a cluster of genes all up-regulated in
one condition — visually apparent.

### Row scaling vs column scaling

| Direction | What is normalised | Typical use |
|-----------|-------------------|-------------|
| **Row** (`apply(mat, 1, ...)`) | Each gene across samples | Compare sample patterns per gene |
| **Column** (`apply(mat, 2, ...)`) | Each sample across genes | Compare gene patterns per sample (rare) |

Row scaling is by far the more common choice in genomics.

```{r zscore_function}
cal_z_score <- function(x) (x - mean(x)) / sd(x)

# Apply row-wise: MARGIN = 1
data_subset_norm <- t(apply(data_subset, 1, cal_z_score))

# Sanity check — each row should now have mean ≈ 0 and sd ≈ 1
round(head(rowMeans(data_subset_norm)), 10)
round(head(apply(data_subset_norm, 1, sd)), 10)
```

### Raw vs scaled side-by-side

Placing the raw and z-scored heatmaps next to each other makes the benefit of
scaling clear.

```{r raw_vs_scaled, fig.width=12, fig.height=8}
ht_raw  <- Heatmap(data_subset,      column_title = "Raw counts",  name = "counts")
ht_norm <- Heatmap(data_subset_norm, column_title = "Z-score",     name = "z-score")

ht_raw + ht_norm
```

### Custom colour scale for Z-scores

The default blue-to-red diverging palette works well for z-scores, but you can
be explicit with `colorRamp2()` from the **circlize** package. Symmetric
limits (e.g. −2 to +2) ensure that zero (the gene mean) always maps to white.

```{r zscore_custom_colour, fig.width=7, fig.height=8}
col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

Heatmap(
  data_subset_norm,
  name            = "z-score",
  col             = col_fun,
  column_title    = "Z-score normalised expression",
  row_title       = paste(nrow(data_subset_norm), "genes"),
  column_names_rot = 45
)
```

Values beyond ±2 are clipped to the extreme colours, which is usually
desirable — it prevents a single outlier gene from washing out the rest of the
palette.

## Basic heatmap options

### Titles and legend name

`column_title` / `row_title` label the heatmap body; `name` is the title shown
above the colour legend.

```{r basic_options, fig.width=7, fig.height=8}
Heatmap(
  data_subset_norm,
  name         = "z-score",
  column_title = "Samples",
  row_title    = "Genes",
  col          = col_fun
)
```

### Turning off clustering

```{r no_cluster, fig.width=7, fig.height=8}
Heatmap(
  data_subset_norm,
  name                  = "z-score",
  col                   = col_fun,
  cluster_rows          = FALSE,
  cluster_columns       = FALSE,
  column_title          = "No clustering"
)
```

## Annotations

[Annotations](https://jokergoo.github.io/ComplexHeatmap-reference/book/heatmap-annotations.html)
are one of the most powerful features of ComplexHeatmap. They sit alongside the
heatmap body and convey additional metadata about samples (column annotations)
or genes (row annotations).

We will build a small synthetic dataset with clear group structure to make the
annotations easy to read.

```{r annotation_data}
set.seed(42)
n_genes   <- 20
n_samples <- 12

# Simulate two conditions with 6 replicates each
condition <- rep(c("Control", "Treatment"), each = 6)
batch     <- rep(c("Batch1", "Batch2", "Batch1", "Batch2", "Batch1", "Batch2"), 2)

# Generate a matrix where Treatment genes have higher expression
mat <- matrix(rnorm(n_genes * n_samples), nrow = n_genes, ncol = n_samples)
mat[1:10, 7:12] <- mat[1:10, 7:12] + 2   # first 10 genes up in Treatment

rownames(mat) <- paste0("Gene", seq_len(n_genes))
colnames(mat) <- paste0(condition, "_", seq_len(n_samples))
```

### Column (sample) annotations

`HeatmapAnnotation()` accepts:

* **Named character/factor vectors** → drawn as coloured tiles (categorical).
* **Named numeric vectors** → drawn as a gradient (continuous).
* **`anno_*()` functions** → barplots, boxplots, density, text, etc.

```{r col_annotation, fig.width=9, fig.height=8}
# Custom colour maps for categorical annotations
cond_colours  <- c(Control = "#4393C3", Treatment = "#D6604D")
batch_colours <- c(Batch1  = "#F1A340", Batch2    = "#998EC3")

col_ha <- HeatmapAnnotation(
  Condition  = condition,
  Batch      = batch,
  MeanExpr   = colMeans(mat),            # continuous: shown as gradient
  col = list(
    Condition = cond_colours,
    Batch     = batch_colours,
    MeanExpr  = colorRamp2(c(-1, 0, 1), c("white", "grey60", "black"))
  ),
  annotation_name_side = "left"
)

Heatmap(
  mat,
  name             = "expr",
  top_annotation   = col_ha,
  column_title     = "Samples with column annotations",
  show_column_names = FALSE
)
```

### Row (gene) annotations

`rowAnnotation()` works the same way but applies to rows. A common use case is
to annotate genes with their log-fold change or membership in a pathway.

```{r row_annotation, fig.width=10, fig.height=8}
gene_group <- rep(c("Up", "Down"), each = 10)
lfc        <- rnorm(n_genes, mean = ifelse(gene_group == "Up", 1.5, -1.5), sd = 0.5)

row_ha <- rowAnnotation(
  Group = gene_group,
  LFC   = lfc,
  col   = list(
    Group = c(Up = "#D6604D", Down = "#4393C3"),
    LFC   = colorRamp2(c(-3, 0, 3), c("blue", "white", "red"))
  ),
  annotation_name_side = "top"
)

Heatmap(
  mat,
  name             = "expr",
  top_annotation   = col_ha,
  right_annotation = row_ha,
  column_title     = "Heatmap with column and row annotations",
  show_column_names = FALSE
)
```

### Bar plot and box plot annotations

`anno_barplot()` and `anno_boxplot()` let you display summary statistics
directly alongside the heatmap, saving the need for a separate figure.

```{r anno_barplot_boxplot, fig.width=10, fig.height=8}
col_ha2 <- HeatmapAnnotation(
  Condition = condition,
  MeanExpr  = anno_barplot(colMeans(mat), height = unit(2, "cm")),
  col = list(Condition = cond_colours)
)

row_ha2 <- rowAnnotation(
  BoxExpr = anno_boxplot(mat, width = unit(3, "cm"))
)

Heatmap(
  mat,
  name             = "expr",
  top_annotation   = col_ha2,
  right_annotation = row_ha2,
  show_column_names = FALSE,
  column_title      = "Bar and box plot annotations"
)
```

### Text annotations

`anno_text()` is useful for labelling a small number of important genes or
samples directly on the plot rather than relying on axis labels.

```{r anno_text, fig.width=10, fig.height=8}
# Highlight the top 5 most variable genes
gene_vars <- apply(mat, 1, var)
top5       <- names(sort(gene_vars, decreasing = TRUE))[1:5]
gene_label <- ifelse(rownames(mat) %in% top5, rownames(mat), "")

row_ha3 <- rowAnnotation(
  Label = anno_text(gene_label, gp = gpar(fontsize = 9, fontface = "bold"))
)

Heatmap(
  mat,
  name             = "expr",
  right_annotation = row_ha3,
  show_row_names   = FALSE,
  show_column_names = FALSE,
  column_title      = "Text annotation for top variable genes"
)
```

## Heatmap splitting

Splitting breaks the heatmap into visual blocks, making clusters easier to
interpret. You can split by k-means clustering or by an explicit grouping
factor.

### K-means splitting

```{r km_split, fig.width=9, fig.height=8}
Heatmap(
  mat,
  name             = "expr",
  top_annotation   = col_ha,
  row_km           = 2,         # 2 row clusters
  column_km        = 2,         # 2 column clusters
  show_column_names = FALSE,
  column_title      = "K-means split (k=2)"
)
```

### Splitting by a factor

When you already know the grouping (e.g. condition), pass it directly. This
guarantees that samples from the same group are adjacent.

```{r factor_split, fig.width=9, fig.height=8}
Heatmap(
  mat,
  name              = "expr",
  top_annotation    = col_ha,
  column_split      = condition,          # columns split by condition
  row_split         = gene_group,         # rows split by gene group
  show_column_names = FALSE,
  column_title_gp   = gpar(fontsize = 10, fontface = "bold"),
  row_title_gp      = gpar(fontsize = 10, fontface = "bold")
)
```

## Putting it all together

A publication-ready heatmap combining z-score scaling, colour scale, full
annotations, and factor-based splitting.

```{r publication_heatmap, fig.width=11, fig.height=9}
mat_z <- t(apply(mat, 1, cal_z_score))

col_ha_final <- HeatmapAnnotation(
  Condition = condition,
  Batch     = batch,
  col = list(
    Condition = cond_colours,
    Batch     = batch_colours
  ),
  annotation_name_side = "left",
  show_legend          = TRUE
)

row_ha_final <- rowAnnotation(
  Group = gene_group,
  LFC   = lfc,
  col = list(
    Group = c(Up = "#D6604D", Down = "#4393C3"),
    LFC   = colorRamp2(c(-3, 0, 3), c("blue", "white", "red"))
  ),
  annotation_name_side = "top"
)

ht_final <- Heatmap(
  mat_z,
  name              = "z-score",
  col               = colorRamp2(c(-2, 0, 2), c("blue", "white", "red")),
  top_annotation    = col_ha_final,
  right_annotation  = row_ha_final,
  column_split      = condition,
  row_split         = gene_group,
  show_column_names = FALSE,
  column_title_gp   = gpar(fontsize = 11, fontface = "bold"),
  row_title_gp      = gpar(fontsize = 11, fontface = "bold"),
  heatmap_legend_param = list(
    title            = "z-score",
    at               = c(-2, -1, 0, 1, 2),
    legend_direction = "horizontal"
  )
)

draw(ht_final, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
```

## UpSet plot

ComplexHeatmap has an [implementation of UpSet plots](https://jokergoo.github.io/ComplexHeatmap-reference/book/upset-plot.html)
that scales better than Venn diagrams for more than three sets.

Three lists.

```{r upset_plot_eg1}
set.seed(123)
lt <- list(a = sample(letters, 10),
          b = sample(letters, 15),
          c = sample(letters, 20))
m <- make_comb_mat(lt)
UpSet(m)
```

> union mode: 1 means in that set and 0 is not taken into account. When there
> are multiple 1, the relationship is OR. Then, 1 1 0 means a set of elements
> in set A or B, and they can also in C or not in C (union(A, B)). Under this
> mode, the seven combination sets can overlap.

```{r upset_plot_eg2}
m = make_comb_mat(lt, mode = "union")
UpSet(m)
```

With colours.

```{r upset_plot_eg2_col}
UpSet(m, comb_col = c(rep(2, 3), rep(3, 3), 1))
```

Compare two UpSet plots.

```{r upset_plot_eg3}
set.seed(123)
lt1 = list(a = sample(letters, 10),
          b = sample(letters, 15),
          c = sample(letters, 20))
m1 = make_comb_mat(lt1)

set.seed(456)
lt2 = list(a = sample(letters, 10),
          b = sample(letters, 15),
          c = sample(letters, 20))
m2 = make_comb_mat(lt2)

max1 = max(c(set_size(m1), set_size(m2)))
max2 = max(c(comb_size(m1), comb_size(m2)))

UpSet(m1, top_annotation = upset_top_annotation(m1, ylim = c(0, max2)),
    right_annotation = upset_right_annotation(m1, ylim = c(0, max1)),
    column_title = "UpSet1") +
UpSet(m2, top_annotation = upset_top_annotation(m2, ylim = c(0, max2)),
    right_annotation = upset_right_annotation(m2, ylim = c(0, max1)),
    column_title = "UpSet2")
```

## Exporting to PNG

### The basic pattern

ComplexHeatmap draws via the **grid** graphics system, so you must open a
graphics device, call `draw()`, and then close the device with `dev.off()`.
Calling `png()` alone and then printing the heatmap object works too, but
using `draw()` gives you control over legend placement at export time.

```{r export_basic, eval=FALSE}
png(
  filename = "heatmap.png",
  width    = 8,      # inches
  height   = 6,      # inches
  units    = "in",
  res      = 300,    # dots per inch — 300 is standard for publication
  bg       = "white" # explicit white background (avoids transparency issues)
)
draw(ht_final, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
dev.off()
```

### Key parameters

| Parameter | Typical values | Notes |
|-----------|---------------|-------|
| `width` / `height` | 6–14 | In the units given by `units` |
| `units` | `"in"`, `"cm"`, `"px"` | `"in"` is most intuitive for print |
| `res` | 72, 150, 300, 600 | See table below |
| `bg` | `"white"`, `"transparent"` | Journals usually require white |
| `pointsize` | 8–14 | Base font size in points; default is 12 |

Common `res` (DPI) values and their intended use:

| DPI | Use case |
|-----|----------|
| 72  | Screen / web preview |
| 150 | Draft print, presentations |
| 300 | Standard publication quality |
| 600 | High-resolution print / microscopy figures |

### Sizing the image to the heatmap content

A heatmap with many rows needs more height than one with few rows. A simple
rule of thumb: allocate ~0.25 inches per row and ~0.6 inches per column, then
add a fixed margin for titles, legends, and annotations.

```{r export_sized, eval=FALSE}
n_rows <- nrow(mat_z)
n_cols <- ncol(mat_z)

img_width  <- n_cols * 0.6 + 3   # 3 in of margin for row annotations + legend
img_height <- n_rows * 0.25 + 3  # 3 in of margin for col annotations + legend

png(
  filename = "heatmap_sized.png",
  width    = img_width,
  height   = img_height,
  units    = "in",
  res      = 300,
  bg       = "white"
)
draw(ht_final, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
dev.off()
```

For very large matrices (hundreds of rows) you may want a fixed height with
`show_row_names = FALSE` rather than scaling linearly.

### Adjusting font sizes for high-resolution output

At 300 DPI the default `pointsize = 12` renders fine. If you increase `res`
to 600 DPI the plot appears to shrink unless you increase `pointsize` to
match, or enlarge `width` and `height`. For example, to maintain the same
visual size at 600 DPI as a 300-DPI export:

```{r export_600dpi, eval=FALSE}
png(
  filename  = "heatmap_600dpi.png",
  width     = img_width * 2,   # double the physical size ...
  height    = img_height * 2,
  units     = "in",
  res       = 600,             # ... at double the DPI = same pixel count
  pointsize = 16,              # larger base font to stay legible
  bg        = "white"
)
draw(ht_final, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
dev.off()
```

### Vector formats (PDF and SVG)

For formats that require infinite scalability (e.g. figures submitted as PDF
to a journal), replace `png()` with `pdf()` or `svg()`. These devices ignore
`res` because they are resolution-independent.

```{r export_pdf, eval=FALSE}
pdf(
  file   = "heatmap.pdf",
  width  = img_width,
  height = img_height
)
draw(ht_final, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
dev.off()
```

```{r export_svg, eval=FALSE}
svg(
  filename = "heatmap.svg",
  width    = img_width,
  height   = img_height
)
draw(ht_final, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
dev.off()
```

### Exporting multiple heatmaps to one file

A concatenated heatmap (created with `+`) is itself a `HeatmapList` object
and is drawn exactly the same way.

```{r export_multi, eval=FALSE}
ht_list <- ht_raw + ht_norm   # the two-heatmap object built earlier

png(
  filename = "two_heatmaps.png",
  width    = 14,
  height   = 8,
  units    = "in",
  res      = 300,
  bg       = "white"
)
draw(ht_list)
dev.off()
```

## Further reading

* [ComplexHeatmap Complete Reference](https://jokergoo.github.io/ComplexHeatmap-reference/book/) — the authoritative guide by the package author.
* [Bioconductor package page](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) — vignettes and citation.
* `colorRamp2()` from the [circlize](https://cran.r-project.org/package=circlize) package is the recommended way to define colour scales.

## Session info

```{r session_info}
sessionInfo()
```
