---
title: "Downloading files using the curl package"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(curl)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The [{curl} package](https://cran.r-project.org/web/packages/curl/index.html) provides:

> The `curl()` and `curl_download()` functions provide highly configurable drop-in replacements for base `url()` and `download.file()` with better performance, support for encryption (https, ftps), gzip compression, authentication, and other 'libcurl' goodies. The core of the package implements a framework for performing fully customized requests where data can be processed either in memory, on disk, or streaming via the callback or connection interfaces. Some knowledge of 'libcurl' is recommended; for a more-user-friendly web client see the 'httr' package which builds on this package with http specific tools and logic.

Use `multi_download()` to download a relatively big file (9 GB).

```{r multi_download}
outdir <- 'data'
dir.create(outdir, showWarnings = FALSE)

library_file <- "seattle-library-checkouts.csv"
outfile <- paste0(outdir, '/', library_file)

if(!file.exists(outfile)){
  curl::multi_download(
    urls = "https://r4ds.s3.us-west-2.amazonaws.com/seattle-library-checkouts.csv",
    destfiles = outfile,
    resume = TRUE
  )
}
```

File size.

```{r file_size}
file.size(outfile) |> utils:::format.object_size(units = 'Gb')
```
