---
title: "Using dplyr to aggregate in R"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

Investopedia has a good definition of an [aggregate function](https://www.investopedia.com/terms/a/aggregate-function.asp):

>An aggregate function is a mathematical computation involving a range of values that results in just a single value expressing the significance of the accumulated data it is derived from. Aggregate functions are often used to derive descriptive statistics.

In base R, aggregation is achieved using the `aggregate` function, which according to its help page, computes summary statistics of data subsets. I wrote a [post](https://davetang.org/muse/2013/05/22/using-aggregate-and-apply-in-r/) on using the `aggregate` function because the function was not intuitive to me (at the time I wrote the post). In this post, I will use the `ChickWeight` dataset to illustrate aggregation. The `ChickWeight` data frame contains 578 rows and 4 columns from an experiment on the effect of diet on early growth of chicks. Use `?ChickWeight` to find out more about the dataset.

```{r chick_weight_dataset}
data("ChickWeight")
str(ChickWeight)
```

Groups of chicks were fed the same diet and most chicks had 12 measurements.

```{r table_diet_chick}
table(ChickWeight$Diet, ChickWeight$Chick)
```

These 12 measurements correspond to different timepoints (days since birth) where their weight was measured.

```{r chick_13}
ChickWeight[ChickWeight$Chick == 13, ]
```

Aggregating weight (using mean) as a function of diet can show us whether different diets resulted in different weights.

```{r weight_vs_diet}
aggregate(weight ~ Diet, data = ChickWeight, mean)
```

The `weight ~ Diet` expression is a [R formula](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html#formulas), which is commonly used to generate [design matrices](https://f1000research.com/articles/9-1444) but can be used as a general expression.

```{r formula}
class(weight ~ Diet)
```

The same expression can be used for boxplots.

```{r boxplot}
boxplot(weight ~ Diet, ChickWeight)
```

However, using R formula may not be intuitive and the following `dplyr` approach may make more sense, especially to those familiar with the [group by](https://en.wikipedia.org/wiki/Group_by_(SQL)) statement.
 
```{r summarise_weight}
group_by(ChickWeight, Diet) %>%
  summarise(weight = mean(weight))
```

To aggregate with two factors.

```{r agg_time_diet}
head(aggregate(weight ~ Diet + Time, data = ChickWeight, mean))
```

Using a `dplyr` approach.

```{r agg_time_diet_dplyr}
group_by(ChickWeight, Diet, Time) %>%
  summarise(weight = mean(weight)) %>%
  head()
```
 
Aggregating and calculating two summaries.

```{r agg_two_sum}
aggregate(
  weight ~ Diet,
  data = ChickWeight,
  FUN = function(x) c(mean = mean(x), n = length(x))
)
```

Using a `dplyr` approach.

```{r agg_two_sum_dplyr}
group_by(ChickWeight, Diet) %>%
  summarise(
    weight.mean = mean(weight),
    weight.n = length(weight)
  )
```

Aggregating on a data subset, for example only keeping chicks with 12 measurements.

```{r agg_subset}
chick_table <- table(ChickWeight$Chick)
my_keep <- as.integer(names(chick_table[chick_table == 12]))

aggregate(
  weight ~ Diet,
  data = subset(ChickWeight, Chick %in% my_keep),
  FUN = function(x) c(mean = mean(x), n = length(x))
)
```

Using a `dplyr` approach.

```{r agg_subset_dplyr}
ChickWeight %>%
  filter(Chick %in% my_keep) %>%
  group_by(Diet) %>%
  summarise(
    weight.mean = mean(weight),
    weight.n = length(weight)
  )
```

In summary, the `group_by` function from `dplyr` helps with carrying out aggregation functions within factors.
