---
title: "Normalisation methods implemented in edgeR"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

> Differential expression analysis of RNA-seq expression profiles with biological replication. Implements a range of statistical methodology based on the negative binomial distributions, including empirical Bayes estimation, exact tests, generalized linear models and quasi-likelihood tests. As well as RNA-seq, it be applied to differential signal analysis of other types of genomic data that produce read counts, including ChIP-seq, ATAC-seq, Bisulfite-seq, SAGE and CAGE. 

## Installation

Install using `BiocManager::install()`.

```{r install_edger, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("edgeR")
```

## Documentation

From `?edgeR::normLibSizes`:

> Calculate scaling factors to convert the raw library sizes for a set of sequenced samples into normalized effective library sizes.
>
> This function computes scaling factors to convert observed library sizes into normalized library sizes, also called "effective library sizes". The effective library sizes for use in downstream analysis are lib.size * norm.factors where lib.size contains the original library sizes and norm.factors is the vector of scaling factors computed by this function.
>
> The TMM method implements the trimmed mean of M-values method proposed by Robinson and Oshlack (2010). By default, the M-values are weighted according to inverse variances, as computed by the delta method for logarithms of binomial random variables. If refColumn is unspecified, then the column whose count-per-million upper quartile is closest to the mean upper quartile is set as the reference library.
>
> The TMMwsp method stands for "TMM with singleton pairing". This is a variant of TMM that is intended to have more stable performance when the counts have a high proportion of zeros. In the TMM method, genes that have zero count in either library are ignored when comparing pairs of libraries. In the TMMwsp method, the positive counts from such genes are reused to increase the number of features by which the libraries are compared. The singleton positive counts are paired up between the libraries in decreasing order of size and then a slightly modified TMM method is applied to the re-ordered libraries. If refColumn is unspecified, then the column with largest sum of square-root counts is used as the reference library.
>
> RLE is the scaling factor method proposed by Anders and Huber (2010). We call it "relative log expression", as median library is calculated from the geometric mean of all columns and the median ratio of each sample to the median library is taken as the scale factor.
>
> The upperquartile method is the upper-quartile normalization method of Bullard et al (2010), in which the scale factors are calculated from the 75% quantile of the counts for each library, after removing genes that are zero in all libraries. The idea is generalized here to allow normalization by any quantile of the count distributions.
>
> If method="none", then the normalization factors are set to 1.
>
> For symmetry, normalization factors are adjusted to multiply to 1. Rows of object that have zero counts for all columns are removed before normalization factors are computed. The number of such rows does not affect the estimated normalization factors. 

## A test dataset

I created a dataset to test the different normalisation methods. There are four samples; column one and two are the controls and column three and four are the patients. 25 transcripts are in all four samples in equal amount. Another 25 transcripts are **only** present in the controls and their expression is the same as the first 25 transcripts in the controls. The four samples have exactly the same depth (500 counts). However, since the patient samples have half the number of transcripts than the controls (25 versus 50), they are sequenced at twice the depth. This hypothetical situation was described in [Robinson and Oshlack Genome Biology 2010](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25).

```{r create_df}
control_1 <- rep(10, 50)
control_2 <- rep(10, 50)
patient_1 <- c(rep(20, 25),rep(0,25))
patient_2 <- c(rep(20, 25),rep(0,25))

df <- data.frame(c1=control_1,
                 c2=control_2,
                 p1=patient_1,
                 p2=patient_2)

head(df)
```

Tail of the dataset.

```{r tail_df}
tail(df)
```

Equal depth.

```{r depth_df}
colSums(df)
```

## No normalisation

Let's run the differential expression analysis without any normalisation step:

```{r no_norm}
library(edgeR)

# create group vector
group <- c('control','control','patient','patient')

# create DGEList object
d <- DGEList(counts=df, group=group)

# check out the DGEList object
d
```

Perform a differential gene expression analysis.

```{r no_norm_de}
d <- DGEList(counts=df, group=group)
d <- estimateCommonDisp(d)

# perform the DE test
de <- exactTest(d)

# how many differentially expressed transcripts?
table(p.adjust(de$table$PValue, method="BH")<0.05)
```

Without normalisation, every transcript is differentially expressed.

## TMM normalisation

From the [edgeR manual](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf):

{edgeR} is concerned with differential expression analysis rather than with the quantification of expression levels. It is concerned with relative changes in expression levels between conditions, but not directly with estimating absolute expression levels.

The `normLibSizes()` function normalises the library sizes in such a way to minimise the log-fold changes between the samples for most genes. The default method for computing these scale factors uses a trimmed mean of M-values (TMM) between each pair of samples. We call the product of the original library size and the scaling factor the effective library size, i.e., the normalised library size. The effective library size replaces the original library size in all downstream analyses

Let's test the weighted trimmed mean of M-values method:

```{r d_tmm}
d_tmm <- normLibSizes(d, method="TMM")
d_tmm
```

If we use the scaling, we will observe that the differences are now less pronounced.

```{r head_d_tmm}
head(d$counts)
head(d_tmm$counts * d_tmm$samples$norm.factors)
```

Perform the differential expression test.

```{r d_tmm_de}
d_tmm <- estimateCommonDisp(d_tmm)
d_tmm <- exactTest(d_tmm)
table(p.adjust(d_tmm$table$PValue, method="BH")<0.05)
```

Only half of the transcripts are differentially expressed (the last 25 transcripts).

## RLE normalisation

Let's test the relative log expression normalisation method:

```{r d_rle}
d_rle <- normLibSizes(d, method="RLE")
d_rle
```

Perform the differential gene expression analysis.

```{r d_rle_de}
d_rle <- estimateCommonDisp(d_rle)
d_rle <- exactTest(d_rle)
table(p.adjust(d_rle$table$PValue, method="BH")<0.05)
```

## Upper-quartile normalisation

Lastly let's try the upper-quartile normalisation method:

```{r d_uq}
d_uq <- normLibSizes(d, method="upperquartile")
d_uq
```

DE test.

```{r d_uq_de}
d_uq <- estimateCommonDisp(d_uq)
d_uq <- exactTest(d_uq)
table(p.adjust(d_uq$table$PValue, method="BH")<0.05)
```

## Testing a real dataset

Perform differential gene expression analysis using various normalisation methods on the [pnas_expression.txt](https://pubmed.ncbi.nlm.nih.gov/19088194/) dataset.

```{r data}
my_url <- "https://davetang.org/file/pnas_expression.txt"
data <- read.table(my_url, header=TRUE, sep="\t")

dim(data)
```

Prepare a DGEList object.

```{r data_d}
d <- data[,2:8]
rownames(d) <- data[,1]
group <- c(rep("Control",4),rep("DHT",3))
d <- DGEList(counts = d, group=group)
```

Filter out lowly expressed genes.

```{r keep}
keep <- rowSums(cpm(d) > 0.5) >= 2
d <- d[keep, , keep.lib.sizes=FALSE]
```

Function to run edgeR.

```{r run_edger}
run_edger <- function(d, method = NULL){
  if(!is.null(method)){
    d <- normLibSizes(d, method = method)
  }
  design <- model.matrix(~d$samples$group)
  d <- estimateDisp(d, design)
  # use Quasi-likelihood F-tests
  fit <- glmQLFit(d, design)
  qlf <- glmQLFTest(fit)
}
```

Carry out differential gene expression analysis with no normalisation.

```{r run_edger_no_norm}
no_norm <- run_edger(d)
table(p.adjust(no_norm$table$PValue, method="fdr")<0.05)
```

With TMM normalisation.

```{r run_edger_tmm}
TMM <- run_edger(d, method="TMM")
table(p.adjust(TMM$table$PValue, method="fdr")<0.05)
```

With RLE.

```{r run_edger_rle}
RLE <- run_edger(d, method="RLE")
table(p.adjust(RLE$table$PValue, method="fdr")<0.05)
```

With the upper quartile method.

```{r run_edger_uq}
uq <- run_edger(d, method="upperquartile")
table(p.adjust(uq$table$PValue, method="fdr")<0.05)
```

Finding the overlaps between the differential gene expression analyses.

```{r run_edger_overlap}
library(gplots)

get_de <- function(x, pvalue){
  my_i <- p.adjust(x$PValue, method="fdr") < pvalue
  row.names(x)[my_i]
}

my_de_no_norm <- get_de(no_norm$table, 0.05)
my_de_tmm <- get_de(TMM$table, 0.05)
my_de_rle <- get_de(RLE$table, 0.05)
my_de_uq <- get_de(uq$table, 0.05)

gplots::venn(list(no_norm = my_de_no_norm, TMM = my_de_tmm, RLE = my_de_rle, UQ = my_de_uq))
```

There is a large overlap of differentially expressed genes in the different normalisation approaches.

```{r run_edger_norm}
gplots::venn(list(TMM = my_de_tmm, RLE = my_de_rle, UQ = my_de_uq))
```

## Summary

The normalisation factors were quite similar between all normalisation methods, which is why the results of the differential expression were quite concordant. Most methods down sized the DHT samples with a normalisation factor of less than one to account for the larger library sizes of these samples.
