---
title: "The Expectation-Maximisation (EM) Algorithm"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r packages, message=FALSE}
library(ggplot2)
```

# Introduction

The **Expectation-Maximisation (EM) algorithm** is one of the most widely used tools in statistics and machine learning. It shows up in Gaussian mixture models, hidden Markov models, missing-data problems, and many other settings. But most explanations of EM jump into abstract notation very quickly, which makes it feel more complicated than it actually is.

The core idea is simple. EM is for situations where you are trying to estimate something, but there is **hidden information** that, if you knew it, would make the problem easy. Since you don't know the hidden information, you alternate between:

1. **Guessing** the hidden information (using your current parameter estimates).
2. **Updating** the parameter estimates (using your guesses of the hidden information).

Repeat until things stabilise. That's it. The rest of this article is about building solid intuition for what this means and why it works.

# An Everyday Analogy

Imagine you are a teacher with two teaching assistants, Alice and Bob, who grade homework. You receive a stack of graded assignments but their names have fallen off — you can see the scores each student received, but you do not know which TA graded which assignment.

You want to figure out each TA's grading tendency (say, the average score they give). If you **knew** who graded each assignment, this would be trivial: split the assignments into two piles and compute the average for each pile.

But you don't know who graded what. So you do the following:

1. **Start with a guess**: maybe Alice grades generously (average ~85) and Bob grades strictly (average ~70).
2. **E-step**: For each assignment, ask: "Given these guessed averages, which TA probably graded this one?" An assignment with a score of 90 is more likely from Alice; a score of 65 is more likely from Bob. A score of 78 could go either way, so you assign **partial credit** — maybe 60% chance Alice, 40% chance Bob.
3. **M-step**: Using these soft assignments, recompute the averages. Alice's new average is the weighted mean of all scores, weighted by how likely she was to have graded each one. Same for Bob.
4. **Repeat** steps 2 and 3 until the averages stop changing.

That is the EM algorithm. The hidden information is "who graded what", and the parameters are the grading averages.

# The Two-Coin Problem

The classic teaching example for EM involves two coins. This problem is small enough to work through by hand, which makes the mechanics completely transparent.

## Setup

You have two coins, **A** and **B**, with unknown probabilities of landing heads:

- Coin A lands heads with probability $\theta_A$ (unknown).
- Coin B lands heads with probability $\theta_B$ (unknown).

Someone runs several rounds of an experiment. In each round, they **secretly pick** one of the two coins (you don't know which), flip it 10 times, and tell you how many heads came up. Your job: estimate $\theta_A$ and $\theta_B$.

Let's simulate this. We'll set the true biases to $\theta_A = 0.8$ and $\theta_B = 0.3$, generate the data, and then see how close EM can get — pretending we don't know the true values.

```{r two_coins_setup}
set.seed(1984)

# True (hidden) parameters
true_theta_A <- 0.8
true_theta_B <- 0.3

# Simulate 10 rounds: randomly pick a coin, flip it 10 times
n_rounds <- 10
n_flips <- 10

coin_used <- sample(c("A", "B"), n_rounds, replace = TRUE)
heads <- ifelse(
  coin_used == "A",
  rbinom(n_rounds, size = n_flips, prob = true_theta_A),
  rbinom(n_rounds, size = n_flips, prob = true_theta_B)
)
```

Here is what the experimenter tells us — just the number of heads per round. The coin column is hidden from us.

```{r two_coins_table}
data.frame(
  Round = seq_len(n_rounds),
  Heads = heads,
  Coin_used = paste0(coin_used, " (hidden)")
)
```

If we knew the hidden column, we could just compute the proportion of heads for each coin. Since we don't, we use EM.

## Step by Step

Before we look at the code, let's unpack the key ideas that the E-step and M-step rely on.

### Probability vs Likelihood

These two words sound interchangeable, but they ask different questions:

- **Probability** asks: given that I know the coin has bias $\theta = 0.8$, what is the chance of seeing 7 heads in 10 flips? The parameter is fixed and we are asking about the data.
- **Likelihood** asks: given that I observed 7 heads in 10 flips, how plausible is it that the coin has bias $\theta = 0.8$? The data is fixed and we are asking about the parameter.

Mathematically, the same formula is used for both — the binomial probability mass function `dbinom()`. The difference is in what we treat as known and what we treat as unknown. When we hold the data fixed and vary the parameter, we call the result the **likelihood**.

### Conditional Probability and Bayes' Rule

In the E-step, we need to answer: "Given the heads count in this round **and** our current guesses for $\theta_A$ and $\theta_B$, what is the probability that coin A was used?" This is a **conditional probability** — the probability of one event (coin A was used) given that another event has occurred (we observed a specific number of heads).

Bayes' rule gives us the answer:

$$P(\text{coin A} \mid \text{data}) = \frac{P(\text{data} \mid \text{coin A}) \times P(\text{coin A})}{P(\text{data} \mid \text{coin A}) \times P(\text{coin A}) + P(\text{data} \mid \text{coin B}) \times P(\text{coin B})}$$

Breaking this down:

- $P(\text{data} \mid \text{coin A})$ is the **likelihood**: how probable is this round's outcome if coin A was used? We compute this with `dbinom(heads, size = n_flips, prob = theta_A)`.
- $P(\text{coin A})$ is the **prior**: before looking at the data, how likely is it that coin A was picked? We assume each coin is equally likely, so $P(\text{coin A}) = P(\text{coin B}) = 0.5$. Because the priors are equal, they cancel out and the formula simplifies to:

$$P(\text{coin A} \mid \text{data}) = \frac{P(\text{data} \mid \text{coin A})}{P(\text{data} \mid \text{coin A}) + P(\text{data} \mid \text{coin B})}$$

This is just "coin A's likelihood divided by the total likelihood". If coin A makes the observed data twice as likely as coin B, then coin A gets twice the credit.

### Responsibility

The result of this calculation — $P(\text{coin A} \mid \text{data})$ — is called the **responsibility**. It is a number between 0 and 1 for each round. A responsibility of 0.9 for coin A means: "given our current parameter guesses, we are 90% confident coin A was used in this round."

### The Weighted Update (M-step)

Once we have responsibilities, we update the bias estimates. The M-step computes:

$$\theta_A^{\text{new}} = \frac{\sum_{i} r_{A,i} \times h_i}{\sum_{i} r_{A,i} \times n}$$

where $r_{A,i}$ is the responsibility of coin A for round $i$, $h_i$ is the number of heads, and $n$ is the number of flips per round.

What does this mean? Consider two extremes:

- If $r_{A,i} = 1$ for some rounds and 0 for others, this becomes the ordinary proportion of heads across only the rounds assigned to coin A. This is what you would compute if you **knew** the labels.
- If $r_{A,i} = 0.7$ for a round, that round contributes to coin A's estimate with 70% weight and to coin B's with 30%. Each round "splits" its evidence between the two coins according to how likely each coin was.

In this way, the responsibilities act as **soft labels**: instead of assigning each round to exactly one coin, they spread each round's data proportionally across both coins.

### Initial Guess

We need a starting point. Let's guess $\theta_A = 0.6$ and $\theta_B = 0.5$.

```{r two_coins_em}
# EM for the two-coin problem
em_two_coins <- function(heads, n_flips, theta_A_init, theta_B_init,
                         max_iter = 20, tol = 1e-6) {

  theta_A <- theta_A_init
  theta_B <- theta_B_init
  n_rounds <- length(heads)
  history <- data.frame(
    iteration = integer(),
    theta_A = numeric(),
    theta_B = numeric()
  )
  history <- rbind(history, data.frame(
    iteration = 0, theta_A = theta_A, theta_B = theta_B
  ))

  for (iter in seq_len(max_iter)) {
    # --- E-step ---
    # For each round, compute the probability that coin A was used
    # P(coin A | data) is proportional to P(data | coin A) * P(coin A)
    # Assume equal prior: P(coin A) = P(coin B) = 0.5

    # Likelihood of observing the data under each coin (binomial)
    lik_A <- dbinom(heads, size = n_flips, prob = theta_A)
    lik_B <- dbinom(heads, size = n_flips, prob = theta_B)

    # Responsibility: probability that coin A generated each round
    resp_A <- lik_A / (lik_A + lik_B)
    resp_B <- 1 - resp_A

    # --- M-step ---
    # Update theta_A: weighted average of head proportions, weighted by resp_A
    theta_A_new <- sum(resp_A * heads) / sum(resp_A * n_flips)
    theta_B_new <- sum(resp_B * heads) / sum(resp_B * n_flips)

    history <- rbind(history, data.frame(
      iteration = iter, theta_A = theta_A_new, theta_B = theta_B_new
    ))

    # Check convergence
    if (abs(theta_A_new - theta_A) < tol && abs(theta_B_new - theta_B) < tol) {
      theta_A <- theta_A_new
      theta_B <- theta_B_new
      break
    }

    theta_A <- theta_A_new
    theta_B <- theta_B_new
  }

  list(
    theta_A = theta_A,
    theta_B = theta_B,
    iterations = iter,
    history = history
  )
}

result <- em_two_coins(heads, n_flips,
                       theta_A_init = 0.6, theta_B_init = 0.5)

cat(sprintf("Converged in %d iterations\n\n", result$iterations))
cat(sprintf("  Estimated: theta_A = %.4f   True: %.4f\n", result$theta_A, true_theta_A))
cat(sprintf("  Estimated: theta_B = %.4f   True: %.4f\n", result$theta_B, true_theta_B))
```

With only `r n_rounds` rounds of data, the estimates may not be perfect — but they should be in the right ballpark. EM correctly identifies which coin is biased toward heads and which toward tails.

### What Happened at Each Iteration

Let's look at the first few iterations in detail to see the E-step and M-step playing out.

```{r first_iterations}
# Show detailed E-step and M-step for the first 3 iterations
theta_A <- 0.6
theta_B <- 0.5

for (iter in 1:3) {
  cat(sprintf("=== Iteration %d ===\n", iter))
  cat(sprintf("Current guesses: theta_A = %.4f, theta_B = %.4f\n\n", theta_A, theta_B))

  lik_A <- dbinom(heads, size = n_flips, prob = theta_A)
  lik_B <- dbinom(heads, size = n_flips, prob = theta_B)
  resp_A <- lik_A / (lik_A + lik_B)

  cat("E-step — responsibility of coin A for each round:\n")
  for (i in seq_along(heads)) {
    cat(sprintf("  Round %d (%d heads): P(coin A) = %.3f, P(coin B) = %.3f\n",
                i, heads[i], resp_A[i], 1 - resp_A[i]))
  }

  theta_A <- sum(resp_A * heads) / sum(resp_A * n_flips)
  theta_B <- sum((1 - resp_A) * heads) / sum((1 - resp_A) * n_flips)

  cat(sprintf("\nM-step — updated estimates: theta_A = %.4f, theta_B = %.4f\n\n", theta_A, theta_B))
}
```

Notice how the responsibilities shift over iterations. In the beginning, the guesses are close together so the model is unsure which coin generated which round. As the estimates improve, the model becomes more confident: rounds with many heads are assigned more strongly to the higher-probability coin and vice versa.

### Convergence Plot

```{r coins_convergence, fig.width=8, fig.height=5, fig.cap="Convergence of coin bias estimates across EM iterations. Dashed lines show the true values."}
history_long <- data.frame(
  iteration = rep(result$history$iteration, 2),
  theta = c(result$history$theta_A, result$history$theta_B),
  coin = rep(c("Coin A", "Coin B"), each = nrow(result$history))
)

ggplot(history_long, aes(x = iteration, y = theta, colour = coin)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = true_theta_A, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = true_theta_B, linetype = "dashed", colour = "grey40") +
  annotate("text", x = max(history_long$iteration), y = true_theta_A + 0.03,
           label = paste0("True theta_A = ", true_theta_A),
           hjust = 1, colour = "grey40") +
  annotate("text", x = max(history_long$iteration), y = true_theta_B - 0.03,
           label = paste0("True theta_B = ", true_theta_B),
           hjust = 1, colour = "grey40") +
  labs(x = "Iteration", y = "Estimated Bias (probability of heads)",
       colour = "", title = "EM convergence for the two-coin problem") +
  theme_minimal()
```

The two estimates start close together and gradually separate as the algorithm figures out which rounds go with which coin. The dashed lines show the true biases we used to simulate the data.

## More Data, Better Estimates

With only `r n_rounds` rounds, the estimates are reasonable but not exact. What happens if we give EM more data? Just as you would get a better estimate of a coin's bias by flipping it more times, EM gets closer to the truth with more rounds.

```{r more_rounds, fig.width=9, fig.height=5, fig.cap="EM estimates improve as the number of rounds increases. With enough data, the estimates converge to the true values."}
round_sizes <- c(10, 25, 50, 100, 250, 500)
n_reps <- 50  # repeat each size to show variability

sim_results <- do.call(rbind, lapply(round_sizes, function(nr) {
  do.call(rbind, lapply(seq_len(n_reps), function(rep) {
    coin <- sample(c("A", "B"), nr, replace = TRUE)
    h <- ifelse(
      coin == "A",
      rbinom(nr, size = n_flips, prob = true_theta_A),
      rbinom(nr, size = n_flips, prob = true_theta_B)
    )
    fit <- em_two_coins(h, n_flips, theta_A_init = 0.6, theta_B_init = 0.5)

    # EM may swap the labels — assign higher estimate to coin A
    tA <- max(fit$theta_A, fit$theta_B)
    tB <- min(fit$theta_A, fit$theta_B)

    data.frame(n_rounds = nr, rep = rep, theta_A = tA, theta_B = tB)
  }))
}))

sim_long <- data.frame(
  n_rounds = rep(sim_results$n_rounds, 2),
  estimate = c(sim_results$theta_A, sim_results$theta_B),
  coin = rep(c("Coin A", "Coin B"), each = nrow(sim_results))
)

ggplot(sim_long, aes(x = factor(n_rounds), y = estimate, fill = coin)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  geom_hline(yintercept = true_theta_A, linetype = "dashed", colour = "grey30") +
  geom_hline(yintercept = true_theta_B, linetype = "dashed", colour = "grey30") +
  annotate("text", x = 6.4, y = true_theta_A + 0.02,
           label = paste0("True = ", true_theta_A), colour = "grey30", hjust = 1) +
  annotate("text", x = 6.4, y = true_theta_B - 0.02,
           label = paste0("True = ", true_theta_B), colour = "grey30", hjust = 1) +
  labs(x = "Number of Rounds", y = "Estimated Bias",
       fill = "", title = "EM estimates improve with more data") +
  theme_minimal()
```

With 10 rounds, the estimates are noisy — each repetition of the experiment gives a somewhat different answer. By 100 rounds, the boxes have tightened considerably. By 500 rounds, the estimates are tightly clustered around the true values. This is a general property: EM is a **consistent** estimator, meaning it converges to the truth given enough data.

# The Grading Example — Made Concrete

Earlier we described a teacher trying to figure out whether Alice or Bob graded each assignment, given only the scores. Let's turn that analogy into a full worked example with simulated data.

## Setting the Scene

Alice is a generous grader — her scores tend to cluster around 82 with some spread. Bob is stricter — his scores cluster around 65. Over the semester, they graded 200 assignments between them (Alice graded about 55%, Bob about 45%). Unfortunately the TA names fell off, so all we have is a pile of 200 scores.

```{r grading_simulate}
set.seed(1984)

n_alice <- 110
n_bob <- 90

alice_scores <- rnorm(n_alice, mean = 82, sd = 7)
bob_scores <- rnorm(n_bob, mean = 65, sd = 6)

# What we observe: a single pile of scores with no labels
scores <- c(alice_scores, bob_scores)
true_grader <- c(rep("Alice", n_alice), rep("Bob", n_bob))
```

```{r grading_hist, fig.width=9, fig.height=5, fig.cap="The pile of 200 scores. Two overlapping humps hint at two graders, but we don't know who graded what."}
ggplot(data.frame(score = scores), aes(x = score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", colour = "white", alpha = 0.7) +
  geom_density(colour = "red", linewidth = 0.8) +
  labs(x = "Score", y = "Density",
       title = "All 200 assignment scores (grader unknown)") +
  theme_minimal()
```

The histogram shows a hint of two humps. EM will try to recover the two grading distributions.

## Running EM

The parameters we want to estimate are:

- $\mu_A$, $\sigma_A$: Alice's average score and spread.
- $\mu_B$, $\sigma_B$: Bob's average score and spread.
- $\pi_A$: the proportion of assignments Alice graded (and $\pi_B = 1 - \pi_A$).

We start with a rough guess: grader 1 averages around 75 and grader 2 averages around 70 — intentionally mediocre starting values.

```{r grading_em}
em_grading <- function(scores, max_iter = 50, tol = 1e-6) {
  n <- length(scores)

  # Initial guesses (deliberately rough)
  mu <- c(75, 70)
  sigma <- c(10, 10)
  pi_k <- c(0.5, 0.5)

  history <- list()

  for (iter in seq_len(max_iter)) {
    # E-step: for each score, how likely is it from grader 1 vs grader 2?
    lik_1 <- pi_k[1] * dnorm(scores, mean = mu[1], sd = sigma[1])
    lik_2 <- pi_k[2] * dnorm(scores, mean = mu[2], sd = sigma[2])
    resp_1 <- lik_1 / (lik_1 + lik_2)
    resp_2 <- 1 - resp_1

    # M-step: re-estimate parameters using responsibilities as weights
    N_1 <- sum(resp_1)
    N_2 <- sum(resp_2)

    mu_new <- c(sum(resp_1 * scores) / N_1,
                sum(resp_2 * scores) / N_2)
    sigma_new <- c(sqrt(sum(resp_1 * (scores - mu_new[1])^2) / N_1),
                   sqrt(sum(resp_2 * (scores - mu_new[2])^2) / N_2))
    pi_new <- c(N_1 / n, N_2 / n)

    history[[iter]] <- data.frame(
      iteration = iter,
      grader = c("Grader 1", "Grader 2"),
      mu = mu_new,
      sigma = sigma_new,
      weight = pi_new
    )

    if (max(abs(mu_new - mu)) < tol) break
    mu <- mu_new
    sigma <- sigma_new
    pi_k <- pi_new
  }

  list(mu = mu_new, sigma = sigma_new, pi_k = pi_new,
       resp_1 = resp_1, history = do.call(rbind, history),
       iterations = iter)
}

fit_grade <- em_grading(scores)

cat(sprintf("Converged in %d iterations\n\n", fit_grade$iterations))
cat("Estimated parameters:\n")
cat(sprintf("  Grader 1: mean = %.1f, sd = %.1f, proportion = %.0f%%\n",
            fit_grade$mu[1], fit_grade$sigma[1], fit_grade$pi_k[1] * 100))
cat(sprintf("  Grader 2: mean = %.1f, sd = %.1f, proportion = %.0f%%\n",
            fit_grade$mu[2], fit_grade$sigma[2], fit_grade$pi_k[2] * 100))
cat("\nTrue parameters:\n")
cat(sprintf("  Alice:    mean = 82.0, sd = 7.0, proportion = 55%%\n"))
cat(sprintf("  Bob:      mean = 65.0, sd = 6.0, proportion = 45%%\n"))
```

EM recovers values close to the truth, despite starting with guesses that were well off.

## Watching It Converge

```{r grading_convergence, fig.width=9, fig.height=5, fig.cap="The estimated grading averages converge from the initial guesses (75 and 70) toward the true values."}
ggplot(fit_grade$history, aes(x = iteration, y = mu, colour = grader)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 82, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 65, linetype = "dashed", colour = "grey40") +
  annotate("text", x = max(fit_grade$history$iteration), y = 83.5,
           label = "Alice's true mean (82)", hjust = 1, colour = "grey40") +
  annotate("text", x = max(fit_grade$history$iteration), y = 63.5,
           label = "Bob's true mean (65)", hjust = 1, colour = "grey40") +
  labs(x = "Iteration", y = "Estimated Mean Score",
       colour = "", title = "EM convergence for the grading problem") +
  theme_minimal()
```

The two estimates start close together (75 and 70) and separate rapidly as EM figures out the two grading styles.

## Looking at Individual Assignments

After convergence, each assignment has a responsibility — the probability that grader 1 (Alice) gave that score. Let's look at a few assignments to see how EM reasons about them.

```{r grading_individual}
# Pick a few illustrative assignments
examples <- c(
  which.max(scores),                          # highest score
  which.min(scores),                          # lowest score
  which.min(abs(scores - mean(scores)))       # most ambiguous (near overall mean)
)

cat("How EM assigns individual assignments:\n\n")
for (i in examples) {
  cat(sprintf(
    "  Score = %.1f → P(Alice) = %.1f%%, P(Bob) = %.1f%%  [truly graded by %s]\n",
    scores[i],
    fit_grade$resp_1[i] * 100,
    (1 - fit_grade$resp_1[i]) * 100,
    true_grader[i]
  ))
}
```

High scores are confidently assigned to Alice (the generous grader), low scores to Bob (the strict grader), and middle scores are genuinely ambiguous.

```{r grading_assignments, fig.width=9, fig.height=5, fig.cap="Each assignment coloured by EM's confidence that Alice graded it. Scores in the overlap region (around 70-75) are genuinely uncertain."}
df_assign <- data.frame(
  score = scores,
  p_alice = fit_grade$resp_1,
  true_grader = true_grader
)

ggplot(df_assign, aes(x = score, y = p_alice, colour = true_grader)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  labs(x = "Score", y = "P(Alice graded this)",
       colour = "True Grader",
       title = "EM's soft assignments vs the truth") +
  theme_minimal()
```

Points above the dashed line are assigned to Alice; points below to Bob. The colour shows the truth. Most assignments are classified correctly, with mistakes concentrated in the overlap region where the two graders' score distributions blend together — exactly where you would expect genuine uncertainty.

## How Accurate Are the Hard Assignments?

If we assign each paper to whichever grader has the higher responsibility, how often do we get it right?

```{r grading_accuracy}
hard_assignment <- ifelse(fit_grade$resp_1 > 0.5, "Alice", "Bob")
accuracy <- mean(hard_assignment == true_grader)
cat(sprintf("Classification accuracy: %.1f%%\n", accuracy * 100))

cat("\nConfusion matrix:\n")
print(table(Predicted = hard_assignment, True = true_grader))
```

The errors are concentrated among scores in the 70--78 range where Alice's low scores and Bob's high scores overlap. No method can perfectly separate those, because the data genuinely looks the same regardless of which grader produced it.

# EM for Gaussian Mixtures

The two-coin problem illustrates the logic of EM in a discrete setting. The same logic applies to continuous data. The most common application is fitting **Gaussian mixture models** (GMMs), where you assume your data comes from a blend of several normal (bell-curve) distributions and you want to recover their parameters.

For a detailed treatment of GMMs, see the [Gaussian mixture model](gmm.html) article. Here we focus on the EM mechanics.

## Simulated Data

```{r simulate_gmm}
set.seed(1984)

# Two-component Gaussian mixture
n <- 800
n1 <- round(n * 0.65)
n2 <- n - n1

x <- c(rnorm(n1, mean = 2, sd = 0.8),
       rnorm(n2, mean = 6, sd = 1.0))

true_labels <- c(rep(1, n1), rep(2, n2))
```

```{r show_data, fig.width=9, fig.height=4, fig.cap="Simulated data from two overlapping Gaussian components."}
ggplot(data.frame(x = x), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "steelblue", colour = "white", alpha = 0.7) +
  geom_density(colour = "red", linewidth = 0.8) +
  labs(x = "Value", y = "Density",
       title = "Observed data from an unknown mixture") +
  theme_minimal()
```

We know this data was generated by two Gaussians, but we will pretend we don't and let EM figure it out.

## Watching EM Iterate

To see exactly what EM is doing, we'll record the state at each iteration and plot it.

```{r em_gmm_detailed}
em_gaussian <- function(x, K = 2, max_iter = 30, tol = 1e-6) {
  n <- length(x)

  # Initialise: spread means across the data range
  mu <- quantile(x, probs = seq(0.3, 0.7, length.out = K))
  sigma <- rep(sd(x), K)
  pi_k <- rep(1 / K, K)

  snapshots <- list()

  for (iter in seq_len(max_iter)) {
    # E-step
    resp <- matrix(0, nrow = n, ncol = K)
    for (k in seq_len(K)) {
      resp[, k] <- pi_k[k] * dnorm(x, mean = mu[k], sd = sigma[k])
    }
    resp <- resp / rowSums(resp)

    # M-step
    N_k <- colSums(resp)
    mu_new <- colSums(resp * x) / N_k
    sigma_new <- sqrt(colSums(resp * outer(x, mu_new, "-")^2) / N_k)
    pi_new <- N_k / n

    # Record snapshot
    snapshots[[iter]] <- list(
      iter = iter, mu = mu_new, sigma = sigma_new,
      pi_k = pi_new, resp = resp
    )

    if (max(abs(mu_new - mu)) < tol) break
    mu <- mu_new
    sigma <- sigma_new
    pi_k <- pi_new
  }

  list(mu = mu_new, sigma = sigma_new, pi_k = pi_new,
       resp = resp, snapshots = snapshots, iterations = iter)
}

fit <- em_gaussian(x, K = 2)
cat(sprintf("Converged in %d iterations\n\n", fit$iterations))
cat(sprintf("Component 1: mean = %.2f, sd = %.2f, weight = %.2f\n",
            fit$mu[1], fit$sigma[1], fit$pi_k[1]))
cat(sprintf("Component 2: mean = %.2f, sd = %.2f, weight = %.2f\n",
            fit$mu[2], fit$sigma[2], fit$pi_k[2]))
```

```{r em_iterations_plot, fig.width=10, fig.height=10, fig.cap="EM at iterations 1, 3, 5, and after convergence. The two Gaussian components (orange and blue curves) gradually separate and fit the data."}
# Select a few iterations to visualise
show_iters <- c(1, 3, 5, fit$iterations)
show_iters <- unique(pmin(show_iters, fit$iterations))

x_grid <- seq(min(x) - 1, max(x) + 1, length.out = 500)

plot_data <- do.call(rbind, lapply(show_iters, function(i) {
  s <- fit$snapshots[[i]]
  data.frame(
    x = rep(x_grid, 3),
    density = c(
      s$pi_k[1] * dnorm(x_grid, s$mu[1], s$sigma[1]),
      s$pi_k[2] * dnorm(x_grid, s$mu[2], s$sigma[2]),
      s$pi_k[1] * dnorm(x_grid, s$mu[1], s$sigma[1]) +
        s$pi_k[2] * dnorm(x_grid, s$mu[2], s$sigma[2])
    ),
    curve = rep(c("Component 1", "Component 2", "Mixture"), each = length(x_grid)),
    iteration = paste("Iteration", i)
  )
}))

plot_data$iteration <- factor(plot_data$iteration,
                              levels = paste("Iteration", show_iters))

ggplot() +
  geom_histogram(data = data.frame(x = x), aes(x = x, y = after_stat(density)),
                 bins = 50, fill = "grey80", colour = "white") +
  geom_line(data = plot_data, aes(x = x, y = density, colour = curve),
            linewidth = 0.9) +
  scale_colour_manual(values = c("Component 1" = "#E69F00",
                                  "Component 2" = "#56B4E9",
                                  "Mixture" = "red")) +
  facet_wrap(~iteration, ncol = 2) +
  labs(x = "Value", y = "Density", colour = "",
       title = "EM algorithm iteration by iteration") +
  theme_minimal()
```

At iteration 1, the two components are poorly placed — they are too close together and too wide. By iteration 3, they have started to separate. By convergence, each component sits neatly over one of the two humps and the combined mixture (red) traces the overall shape of the data.

## How Responsibilities Evolve

The E-step assigns each data point a responsibility (a probability between 0 and 1) for each component. Let's see how these responsibilities sharpen over iterations.

```{r responsibilities_evolve, fig.width=10, fig.height=8, fig.cap="Responsibilities at different iterations. Early on (top), the model is uncertain about most points. By convergence (bottom), assignments are crisp except in the overlap region."}
resp_iters <- c(1, 3, fit$iterations)
resp_iters <- unique(pmin(resp_iters, fit$iterations))

resp_data <- do.call(rbind, lapply(resp_iters, function(i) {
  data.frame(
    x = x,
    resp_2 = fit$snapshots[[i]]$resp[, 2],
    iteration = paste("Iteration", i)
  )
}))

resp_data$iteration <- factor(resp_data$iteration,
                              levels = paste("Iteration", resp_iters))

ggplot(resp_data, aes(x = x, y = resp_2)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "steelblue") +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  facet_wrap(~iteration, ncol = 1) +
  labs(x = "Value", y = "P(Component 2)",
       title = "How responsibilities sharpen over iterations") +
  theme_minimal()
```

The dashed line at 0.5 marks the decision boundary. Points above the line are assigned to component 2; points below to component 1. In early iterations, many points cluster near 0.5 (the model is uncertain). By convergence, most points are pushed close to 0 or 1, with only the overlap region retaining genuine ambiguity.

# Why Does EM Work?

You might wonder: why does this iterative guessing game converge to a good answer instead of going in circles? There is a mathematical guarantee.

## The Log-Likelihood Never Decreases

At each iteration, EM is guaranteed to **increase** (or at least not decrease) the **log-likelihood** of the data. The log-likelihood measures how well the current parameters explain the observed data. Think of it as a score: higher is better.

```{r log_likelihood_trace, fig.width=8, fig.height=5, fig.cap="The log-likelihood increases at every EM iteration and plateaus at convergence."}
# Compute log-likelihood at each iteration
compute_ll <- function(x, mu, sigma, pi_k) {
  K <- length(mu)
  ll <- 0
  for (i in seq_along(x)) {
    p <- 0
    for (k in seq_len(K)) {
      p <- p + pi_k[k] * dnorm(x[i], mu[k], sigma[k])
    }
    ll <- ll + log(p)
  }
  ll
}

ll_values <- sapply(fit$snapshots, function(s) {
  compute_ll(x, s$mu, s$sigma, s$pi_k)
})

df_ll <- data.frame(iteration = seq_along(ll_values), log_likelihood = ll_values)

ggplot(df_ll, aes(x = iteration, y = log_likelihood)) +
  geom_line(linewidth = 1, colour = "darkgreen") +
  geom_point(size = 2, colour = "darkgreen") +
  labs(x = "Iteration", y = "Log-Likelihood",
       title = "Log-likelihood increases at every EM iteration") +
  theme_minimal()
```

The curve goes up steeply at first (big improvements) then levels off (fine-tuning). EM stops when the improvements become negligibly small.

## Local Optima

One important caveat: EM is guaranteed to find **a** peak in the log-likelihood, but not necessarily the **highest** peak. The log-likelihood surface can have multiple peaks (local optima), and which one EM finds depends on where it starts.

This means **initialisation matters**. Let's demonstrate this by running EM with two different starting points.

```{r local_optima, fig.width=9, fig.height=5, fig.cap="Different starting points can lead to different solutions. Here, a poor initialisation gets stuck."}
# Good initialisation
fit_good <- em_gaussian(x, K = 2)

# Force a bad initialisation by hacking the function
em_gaussian_custom <- function(x, mu_init, sigma_init, pi_init,
                               max_iter = 30, tol = 1e-6) {
  n <- length(x)
  K <- length(mu_init)
  mu <- mu_init
  sigma <- sigma_init
  pi_k <- pi_init

  for (iter in seq_len(max_iter)) {
    resp <- matrix(0, nrow = n, ncol = K)
    for (k in seq_len(K)) {
      resp[, k] <- pi_k[k] * dnorm(x, mean = mu[k], sd = sigma[k])
    }
    resp <- resp / rowSums(resp)

    N_k <- colSums(resp)
    mu_new <- colSums(resp * x) / N_k
    sigma_new <- sqrt(colSums(resp * outer(x, mu_new, "-")^2) / N_k)
    pi_new <- N_k / n

    if (max(abs(mu_new - mu)) < tol) break
    mu <- mu_new
    sigma <- sigma_new
    pi_k <- pi_new
  }
  list(mu = mu_new, sigma = sigma_new, pi_k = pi_new)
}

# Bad initialisation: both means on the same side
fit_bad <- em_gaussian_custom(x,
  mu_init = c(5.5, 6.5),
  sigma_init = c(1, 1),
  pi_init = c(0.5, 0.5)
)

# Plot both results
x_grid <- seq(min(x) - 1, max(x) + 1, length.out = 500)

df_good <- data.frame(
  x = x_grid,
  density = fit_good$pi_k[1] * dnorm(x_grid, fit_good$mu[1], fit_good$sigma[1]) +
    fit_good$pi_k[2] * dnorm(x_grid, fit_good$mu[2], fit_good$sigma[2]),
  init = "Good initialisation"
)
df_bad <- data.frame(
  x = x_grid,
  density = fit_bad$pi_k[1] * dnorm(x_grid, fit_bad$mu[1], fit_bad$sigma[1]) +
    fit_bad$pi_k[2] * dnorm(x_grid, fit_bad$mu[2], fit_bad$sigma[2]),
  init = "Bad initialisation"
)

df_inits <- rbind(df_good, df_bad)

ggplot() +
  geom_histogram(data = data.frame(x = x), aes(x = x, y = after_stat(density)),
                 bins = 50, fill = "grey80", colour = "white") +
  geom_line(data = df_inits, aes(x = x, y = density, colour = init),
            linewidth = 1) +
  labs(x = "Value", y = "Density", colour = "",
       title = "The same algorithm, two different starting points") +
  theme_minimal()
```

```{r print_both_fits}
cat("Good initialisation:\n")
cat(sprintf("  Component 1: mean = %.2f, sd = %.2f\n", fit_good$mu[1], fit_good$sigma[1]))
cat(sprintf("  Component 2: mean = %.2f, sd = %.2f\n\n", fit_good$mu[2], fit_good$sigma[2]))

cat("Bad initialisation:\n")
cat(sprintf("  Component 1: mean = %.2f, sd = %.2f\n", fit_bad$mu[1], fit_bad$sigma[1]))
cat(sprintf("  Component 2: mean = %.2f, sd = %.2f\n", fit_bad$mu[2], fit_bad$sigma[2]))
```

In practice, the standard defence against local optima is to run EM multiple times with different random starting points and keep the result with the highest log-likelihood. Packages like `mclust` in R do this automatically.

# The General Pattern

EM always follows the same template, regardless of the specific model:

| Step | What you do | Two-coin example | Grading example | Gaussian mixture |
|------|-------------|------------------|-----------------|------------------|
| **Hidden data** | Identify what is unobserved | Which coin was used in each round | Which TA graded each assignment | Which component generated each data point |
| **Initialise** | Make an initial guess for the parameters | $\theta_A = 0.6$, $\theta_B = 0.5$ | Mean scores of 75 and 70 | Means, SDs, and weights |
| **E-step** | Compute the expected value of the hidden data, given current parameters | P(coin A \| round data) for each round | P(Alice \| score) for each assignment | P(component $k$ \| data point) for each point |
| **M-step** | Re-estimate parameters, treating the expected hidden data as if it were real | Weighted proportion of heads for each coin | Weighted mean score for each TA | Weighted mean, SD, and proportion for each component |
| **Repeat** | Until parameters stop changing | — | — | — |

The beauty of EM is that both the E-step and the M-step are usually **easy problems on their own**. The E-step is just plugging numbers into a formula. The M-step is often a familiar calculation (like computing a mean). It is only the combination — the fact that you don't know the hidden data — that makes the original problem hard. EM turns one hard problem into a sequence of easy ones.

# Common Pitfalls

## Choosing K

EM assumes you already know the number of components $K$. In the two-coin problem, we knew there were 2 coins. In practice, you may not know $K$. Common approaches:

- **BIC (Bayesian Information Criterion)**: fit models with different values of $K$ and pick the one with the best BIC, which balances fit against model complexity. This is what `mclust` does in R.
- **Domain knowledge**: sometimes the number of groups is known from the subject matter (e.g., two sexes, three treatment arms).

## Convergence Speed

EM can be slow when components overlap heavily. In such cases, the responsibilities stay uncertain for many iterations, and the parameters update by tiny amounts. This is usually not a problem in practice (modern computers handle thousands of iterations quickly), but it is worth knowing.

## Degenerate Solutions

If a Gaussian component collapses onto a single data point (mean equals that point, standard deviation shrinks to zero), the likelihood goes to infinity. This is a degenerate solution. Well-implemented packages guard against this by adding small regularisation terms or using constrained covariance structures.

# Summary

| Concept | Plain-language meaning |
|---|---|
| **EM algorithm** | An iterative method for estimating parameters when there is hidden (unobserved) information |
| **E-step** | "Given my current guesses, how should I fill in the blanks?" — compute soft assignments of hidden variables |
| **M-step** | "Given the filled-in blanks, what are the best parameter estimates?" — update parameters using standard formulas |
| **Responsibility** | The probability that a particular hidden state generated a particular data point |
| **Log-likelihood** | A score that measures how well the current parameters explain the data; EM never decreases it |
| **Local optimum** | EM finds a peak, but not necessarily the highest peak; run it multiple times with different starting points |
| **Convergence** | When the parameter updates become negligibly small, the algorithm stops |

The EM algorithm is not magic. It is a disciplined way of handling the chicken-and-egg problem that arises whenever you have incomplete data. Once you see it as "guess the hidden stuff, update your parameters, repeat", the notation in textbooks becomes much less intimidating.

# Session Info

```{r session_info}
sessionInfo()
```
