---
title: "Using fgsea with edgeR results"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

First install [fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html).

```{r install_fgsea, message=FALSE, warning=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("fgsea", quietly = TRUE))
  BiocManager::install("fgsea")

library(fgsea)
```

## edgeR results

[An example differential gene expression results table](https://davetang.org/muse/2024/10/31/an-example-differential-gene-expression-results-table/).

```{r edger_results}
edger_res <- readr::read_csv("https://raw.githubusercontent.com/davetang/muse/refs/heads/main/data/13970886_edger_res.csv", show_col_types = FALSE)
head(edger_res)
```

Add ranking metric.

```{r rank_metric}
edger_res |>
  dplyr::mutate(rank_metric = logFC * -log10(PValue)) -> edger_res
```

## Convert to Entrez Gene IDs

Use {org.Hs.eg.db}.

```{r install_org_hs_eg_db}
if (!require("org.Hs.eg.db", quietly = TRUE))
  BiocManager::install("org.Hs.eg.db")

suppressPackageStartupMessages(library(org.Hs.eg.db))
```

Convert to Entrez Gene IDs.

```{r ensembl_to_entrez}
AnnotationDbi::select(
  org.Hs.eg.db,
  keys = edger_res$ensembl_gene_id,
  columns=c("ENSEMBL","ENTREZID"),
  keytype="ENSEMBL"
) -> ensembl_to_entrez

ensembl_to_entrez <- dplyr::rename(ensembl_to_entrez, "ensembl_gene_id" = ENSEMBL)

head(ensembl_to_entrez)
```

Number of NAs.

```{r na_entrez_id}
table(is.na(ensembl_to_entrez$ENTREZID))
```

## Hallmark gene sets

Use {msigdb}.

```{r install_msigdb}
if (!require("msigdb", quietly = TRUE))
  BiocManager::install("msigdb")

if (!require("ExperimentHub", quietly = TRUE))
  BiocManager::install("ExperimentHub")

if (!require("GSEABase", quietly = TRUE))
  BiocManager::install("GSEABase")

suppressPackageStartupMessages(library(msigdb))
suppressPackageStartupMessages(library(ExperimentHub))
suppressPackageStartupMessages(library(GSEABase))
```

Query an `ExperimentHub` object.

```{r query_eh}
eh <- ExperimentHub(ask = FALSE)
AnnotationHub::query(x = eh, pattern = 'msigdb')
```

Latest version.

```{r msigdb_hs_latest}
AnnotationHub::query(x = eh, pattern = 'msigdb.*hs.EZID') |>
  tail(1) -> msigdb_hs_latest
msigdb_hs_latest
```

Download.

```{r download_msigdb}
msigdb_hs_ezid <- eh[[names(msigdb_hs_latest)]]
```

Collections.

```{r table_of_collections}
table(sapply(lapply(msigdb_hs_ezid, collectionType), bcCategory))
```

Create gene lists from the Hallmark collection.

```{r hallmark_gs_list}
wanted <- sapply(lapply(msigdb_hs_ezid, collectionType), bcCategory) == "h"
hallmark_gs <- msigdb_hs_ezid[wanted]

hallmark_gs_list <- lapply(hallmark_gs, geneIds)
class(hallmark_gs_list)
names(hallmark_gs_list) <- names(hallmark_gs)
head(hallmark_gs_list)
```

## fgsea analysis

Rank metric distribution.

```{r rank_metric_dist}
plot(hist(edger_res$rank_metric, breaks = 50))
```

Create ranks vector.

```{r my_ranks}
edger_res |>
  dplyr::inner_join(y = ensembl_to_entrez, by = "ensembl_gene_id", relationship = "many-to-many") |>
  dplyr::filter(!is.na(ENTREZID)) |>
  dplyr::group_by(ENTREZID) |>
  dplyr::mutate(ambiguous = ifelse(dplyr::n()>1, TRUE, FALSE)) |>
  dplyr::filter(!ambiguous) -> res
  
my_ranks <- res$rank_metric
my_names <- as.character(res$ENTREZID)
names(my_ranks) <- my_names
plot(hist(my_ranks, breaks = 50))
```

The `fgsea()` function runs the pre-ranked gene set enrichment analysis.

```{r fgsea}
set.seed(1984)

fgseaRes <- fgsea(
  pathways = hallmark_gs_list, 
  stats = my_ranks,
  minSize=15,
  maxSize=500,
  nPermSimple = 200000
)
```

Top 6 enriched pathways.

```{r fgsea_result}
head(fgseaRes[order(pval), ])
```

Plot the most significantly enriched pathway.

```{r plot_sign_pathway}
plotEnrichment(
  hallmark_gs_list[[head(fgseaRes[order(pval), ], 1)$pathway]],
  my_ranks
) +
  ggplot2::labs(title=head(fgseaRes[order(pval), ], 1)$pathway)
```
