---
title: "Fitting Generalised Linear Models"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

On [Wikipedia](https://en.wikipedia.org/wiki/Linear_regression), a Generalised Linear Model is described as follows:

> In statistics, a generalized linear model (GLM) is a flexible generalization of ordinary linear regression. The GLM generalizes linear regression by allowing the linear model to be related to the response variable via a link function and by allowing the magnitude of the variance of each measurement to be a function of its predicted value.

## Quick start

Predicting a the miles per gallon (`mpg`) of cars based on their weight (`wt`) using the `mtcars` dataset.

The `glm()` function can handle a variety of models by specifying a family (e.g., Gaussian, binomial, Poisson). For a basic linear regression, the `family = gaussian()` is used.

```{r glm_model}
glm_model <- glm(mpg ~ wt, data = mtcars, family = gaussian())
summary(glm_model)
```

Miles per gallon decreases -5.3445 for a unit gained in weight. In order words, the heavier the car, the less distance travelled per gallon of fuel.

The `lm()` function performs linear regression.

```{r lm_model}
lm_model <- lm(mpg ~ wt, data = mtcars)
summary(lm_model)
```

Both functions produce similar results for simple linear regression because `lm()` is essentially a special case of `glm()` with `family = gaussian()`.

`glm()` provides supports for various distributions and link functions but requires specifying the family (e.g., Gaussian, binomial).

`lm()` is designed specifically for linear regression and assumes the Gaussian distribution and identity link by default.

Use `glm()` when working with non-normal response variables or when needing other link functions. A link function is a mathematical function that connects the linear predictor of a generalised linear model (GLM) to the mean of the response variable's distribution. It allows the model to handle a wide range of response variable types (e.g., binary, count, continuous) by transforming the expected value of the response variable to a scale that matches the linear predictor.
