---
title: "Introduction to Gaussian Mixture Models"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r packages, message=FALSE}
if (!require("mclust", quietly = TRUE)) {
  install.packages("mclust")
}

if (!require("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}

library(mclust)
library(ggplot2)
```

# Introduction

Imagine you measure the heights of 1,000 people at a conference and plot a histogram. If the conference is all adults from the same population, you would probably see a single bell-shaped curve centred around the average height. But what if the conference, for whatever reason, includes both professional basketball players and professional jockeys? You would likely see **two** overlapping bell curves: one centred around a taller average and one around a shorter average.

A **Gaussian mixture model** (GMM) is a way to describe data that looks like it comes from multiple overlapping bell curves. "Gaussian" is another name for the normal distribution (the bell curve), and "mixture" means we are combining several of them. The model assumes that each data point was generated by one of several Gaussian distributions, but we do not know which one generated which point.

GMMs are useful whenever you suspect your data contains hidden subgroups. In biology, different cell types may express a gene at different levels. In marketing, customers may fall into distinct spending segments. In astronomy, stars may cluster into different populations. In all these cases, the observed data is a blend of several underlying groups, and a GMM can help you tease them apart.

# The Normal Distribution

The **normal** (Gaussian) distribution is the classic bell curve. It is defined by two parameters:

- **Mean** ($\mu$): the centre of the bell — where the peak is.
- **Standard deviation** ($\sigma$): how spread out the bell is. A small $\sigma$ gives a tall, narrow curve; a large $\sigma$ gives a short, wide curve.

The probability density function is:

$$f(x \mid \mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)$$

The key idea is that data drawn from a normal distribution clusters symmetrically around the mean, with most values falling within a couple of standard deviations.

```{r normal_curves, fig.width=8, fig.height=4, fig.cap="Three normal distributions with different means and standard deviations."}
set.seed(1984)
x <- seq(-6, 12, length.out = 500)

df_norm <- data.frame(
  x = rep(x, 3),
  density = c(
    dnorm(x, mean = 0, sd = 1),
    dnorm(x, mean = 4, sd = 1.5),
    dnorm(x, mean = 7, sd = 0.7)
  ),
  distribution = rep(
    c("mean=0, sd=1", "mean=4, sd=1.5", "mean=7, sd=0.7"),
    each = length(x)
  )
)

ggplot(df_norm, aes(x = x, y = density, colour = distribution)) +
  geom_line(linewidth = 1) +
  labs(x = "Value", y = "Density", colour = "Parameters",
       title = "Normal distributions with different parameters") +
  theme_minimal()
```

Each curve above is a single Gaussian. A GMM says: "My data is a weighted combination of several of these curves."

# Mixture

A **mixture distribution** combines multiple component distributions. For a Gaussian mixture with $K$ components, the overall density is:

$$p(x) = \sum_{k=1}^{K} \pi_k \; f(x \mid \mu_k, \sigma_k)$$

where:

- $K$ is the number of components (subgroups).
- $\pi_k$ is the **mixing weight** for component $k$; the proportion of data that belongs to that group. The weights must be positive and sum to 1: $\sum_{k=1}^{K} \pi_k = 1$.
- $f(x \mid \mu_k, \sigma_k)$ is the normal density for component $k$, with its own mean and standard deviation.

Think of it like a recipe. If 60% of conference attendees are basketball players (tall) and 40% are jockeys (shorter), the mixture is:

$$p(\text{height}) = 0.6 \times f(\text{height} \mid \mu_{\text{basketball}}, \sigma_{\text{basketball}}) + 0.4 \times f(\text{height} \mid \mu_{\text{jockey}}, \sigma_{\text{jockey}})$$

# Simulating a Mixture

The best way to understand a mixture is to build one. Let's simulate data from two Gaussian components and see what the combined distribution looks like.

```{r simulate_mixture}
set.seed(1984)

n <- 1000
# Component 1: 60% of the data, mean = 3, sd = 1
# Component 2: 40% of the data, mean = 7, sd = 1.2
n1 <- round(n * 0.6)
n2 <- n - n1

component1 <- rnorm(n1, mean = 3, sd = 1)
component2 <- rnorm(n2, mean = 7, sd = 1.2)

# The mixture: combine the two samples
mixture <- c(component1, component2)

# Keep track of the true labels (in practice, we don't have these)
true_labels <- c(rep(1, n1), rep(2, n2))
```

```{r plot_mixture, fig.width=9, fig.height=5, fig.cap="Histogram of mixture data coloured by true component membership."}
df_mix <- data.frame(
  value = mixture,
  component = factor(true_labels)
)

ggplot(df_mix, aes(x = value, fill = component)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 alpha = 0.6, position = "identity", colour = "white") +
  geom_density(aes(colour = component), linewidth = 0.8, fill = NA) +
  labs(x = "Value", y = "Density",
       title = "Simulated data from a two-component Gaussian mixture",
       fill = "True Component", colour = "True Component") +
  theme_minimal()
```

You can see two overlapping humps. In practice, we would observe only the combined histogram (without the colours) and would need a GMM to recover the two underlying groups.

```{r plot_combined, fig.width=9, fig.height=5, fig.cap="The same data viewed without component labels — this is what we actually observe."}
ggplot(df_mix, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "steelblue", colour = "white", alpha = 0.7) +
  geom_density(linewidth = 0.8, colour = "red") +
  labs(x = "Value", y = "Density",
       title = "Observed data (true component labels hidden)") +
  theme_minimal()
```

This is the problem a GMM solves: given only the combined histogram, recover the number of components, their means, standard deviations, and mixing weights.

# The EM Algorithm

Fitting a GMM means estimating all the parameters ($\pi_k$, $\mu_k$, $\sigma_k$ for each component). The standard approach is the **Expectation-Maximisation (EM) algorithm**. The intuition behind EM is surprisingly simple: it is an iterative guessing game.

## The Chicken-and-Egg Problem

If we **knew** which component each data point belonged to, estimating the parameters would be straightforward; just calculate the mean and standard deviation for each group separately. But we don't know the group assignments.

Conversely, if we **knew** the parameters of each component, figuring out which group each point probably belongs to would be easy; just check which bell curve is taller at that point. But we don't know the parameters either.

EM breaks this deadlock by alternating between two steps:

1. **E-step (Expectation)**: Given the current guess at the parameters, calculate the **responsibility** of each component for each data point. The responsibility is a number between 0 and 1 that answers: "How likely is it that this point came from component $k$?" A point sitting right under the peak of component 1 will have a high responsibility for component 1 and a low responsibility for component 2.

2. **M-step (Maximisation)**: Given the current responsibilities, update the parameters. The new mean of component $k$ is the weighted average of all data points, where the weights are the responsibilities. Similarly for the standard deviation and mixing weight.

These two steps repeat until the estimates stop changing (converge). Each iteration is guaranteed to improve the fit (or at least not make it worse), so the algorithm homes in on a good solution.

## A Visual Walkthrough

Let's watch EM in action. We'll implement a simple version for our simulated data.

```{r em_manual}
em_gmm <- function(x, K = 2, max_iter = 50, tol = 1e-6) {
  n <- length(x)

  # Initialise parameters using random quantiles
  set.seed(1984)
  mu <- quantile(x, probs = seq(0.25, 0.75, length.out = K))
  sigma <- rep(sd(x) / K, K)
  pi_k <- rep(1 / K, K)

  # Store history for plotting
  history <- list()

  for (iter in seq_len(max_iter)) {
    # E-step: compute responsibilities
    resp <- matrix(0, nrow = n, ncol = K)
    for (k in seq_len(K)) {
      resp[, k] <- pi_k[k] * dnorm(x, mean = mu[k], sd = sigma[k])
    }
    resp <- resp / rowSums(resp)

    # M-step: update parameters
    N_k <- colSums(resp)
    mu_new <- colSums(resp * x) / N_k
    sigma_new <- sqrt(colSums(resp * outer(x, mu_new, "-")^2) / N_k)
    pi_new <- N_k / n

    history[[iter]] <- data.frame(
      iteration = iter,
      component = seq_len(K),
      mu = mu_new,
      sigma = sigma_new,
      weight = pi_new
    )

    # Check convergence
    if (max(abs(mu_new - mu)) < tol) break

    mu <- mu_new
    sigma <- sigma_new
    pi_k <- pi_new
  }

  list(
    mu = mu_new,
    sigma = sigma_new,
    weights = pi_new,
    responsibilities = resp,
    history = do.call(rbind, history),
    iterations = iter
  )
}

fit <- em_gmm(mixture, K = 2)

cat("EM converged in", fit$iterations, "iterations\n\n")
cat("Estimated parameters:\n")
for (k in 1:2) {
  cat(sprintf("  Component %d: mean = %.2f, sd = %.2f, weight = %.2f\n",
              k, fit$mu[k], fit$sigma[k], fit$weights[k]))
}
cat("\nTrue parameters:\n")
cat("  Component 1: mean = 3.00, sd = 1.00, weight = 0.60\n")
cat("  Component 2: mean = 7.00, sd = 1.20, weight = 0.40\n")
```

The estimated values should be close to the true parameters we used to simulate the data.

```{r em_convergence, fig.width=9, fig.height=4, fig.cap="Convergence of mean estimates across EM iterations."}
ggplot(fit$history, aes(x = iteration, y = mu,
                        colour = factor(component))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = c(3, 7), linetype = "dashed", colour = "grey40") +
  annotate("text", x = max(fit$history$iteration), y = 3.2,
           label = "True mean = 3", hjust = 1, colour = "grey40") +
  annotate("text", x = max(fit$history$iteration), y = 7.2,
           label = "True mean = 7", hjust = 1, colour = "grey40") +
  labs(x = "Iteration", y = "Estimated Mean",
       colour = "Component",
       title = "EM algorithm convergence") +
  theme_minimal()
```

# Visualising the Fitted Model

Now let's overlay the estimated Gaussian components on the data.

```{r fitted_overlay, fig.width=9, fig.height=5, fig.cap="Observed data with fitted Gaussian components overlaid."}
x_grid <- seq(min(mixture) - 1, max(mixture) + 1, length.out = 500)

df_fitted <- data.frame(
  x = rep(x_grid, 3),
  density = c(
    fit$weights[1] * dnorm(x_grid, fit$mu[1], fit$sigma[1]),
    fit$weights[2] * dnorm(x_grid, fit$mu[2], fit$sigma[2]),
    fit$weights[1] * dnorm(x_grid, fit$mu[1], fit$sigma[1]) +
      fit$weights[2] * dnorm(x_grid, fit$mu[2], fit$sigma[2])
  ),
  curve = rep(c("Component 1", "Component 2", "Mixture"),
              each = length(x_grid))
)

ggplot() +
  geom_histogram(data = df_mix, aes(x = value, y = after_stat(density)),
                 bins = 50, fill = "steelblue", colour = "white", alpha = 0.5) +
  geom_line(data = df_fitted, aes(x = x, y = density, colour = curve),
            linewidth = 1) +
  scale_colour_manual(values = c("Component 1" = "#E69F00",
                                  "Component 2" = "#56B4E9",
                                  "Mixture" = "red")) +
  labs(x = "Value", y = "Density", colour = "",
       title = "Fitted Gaussian mixture model") +
  theme_minimal()
```

The red curve (overall mixture) follows the shape of the histogram, while the orange and blue curves show the two estimated components. Each component captures one of the two humps in the data.

# Soft vs Hard Clustering

A distinctive feature of GMMs is **soft** (probabilistic) clustering. Instead of assigning each point to exactly one group, a GMM gives each point a probability of belonging to each component. Points in the overlap region between two Gaussians will have intermediate probabilities.

```{r soft_clustering, fig.width=9, fig.height=5, fig.cap="Data points coloured by their responsibility (probability of belonging to component 2)."}
df_resp <- data.frame(
  value = mixture,
  responsibility = fit$responsibilities[, 2]
)

ggplot(df_resp, aes(x = value, y = 0, colour = responsibility)) +
  geom_jitter(height = 0.4, size = 0.8, alpha = 0.7) +
  scale_colour_gradient(low = "#E69F00", high = "#56B4E9",
                        name = "P(Component 2)") +
  labs(x = "Value", y = "",
       title = "Soft assignments: each point has a probability for each component") +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

Points clearly on the left (low values) are coloured orange — the model is confident they belong to component 1. Points on the right are blue — confidently component 2. Points in the middle have intermediate colours, reflecting genuine uncertainty.

If you need **hard** assignments (each point in exactly one group), you simply assign each point to the component with the highest responsibility:

```{r hard_clustering}
hard_labels <- apply(fit$responsibilities, 1, which.max)
agreement <- mean(hard_labels == true_labels)
cat(sprintf("Agreement with true labels: %.1f%%\n", agreement * 100))
```

# Using mclust

In practice, you would not write the EM algorithm yourself. The `mclust` package in R provides a robust, well-tested implementation that also handles:

- **Choosing the number of components** automatically using the Bayesian Information Criterion (BIC).
- **Multivariate data** (multiple dimensions).
- **Different covariance structures** (components can have different shapes and orientations).

```{r mclust_fit}
mc <- Mclust(mixture)
summary(mc)
```

`Mclust` tests a range of component numbers (by default 1 to 9) and selects the one with the best BIC. Let's see the BIC values.

```{r bic_plot, fig.width=8, fig.height=5, fig.cap="BIC values for different numbers of components. Higher BIC is better in mclust's convention."}
plot(mc, what = "BIC")
```

The model with 2 components should have the highest BIC, matching the truth.

## What Is BIC?

The **Bayesian Information Criterion** (BIC) balances two competing goals:

1. **Fit**: A model with more components can always fit the data more closely (just as a higher-degree polynomial can wiggle through more points).
2. **Simplicity**: More components means more parameters, which risks **overfitting** — fitting noise rather than real structure.

BIC penalises model complexity:

$$\text{BIC} = 2 \ln(\hat{L}) - p \ln(n)$$

where $\hat{L}$ is the maximised likelihood, $p$ is the number of parameters, and $n$ is the sample size. The penalty $p \ln(n)$ grows with the number of parameters, discouraging unnecessarily complex models. We pick the model with the **highest** BIC (in `mclust`'s sign convention).

## Extracting Results

```{r mclust_results}
cat("Selected number of components:", mc$G, "\n\n")

cat("Estimated means:\n")
print(round(mc$parameters$mean, 2))

cat("\nEstimated mixing weights:\n")
print(round(mc$parameters$pro, 2))

cat("\nEstimated variances:\n")
print(round(mc$parameters$variance$sigmasq, 2))
```

```{r mclust_classification, fig.width=9, fig.height=5, fig.cap="mclust classification of the mixture data."}
df_mclust <- data.frame(
  value = mixture,
  classified = factor(mc$classification),
  uncertainty = mc$uncertainty
)

ggplot(df_mclust, aes(x = value, fill = classified)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 alpha = 0.6, position = "identity", colour = "white") +
  labs(x = "Value", y = "Density", fill = "Assigned Component",
       title = "mclust classification of mixture data") +
  theme_minimal()
```

`mclust` also provides an **uncertainty** measure for each point. Points in the overlap zone will have higher uncertainty.

```{r uncertainty_plot, fig.width=9, fig.height=4, fig.cap="Classification uncertainty is highest where the two components overlap."}
ggplot(df_mclust, aes(x = value, y = uncertainty, colour = classified)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = "Value", y = "Uncertainty", colour = "Component",
       title = "Classification uncertainty") +
  theme_minimal()
```

# A More Realistic Example: Old Faithful

The `faithful` dataset in R records eruption durations and waiting times for the Old Faithful geyser in Yellowstone National Park. It is a classic example of bimodal data: there are short eruptions with short waiting times and long eruptions with long waiting times.

```{r faithful_scatter, fig.width=7, fig.height=6, fig.cap="Old Faithful eruption data showing two clusters."}
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_point(alpha = 0.6) +
  labs(x = "Eruption Duration (minutes)", y = "Waiting Time (minutes)",
       title = "Old Faithful Geyser Data") +
  theme_minimal()
```

Let's fit a GMM to this two-dimensional data.

```{r faithful_gmm}
mc_faithful <- Mclust(faithful)
summary(mc_faithful)
```

```{r faithful_classification, fig.width=7, fig.height=6, fig.cap="Old Faithful data classified by a Gaussian mixture model."}
df_faithful <- data.frame(
  eruptions = faithful$eruptions,
  waiting = faithful$waiting,
  component = factor(mc_faithful$classification),
  uncertainty = mc_faithful$uncertainty
)

ggplot(df_faithful, aes(x = eruptions, y = waiting,
                        colour = component, size = uncertainty)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(1, 4), name = "Uncertainty") +
  labs(x = "Eruption Duration (minutes)", y = "Waiting Time (minutes)",
       colour = "Component",
       title = "GMM classification of Old Faithful data") +
  theme_minimal()
```

The GMM separates the two eruption types cleanly. Points at the boundary between groups are displayed larger because they have higher classification uncertainty.

# When to Use GMMs

GMMs are a good choice when:

- You believe your data contains **subgroups** that each follow an approximately normal distribution.
- You want **soft** (probabilistic) cluster assignments rather than hard boundaries.
- You want a **principled way to choose the number of clusters** (via BIC), rather than relying on visual inspection.
- You need a **generative model** — once fitted, a GMM can simulate new data that resembles the original.

GMMs may not be ideal when:

- Your subgroups are not roughly Gaussian-shaped (e.g., crescent or ring-shaped clusters). In such cases, consider kernel density methods or non-parametric approaches.
- You have very high-dimensional data without dimensionality reduction. The number of parameters in a GMM grows quickly with dimensionality.
- The components overlap so heavily that they are effectively indistinguishable.

# Summary

| Concept | What it means |
|---|---|
| **Gaussian** | The normal (bell curve) distribution, defined by a mean and standard deviation |
| **Mixture** | A weighted combination of several distributions |
| **Component** | One of the Gaussian distributions in the mixture |
| **Mixing weight** ($\pi_k$) | The proportion of data belonging to component $k$ |
| **EM algorithm** | An iterative procedure that alternates between estimating group memberships (E-step) and updating parameters (M-step) |
| **Responsibility** | The probability that a data point belongs to a particular component |
| **BIC** | A criterion for choosing the number of components, balancing fit and simplicity |

The core idea is straightforward: if your data looks like it was generated by several overlapping bell curves, a GMM can recover those curves and tell you which data points likely belong to which group. The EM algorithm does the heavy lifting, and tools like `mclust` make it accessible in just a few lines of R code.

# Session Info

```{r session_info}
sessionInfo()
```
