---
title: "Solving problems with graphs"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

## Perfect squares

I saw this question on [Quora](https://www.quora.com/Is-there-any-quick-way-to-solve-this-math-problem):

>A teacher assigns each of her 18 students a different integer from 1 through
>18. The teacher forms pairs of study partners by using the rule that the sum
>of the pair of numbers is a perfect square. Assuming the 9 pairs of students
>follow this rule, the student assigned which number must be paired with the
>student assigned the number 1?
>
>A. 16
>B. 15
>C. 9
>D. 8
>E. 3

Firstly, a perfect square is a number made by squaring a whole number. The first ten perfect squares are listed below.

```{r perfect_squares}
(1:10)^2
```

We can already narrow the answer to either A (16) or C (9), which means that the student assigned 1 can either pair with the student assigned 8 or 15. But the student assigned 8 can also pair with 17 and the student assigned 15 can also pair with 10. A graph is a nice way to visualise the possibilities to narrow down on a solution.

We can use the `combn()` function to generate all the pair combinations.

```{r student_pairs}
student_pairs <- t(combn(1:18, 2))
tail(student_pairs)
```

However, not all pairs form perfect squares so we need to eliminate those pairings. Given the definition of a perfect square, we can write a function to check if a number is a perfect square. The tricky bit is writing a test in R to check whether a number is a whole number because the `sqrt()` function does not return an integer even when a number is a perfect square.

```{r sqrt_eg}
class(sqrt(25))
class(sqrt(26))
```

We can use modulus one because a whole number will equally divide by one.

```{r modulus}
25 %% 1
25.5 %% 1
```

Now we can write a function to check whether a number is a perfect square.

```{r check_perfect_square}
check_ps <- function(n){
  stopifnot(n > 0)
  sqrt(n) %% 1 == 0
}

check_ps(100)
```

Remove the pairs that are not perfect squares.

```{r filter_student_pairs}
wanted <- check_ps(apply(student_pairs, 1, sum))
student_pairs <- student_pairs[wanted, ]
student_pairs
```

Let's check whether all students are present, to see if they were included in at least one pair. The `all()` function checks whether all values are true.

```{r check_pairs}
all(1:18 %in% student_pairs)
```

Finally, we can visualise the pairs as a graph.

```{r graph_pairs}
suppressPackageStartupMessages(library(igraph))
net <- graph.data.frame(student_pairs, directed = FALSE)
plot(net, layout = layout_components(net))
```

Since 16, 17, and 18 must pair with 9, 8, and 7, respectively, 2 must pair with 14, 11 with 5, 4 with 12, 13 with 3, 6 with 10. Therefore 1 has to pair with 15.

I'm curious whether this type of pairing is possible with any class size that is even.

```{r plot_pairs}
gen_pairs <- function(n){
  spairs <- t(combn(1:n, 2))
  wanted <- check_ps(apply(spairs, 1, sum))
  spairs[wanted, ]
}

plot_pairs <- function(spairs){
  net <- graph.data.frame(spairs, directed = FALSE)
  plot(net, layout = layout_components(net))
}

plot_pairs(gen_pairs(20))
```

To check whether a solution exists, we:

1. Check whether all students are in at least one potential pair.
2. Find pairs where one number is unique, i.e. unique pair, (and save them) and then remove other pairs that include either number of the unique pair.
3. Keep repeating step 2 until no pairs are left or if no unique pair exists, we will have to use a backtracking algorithm to pick a random pair (not implemented).

Let's see if this works for our example with a class of 18.

```{r eg1}
eg1 <- gen_pairs(18)

get_row_idx <- function(mat, vec){
  unique(which(t(apply(mat, 1, function(x) x %in% vec)), arr.ind = TRUE)[, 1])
}

get_unique_pair_idx <- function(mat){
  vec <- as.vector(mat)
  dup <- unique(vec[duplicated(vec)])
  wanted <- setdiff(unique(vec), dup)
  get_row_idx(mat, wanted)
}

find_solution <- function(mat, iter = 100){
  solution <- matrix(nrow = 0, ncol = 2)
  i <- 1
  while(nrow(mat) > 0){
    uniq_row <- get_unique_pair_idx(mat)
    if(length(uniq_row) == 0){
      message("No unique row")
      plot_pairs(mat)
      return(mat)
    }
    up <- mat[uniq_row, ]
    solution <- rbind(solution, up, deparse.level = 0)
    vec <- as.vector(up)
    wanted <- get_row_idx(mat, vec)
    mat <- mat[-wanted,, drop = FALSE]
    if(length(unique(as.vector(mat))) %% 2 != 0){
      message("Odd number remaining")
      plot_pairs(mat)
      return(mat)
    }
    i <- i + 1
    if(i == iter){
      print(solution)
      stop("Too many iterations")
    }
  }
  return(solution)
}

find_solution(eg1)
```

The code works nicely for a class size of 18 and 24 but doesn't handle a class size of 20 properly because there is no solution.

```{r eg2}
eg2 <- gen_pairs(20)
plot_pairs(eg2)
```

When 18 is paired with 7, one number will be left without a pair.

The code also doesn't work when we have to make a choice because each number has two or more possible pairs. Backtracking needs to be implemented at this point to find a possible solution.

```{r eg3}
eg3 <- gen_pairs(30)
find_solution(eg3)
```
