---
title: "Working with HDF5 data in R"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(hdf5r)
```

# Introduction

From [Wikipedia](https://en.wikipedia.org/wiki/Hierarchical_Data_Format):

> Hierarchical Data Format (HDF) is a set of file formats (HDF4, HDF5) designed
to store and organize large amounts of data. Originally developed at the U.S.
National Center for Supercomputing Applications, it is supported by The HDF
Group, a non-profit corporation whose mission is to ensure continued development
of HDF5 technologies and the continued accessibility of data stored in HDF.
>
> In keeping with this goal, the HDF libraries and associated tools are
available under a liberal, BSD-like license for general use. HDF is supported by
many commercial and non-commercial software platforms and programming languages.
The freely available HDF distribution consists of the library, command-line
utilities, test suite source, Java interface, and the Java-based HDF Viewer
(HDFView).
>
> The current version, HDF5, differs significantly in design and API from the
major legacy version HDF4.

# h5ad

[Download](https://cellxgene.cziscience.com/collections/e5f58829-1a66-40b5-a624-9046778e74f5) from Tabula Sapiens:

> Tabula Sapiens is a benchmark, first-draft human cell atlas of over 1.1M cells from 28 organs of 24 normal human subjects. This work is the product of the Tabula Sapiens Consortium. Taking the organs from the same individual controls for genetic background, age, environment, and epigenetic effects, and allows detailed analysis and comparison of cell types that are shared between tissues.

```{r ts_heart}
ts_heart_url <- "https://datasets.cellxgene.cziscience.com/97516b79-8d08-46a6-b329-5d0a25b0be98.h5ad"
ts_heart_file <- paste0('data/', basename(ts_heart_url))
if(file.exists(ts_heart_file)){
  message(paste0(ts_heart_file, " already exists"))
} else {
  options(timeout = 1000)
  download.file(url = ts_heart_url, destfile = ts_heart_file)
}
```

Creat new `H5File`.

```{r ts_heart_h5}
ts_heart_h5 <- H5File$new(filename = ts_heart_file, mode = "r")
ts_heart_h5
```

List datasets.

```{r list_datasets_heart_h5}
hdf5r::list.datasets(ts_heart_h5) |>
  head()
```

UMAP.

```{r heart_h5_umap}
ts_heart_umap_h5 <- ts_heart_h5[["obsm/X_umap"]]
ts_heart_umap <- ts_heart_umap_h5[,]
plot(t(ts_heart_umap), pch = '.', xlab = "UMAP1", ylab = "UMAP2", main = "Tabula Sapiens Heart")
```

# 10x HDF5

## Gene expression data

Download [1.3 Million Brain Cells from E18 Mice](https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.3.0/1M_neurons?) (3.93 GB).

```{r download_1m_neurons}
my_url <- 'https://cf.10xgenomics.com/samples/cell-exp/1.3.0/1M_neurons/1M_neurons_filtered_gene_bc_matrices_h5.h5'
my_file <- paste0("data/", basename(my_url))

if(!file.exists(my_file)){
  options(timeout = 10000)
  download.file(url = my_url, destfile = my_file)
}
```

New `H5File` object.

```{r gex_feat_mat}
gex_feat_mat <- H5File$new(filename = my_file, mode = "r")
gex_feat_mat
```

List datasets.

```{r list_datasets}
hdf5r::list.datasets(gex_feat_mat)
```

Get gene names.

```{r gene_names}
gene_names_h5 <- gex_feat_mat[["mm10/gene_names"]]
gene_names <- gene_names_h5[]
class(gene_names)
length(gene_names)
head(gene_names)
```

# ARCHS4

[Download ARCHS4 HDF5 data](https://maayanlab.cloud/archs4/download.html) and
load using the [hdf5r](https://github.com/hhoeflin/hdf5r) package.

```{r human}
human <- H5File$new(filename = "/data2/archs4/human_gene_v2.5.h5", mode = "r")
human
```

Different groups.

```{r groups}
human[["data"]]
human[["meta"]]
```

## Getting information about objects

Use the `names()` function to get all names of objects.

```{r names}
names(human)
```

The `ls()` function provides more information.

```{r ls}
human$ls()
```

List functions.

```{r list}
hdf5r::list.attributes(human)
hdf5r::list.datasets(human)
hdf5r::list.groups(human)
hdf5r::list.objects(human)
```

## Get data

Function to get data.

```{r get_data}
wrap <- function(x){
  x <- sub("^", "[['", x)
  sub("$", "']]", x)
}

get_data <- function(obj, ds, ...){
  s <- unlist(strsplit(x = ds, split = "/"))
  txt <- paste0(obj, paste(sapply(s, wrap), collapse = "", sep = ""), ...)
  eval(parse(text = txt))
}
```

Check out some of the data.

```{r check_out_data}
get_data('human', 'meta/samples/title', '[1:5]')
get_data('human', 'meta/samples/sample', '[1:5]')
get_data('human', 'meta/samples/geo_accession', '[1:5]')
get_data('human', 'meta/samples/series_id', '[1:5]')
get_data('human', 'meta/samples/characteristics_ch1', '[1:5]')
get_data('human', 'meta/samples/extract_protocol_ch1', '[1]')
get_data('human', 'meta/samples/source_name_ch1', '[1:5]')
```

## Store gene symbols

Gene symbols.

```{r meta_genes_symbol}
human[['meta']][['genes']][['symbol']]
```

Store gene symbols.

```{r my_gene_sym}
my_idx <- 1:human[['meta']][['genes']][['symbol']]$maxdims
my_gene_sym <- human[['meta']][['genes']][['symbol']][my_idx]
head(my_gene_sym)
```

## Source

Check out some sources.

```{r example_source}
unique(get_data('human', 'meta/samples/source_name_ch1', '[1:100]'))
```

Store all sources.

```{r store_source}
my_idx <- 1:human[['meta']][['samples']][['source_name_ch1']]$maxdims
my_source <- human[['meta']][['samples']][['source_name_ch1']][my_idx]
head(my_source)
```

Search.

```{r length_source}
length(my_source[grepl("immune", my_source, ignore.case = TRUE)])
length(my_source[grepl("dendritic", my_source, ignore.case = TRUE)])
```

## Expression

Genes are columns and rows are samples.

```{r data_expression}
exp_data <- human[['data']][['expression']]
exp_data
```

Sum across first ten samples.

```{r sum_ten_samples}
apply(exp_data[1:10, ], 1, sum)
```

TP53 position.

```{r tp53}
tp53_idx <- match("TP53", my_gene_sym)
tp53_idx
```

## Clean up

Close.

```{r close_all}
human$close_all()
```

## Related

* [Official ARCHS4 companion package](https://github.com/MaayanLab/archs4py).
* [ARCHS4 help page](https://maayanlab.cloud/archs4/help.html)
