---
title: "Information"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(mpmi))
knitr::opts_chunk$set(echo = TRUE)
```

Install the dependencies required to render this document.

```{r install_dep, eval=FALSE}
install.packages('mpmi')
packageVersion("mpmi")
```

Example data.

```{r eg1}
set.seed(1984)

n <- 300
Org <- sample(c("Org1","Org2","Org3"), n, replace=TRUE, prob=c(0.3, 0.4, 0.3))

Cluster <- character(n)
for (i in 1:n) {
  if (Org[i] == "Org1") {
    Cluster[i] <- sample(c("C1","C2","C3","C4"), 1, prob=c(0.6,0.2,0.1,0.1))
  } else if (Org[i] == "Org2") {
    Cluster[i] <- sample(c("C1","C2","C3","C4"), 1, prob=c(0.2,0.5,0.2,0.1))
  } else {
    Cluster[i] <- sample(c("C1","C2","C3","C4"), 1, prob=c(0.1,0.1,0.3,0.5))
  }
}

df <- data.frame(Cluster, Org)
table(df$Cluster, df$Org)
```

This function calculates MI and BCMI between a set of discrete variables held as columns in a matrix. It also performs jackknife bias correction and provides a z-score for the hypothesis of no association. Also included are the *.pw functions that calculate MI between two vectors only. The *njk functions do not perform the jackknife and are therefore faster. 

```{r dmi}
dmi(df)
```
