---
title: "JSON versus YAML"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

JSON and YAML are popular [serialisation](https://en.wikipedia.org/wiki/Serialization) formats.

>In computing, serialization (or serialisation) is the process of translating a data structure or object state into a format that can be stored (e.g. files in secondary storage devices, data buffers in primary storage devices) or transmitted (e.g. data streams over computer networks) and reconstructed later (possibly in a different computer environment).

Install the [jsonlite](https://github.com/jeroen/jsonlite) and [yaml](https://github.com/vubiostat/r-yaml) packages, so that we can generate JSON and YAML.

```{r install}
install.packages(c("jsonlite", "yaml"))
```

Load libraries.

```{r load}
library(jsonlite)
library(yaml)
```

As a first example, we will convert the `women` data set, which is a small data set with 15 observations for 2 variables.

```{r women}
women
```

Convert `women` to JSON.

```{r women_json}
women_json <- toJSON(women, pretty = TRUE)
women_json
```

Convert `women` to YAML.

```{r women_yaml}
women_yaml <- as.yaml(women, indent = 3)
writeLines(women_yaml)
```

## To data frame

JSON to data frame.

```{r json_to_df}
fromJSON(women_json)
```

YAML to data frame. This does not work for more complex data structures (see below).

```{r yaml_to_df}
yaml.load(women_yaml, handlers = list(map = function(x) as.data.frame(x) ))
```

## Non-tidy data frame

A data frame containing lists.

```{r my_df}
my_df <- data.frame(
  id = 1:3,
  title = letters[1:3]
)
my_df$keywords = list(
    c('aa', 'aaa', 'aaaa'),
    c('bb', 'bbb'),
    c('cc', 'ccc', 'cccc', 'ccccc')
)

my_df
```

Convert `my_df` to JSON.

```{r my_df_json}
my_df_json <- toJSON(my_df, pretty = TRUE)
my_df_json
```

Convert `my_df` to YAML.

```{r my_df_yaml}
my_df_yaml <- as.yaml(my_df, indent = 3)
writeLines(my_df_yaml)
```

## JSON to YAML and vice versa

Converting from JSON to YAML is easy.

```{r json_to_yaml}
identical(writeLines(as.yaml(fromJSON(my_df_json))), writeLines(my_df_yaml))
```

Converting from YAML to JSON for `my_df` is not as straight-forward because of the different number of keywords.

```{r yaml_load}
my_df_list <- yaml.load(my_df_yaml)
my_df_list
```

This conversion is different from the original data frame to JSON conversion because this creates a single object, where as the original conversion creates an array with three objects.

```{r yaml_to_json_wrong}
toJSON(my_df_list, pretty = TRUE)
my_df_json
```

I could probably write a hacky function to make the conversion but I won't.
