---
title: "Matrix Market"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

[Matrix Market exchange formats](https://en.wikipedia.org/wiki/Matrix_Market_exchange_formats):

> The Matrix Market exchange formats are a set of human readable, ASCII-based file formats designed to facilitate the exchange of matrix data. The file formats were designed and adopted for the Matrix Market, a NIST repository for test data for use in comparative studies of algorithms for numerical linear algebra.

Download Matrix Market file.

```{r download_matrix_mtx}
my_url <- 'https://davetang.org/file/pbmc3k/filtered_gene_bc_matrices/hg19/matrix.mtx'

mtx_file <- "data/pbmc3k_matrix.mtx"

if(!file.exists(mtx_file)){
  download.file(url = my_url, destfile = mtx_file)
}
```

The {Matrix} package has support for Matrix Market using `readMM()` and `writeMM()`.

```{r my_mat}
my_mat <- Matrix::readMM(mtx_file)
class(my_mat)
```

Convert dgTMatrix to dgCMatrix.

| Feature                 | `dgCMatrix`                                 | `dgTMatrix`                                    |
| ----------------------- | ------------------------------------------- | ---------------------------------------------- |
| **Name**                | *Double General Compressed Column*          | *Double General Triplet*                       |
| **Format**              | CSC (Compressed Sparse Column)              | COO (Coordinate / Triplet)                     |
| **Storage**             | Uses column pointers + row indices          | Stores each non-zero entry's (row, col, val)   |
| **Construction**        | Less intuitive (needs compressed structure) | Easy to build from scratch                     |
| **Efficiency**          | Fast and memory-efficient                   | Slower and less compact                        |

```{r dgc_matrix}
my_mat <- as(my_mat, "dgCMatrix")
class(my_mat)
```

Export.

```{r write_mm}
Matrix::writeMM(obj = my_mat, file = "data/pbmc3k_export.mtx")
```
