---
title: "Matthews Correlation Coefficient"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(yardstick)
  library(tibble)
})
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The Matthews Correlation Coefficient (MCC) is a single-number measure of how well a **binary classification** model performs. It takes into account **true and false positives and negatives**, providing a **balanced** evaluation even when classes are **unevenly distributed**.

Imagine an email classifier that predicts whether an email is **spam** (`1`) or **not spam** (`0`).

|                     | Predicted Positive  | Predicted Negative  |
| ------------------- | ------------------- | ------------------- |
| **Actual Positive** | True Positive (TP)  | False Negative (FN) |
| **Actual Negative** | False Positive (FP) | True Negative (TN)  |

The MCC is calculated following:

$$
\text{MCC} = \frac{(TP \times TN) - (FP \times FN)}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}}
$$

* +1: perfect prediction
* 0: random prediction
* −1: total disagreement (inverse prediction)

MCC works well **even with class imbalance** and provides a **balanced score**.

## Simple example

Given the following predictions and actual outcomes:

```{r eg1}
actual <- c(1, 0, 1, 1, 0, 0, 1, 0)
predicted <- c(1, 0, 1, 0, 0, 0, 1, 1)
table(actual, predicted)
```

* TP = 3
* TN = 3
* FP = 1
* FN = 1

Calculate the MCC:

```{r eg1_mcc}
TP <- 3
TN <- 3
FP <- 1
FN <- 1

my_mcc <- (TP * TN - FP * FN) /
  sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

my_mcc
```

## Multiclass

Suppose we have a model predicting three possible classes: "A", "B", and "C".

```{r eg2}
# Ground truth (actual labels)
truth <- factor(c("A", "B", "C", "A", "B", "C", "A", "C", "B", "B"))

# Model predictions
predicted <- factor(c("A", "C", "C", "A", "B", "B", "C", "C", "B", "A"))

results <- tibble(
  truth = truth,
  estimate = predicted
)

results
```

Using {yardstick}.

```{r eg2_mcc}
yardstick::mcc_vec(truth, predicted)
```
