---
title: "Mixed-effects models"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
})
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Mixed-effects models (also known as multilevel models, hierarchical models, or random-effects models) are extensions of linear regression that account for both fixed and random effects.

Given students’ test scores across multiple schools:

* Fixed effects are predictors that apply to everyone such as the effect of study hours.
* Random effects are effects that vary across groups such as school-to-school differences.

Each school might have a slightly different average score (intercept) or respond differently to study hours (slope).

Mixed-effects models handle this by modeling:

$$
\text{Score}*{ij} = \beta_0 + \beta_1 \text{Hours}*{ij} + u_{0j} + \epsilon_{ij}
$$

where:

* $( j )$: the group index, i.e., the school.
* $( i )$: the individual (observation) index within each group, i.e., the student within school $( j )$.

Each school $( j )$ has its own intercept $( u_{0j} )$ that represents how that school's mean score differs from the overall population mean $( \beta_0 )$.

Each student $( i )$ within that school has their own hours studied ($( \text{Hours}*{ij} )$) and residual error ($( \epsilon_{ij} )$).

## Random effects

A random effect is a variable whose levels are drawn from a larger population, not specifically fixed.

For instance:

* If we measured 10 schools, they are a sample from all possible schools.
* We assume their effects (intercepts/slopes) come from a distribution (often Normal).

**Random effects account for correlation among observations within the same group and share information across groups**.

## Example in R

Install dependencies.

```{r install_libraries, eval=FALSE}
install.packages("lme4")
```

Example data: Reaction times in a sleep deprivation study.

> These data are from the study described in Belenky et al. (2003), for the most sleep-deprived group (3 hours time-in-bed) and for the first 10 days of the study, up to the recovery period. The original study analyzed speed (1/(reaction time)) and treated day as a categorical rather than a continuous predictor.

```{r sleepstudy}
library(lme4)
data("sleepstudy")
head(sleepstudy)
```

Visualise the data.

```{r sleepstudy_by_subject}
ggplot(sleepstudy, aes(x = Days, y = Reaction, group = Subject, colour = Subject)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Reaction Time over Days of Sleep Deprivation",
    y = "Average reaction time (ms)",
    x = "Number of days of sleep deprivation"
  ) +
  theme(legend.position = "none")
```

Fit a simple linear model (no random effects).

```{r lm_model}
lm_model <- lm(Reaction ~ Days, data = sleepstudy)
summary(lm_model)
```

This assumes everyone has the same intercept and slope but each subject behaves differently.

Fit a mixed-effects model (random intercepts).

```{r m1}
m1 <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
summary(m1)
```

* `Reaction` - The **response variable** (dependent variable).
* `Days` - A **fixed effect** (predictor that applies to everyone).
* `(1 | Subject)` - A **random effect** where each Subject has their own intercept.
    * `lme4::lmer()` (and similar functions), random effects always have the structure: `(random effects terms | grouping factor)`
    * `(1 | Subject)` means allow the intercept (1) to vary by `Subject`; in a model formula, 1 represents the intercept term, which is the baseline level of the response variable when all predictors = 0.
* `+` - Combine fixed and random effects.

The model says:

$$
\text{Reaction}*{ij} = \beta_0 + \beta_1 \text{Days}*{ij} + u_{0j} + \epsilon_{ij}
$$

where:

* $( i )$ = measurement within subject ( j ).
* $( j )$ = subject.
* $( \beta_0 )$: population-level intercept (average reaction time at Day 0).
* $( \beta_1 )$: population-level slope (effect of each additional day).
* $( u_{0j} )$: random intercept (how much each subject deviates from the average intercept).
* $( \epsilon_{ij} )$: residual error.

Each subject has their *own baseline* reaction time, but we assume the *same slope* (same rate of slowing down).
