---
title: "Mixed-effects models"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
})
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Mixed-effects models (also known as multilevel models, hierarchical models, or random-effects models) are extensions of linear regression that account for both fixed and random effects.

Given students’ test scores across multiple schools:

* Fixed effects are predictors that apply to everyone such as the effect of study hours.
* Random effects are effects that vary across groups such as school-to-school differences.

Each school might have a slightly different average score (intercept) or respond differently to study hours (slope).

Mixed-effects models handle this by modeling:

$$
\text{Score}*{ij} = \beta_0 + \beta_1 \text{Hours}*{ij} + u_{0j} + \epsilon_{ij}
$$

where:

* $( j )$: the group index, i.e., the school.
* $( i )$: the individual (observation) index within each group, i.e., the student within school $( j )$.

Each school $( j )$ has its own intercept $( u_{0j} )$ that represents how that school's mean score differs from the overall population mean $( \beta_0 )$.

Each student $( i )$ within that school has their own hours studied ($( \text{Hours}*{ij} )$) and residual error ($( \epsilon_{ij} )$).

## Random effects

A random effect is a variable whose levels are drawn from a larger population, not specifically fixed.

For instance:

* If we measured 10 schools, they are a sample from all possible schools.
* We assume their effects (intercepts/slopes) come from a distribution (often Normal).

**Random effects account for correlation among observations within the same group and share information across groups**.

## Example in R

Install dependencies.

```{r install_libraries, eval=FALSE}
install.packages("lme4")
```

Example data: Reaction times in a sleep deprivation study.

> These data are from the study described in Belenky et al. (2003), for the most sleep-deprived group (3 hours time-in-bed) and for the first 10 days of the study, up to the recovery period. The original study analyzed speed (1/(reaction time)) and treated day as a categorical rather than a continuous predictor.

```{r sleepstudy}
library(lme4)
data("sleepstudy")
head(sleepstudy)
```

Visualise the data.

```{r sleepstudy_by_subject}
ggplot(sleepstudy, aes(x = Days, y = Reaction, group = Subject, colour = Subject)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Reaction Time over Days of Sleep Deprivation",
    y = "Average reaction time (ms)",
    x = "Number of days of sleep deprivation"
  ) +
  theme(legend.position = "none")
```

Fit a simple linear model (no random effects).

```{r lm_model}
lm_model <- lm(Reaction ~ Days, data = sleepstudy)
summary(lm_model)
```

This assumes everyone has the same intercept and slope but each subject behaves differently.

Fit a mixed-effects model (random intercepts).

```{r m1}
m1 <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
summary(m1)
```

* `Reaction` - The **response variable** (dependent variable).
* `Days` - A **fixed effect** (predictor that applies to everyone).
* `(1 | Subject)` - A **random effect** where each Subject has their own intercept.
    * `lme4::lmer()` (and similar functions), random effects always have the structure: `(random effects terms | grouping factor)`
    * `(1 | Subject)` means allow the intercept (1) to vary by `Subject`; in a model formula, 1 represents the intercept term, which is the baseline level of the response variable when all predictors = 0.
* `+` - Combine fixed and random effects.

The model says:

$$
\text{Reaction}*{ij} = \beta_0 + \beta_1 \text{Days}*{ij} + u_{0j} + \epsilon_{ij}
$$

where:

* $( i )$ = measurement within subject ( j ).
* $( j )$ = subject.
* $( \beta_0 )$: population-level intercept (average reaction time at Day 0).
* $( \beta_1 )$: population-level slope (effect of each additional day).
* $( u_{0j} )$: random intercept (how much each subject deviates from the average intercept).
* $( \epsilon_{ij} )$: residual error.

Each subject has their *own baseline* reaction time, but we assume the *same slope* (same rate of slowing down).

Fit multiple models with increasing complexity and calculate AIC (Akaike Information Criterion) and BIC (Bayesian Information Criterion), which are information criteria used for model selection. They help you balance:

* Model fit (how well the model explains the data)
* Model complexity (how many parameters the model uses)

The actual numbers are meaningless by themselves and what matters is the relative comparison between models:

* Lower AIC/BIC = better model fit
* Delta AIC/BIC shows difference from best model
* Delta < 2: substantial support
* Delta 4-7: considerably less support
* Delta > 10: essentially no support

```{r multiple_models}
m0 <- lm(Reaction ~ Days, data = sleepstudy)

# Model 1: Random intercept only
m1 <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)

# Model 2: Random intercept and random slope
m2 <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)

# Model 3: Random intercept and random slope (uncorrelated)
m3 <- lmer(Reaction ~ Days + (Days || Subject), data = sleepstudy)

aic_values <- c(
  M0_lm = AIC(m0),
  M1_random_intercept = AIC(m1),
  M2_random_slope_corr = AIC(m2),
  M3_random_slope_uncorr = AIC(m3)
)

bic_values <- c(
  M0_lm = BIC(m0),
  M1_random_intercept = BIC(m1),
  M2_random_slope_corr = BIC(m2),
  M3_random_slope_uncorr = BIC(m3)
)

comparison_df <- data.frame(
  Model = c("M0: Linear model", 
            "M1: Random intercept", 
            "M2: Random intercept + slope (correlated)",
            "M3: Random intercept + slope (uncorrelated)"),
  AIC = round(aic_values, 2),
  BIC = round(bic_values, 2),
  Delta_AIC = round(aic_values - min(aic_values), 2),
  Delta_BIC = round(bic_values - min(bic_values), 2)
)

comparison_df
```
