---
title: "Monkey patch"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

[Monkey patch](https://en.wikipedia.org/wiki/Monkey_patch):

> In computer programming, monkey patching is a technique used to dynamically update the behavior of a piece of code at run-time. It is used to extend or modify the runtime code of dynamic languages such as Smalltalk, JavaScript, Objective-C, Ruby, Perl, Python, Groovy, Lisp, and Lua without altering the original source code.

Great question! In R, you can use the `assign()` function to monkeypatch by programmatically replacing or redefining a function, including functions in other environments like packages. This is especially useful when you want to inject a new version of a function into an environment like `package:stats`.

Monkey patch `stats::kmeans()` using `assign()` so that it prints a message before executing.

Save the original function.

```{r original_kmeans}
original_kmeans <- stats::kmeans
```

Define the new function.

```{r my_kmeans}
my_kmeans <- function(x, centers, ...) {
  message("Modified kmeans called")
  original_kmeans(x, centers, ...)
}
```

Use `assign()` to overwrite `kmeans` in the `stats` namespace. If run without `unlockBinding()`:

> Error in assign("kmeans", my_kmeans, envir = asNamespace("stats")) : 
>   cannot change value of locked binding for 'kmeans'

```{r assign}
unlockBinding("kmeans", asNamespace("stats"))
assign("kmeans", my_kmeans, envir = asNamespace("stats"))
lockBinding("kmeans", asNamespace("stats"))
```

Test.

```{r test}
set.seed(1984)
data <- matrix(rnorm(100), ncol = 2)
data_kmeans <- stats::kmeans(data, centers = 3)
```

Restore the Original

```{r restore}
unlockBinding("kmeans", asNamespace("stats"))
assign("kmeans", original_kmeans, envir = asNamespace("stats"))
lockBinding("kmeans", asNamespace("stats"))
```
