---
title: "Monte Carlo integration"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I recently came across this question:

>You have a function called random that randomly generates a number between 0 to 1. Use this to calculate pi.

Worded differently, the question is asking you to estimate pi using random numbers. As you can read in the [script for Life of Pi](https://www.scriptslug.com/assets/uploads/scripts/life-of-pi-2012.pdf):

>Pi, the sixteenth letter of the Greek alphabet, which is also used in mathematics to represent the ratio of any circle's circumference to its diameter - an irrational number of infinite length, usually rounded to three digits, as 3.14.

I later learned that the method used to answer this question is called [Monte Carlo integration](https://en.wikipedia.org/wiki/Monte_Carlo_integration). The idea is that you randomly generate many pairs of numbers, $x$ and $y$, representing points on the [Cartesian coordinate system](https://en.wikipedia.org/wiki/Cartesian_coordinate_system) and you check the ratio of points that lie inside a circle drawn on the same coordinate system to the total number of randomly generated points. This will give you an approximate estimation of the circle's area that can be used to estimate pi.

(The question uses a random number generator that generates numbers between 0 and 1 but I'll use -1 to 1 to go full circle.)

To see the method in action, let's draw a [unit circle](https://en.wikipedia.org/wiki/Unit_circle), which is simply a circle of unit radius, i.e., a radius of 1.

```{r plot_circle, fig.asp=1}
circle <- function(x) sqrt(1 - x^2)
circle_inv <- function(x) -sqrt(1 - x^2)
plot_circle <- function(){
  lim <- c(-1, 1)
  plot(1, xlim = lim, ylim = lim, type = "n", ylab = '', xlab = '')
  abline(h = 0, v = 0, lty = 3)
  curve(circle, -1, 1, add = TRUE)
  curve(circle_inv, -1, 1, add = TRUE)
}
plot_circle()
```

Now we randomly generate 1,000 (x,y) coordinates within the unit circle.

```{r random_points, fig.asp=1}
set.seed(1984)
x <- runif(1000, min = -1, max = 1)
y <- runif(1000, min = -1, max = 1)
plot_circle()
points(x, y, pch = 16)
```

The area of the square encompassing the unit circle is $(1+1)^2 = 4$, which represents all the dots. The area of the unit circle is the ratio of all the points inside the circle divided by all the points multiplied by 4.

Now, how do we check whether a point lies inside of our unit circle? We can use the [Pythagorean theorem](https://en.wikipedia.org/wiki/Pythagorean_theorem), which you may recall as $x^2 + y^2 = z^2$. The hypotenuse in this case is the radius of the circle (which is 1), so therefore if $\sqrt{x^2 + y^2} \leq 1$, then a point lies within the circle. Let's illustrate with the point (0.87, 0.87).

```{r outside}
x <- 0.87
y <- 0.87
sqrt((x^2+y^2))
```

Plot point (0.87, 0.87).

```{r outside_plot, fig.asp=1}
plot_circle()
x <- 0.87
y <- 0.87
points(x, y, pch = 16)
```

Now the point (-0.6, 0.6).

```{r inside}
x <- -0.6
y <- 0.6
sqrt((x^2+y^2))
```

Plot point (-0.6, 0.6).

```{r inside_plot, fig.asp=1}
plot_circle()
x <- -0.6
y <- 0.6
points(x, y, pch = 16)
```

But why is the area of unit circle an estimation of pi? Again as you may recall, the area of a circle is $\pi r^2$. When we have a radius that is 1, such as the unit circle, $area = \pi \times 1^2 = \pi$. Therefore the area of a unit circle is also the value of pi.

Now let's implement this into a function that we can use to estimate pi.

```{r estimate_pi}
estimate_pi <- function(n){
  set.seed(1984)
  x <- runif(n, min = -1, max = 1)
  y <- runif(n, min = -1, max = 1)
  sum(sqrt(x^2+y^2) < 1) / n * 4
}

sapply(c(100, 1000, 10000, 100000, 1000000, 5000000, 10000000, 50000000), estimate_pi)
```

Not the best way to estimate pi (since the estimation actually got worse between 5 and 10 million) but it illustrates one way in which random numbers can be useful.
