---
title: "Parallel computation in R"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
params:
  threads: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

As stated in the [foreach vignette](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html#conclusion):

>Much of parallel computing comes to doing three things: splitting the problem into pieces, executing the pieces in parallel, and combining the results back together.

There are several packages that make it easy to run tasks in parallel:

* The [parallel](https://bookdown.org/rdpeng/rprogdatascience/parallel-computation.html#the-parallel-package) package that comes with R.
* The [doParallel](https://cran.r-project.org/web/packages/doParallel/vignettes/gettingstartedParallel.pdf) package is a parallel backend for the `foreach` package and acts as an interface between `foreach` and the `parallel` package.
* The [BiocParallel](https://bioconductor.org/packages/release/bioc/html/BiocParallel.html) package tailored for use with Bioconductor.
* The [furrr](https://furrr.futureverse.org/) package parallelises mapping functions from the [purrr](https://purrr.tidyverse.org/) package.
* The [pbapply](https://github.com/psolymos/pbapply#parallel-backends) package, which is a package for adding a progress bar, but supports parallel execution.
* The [future](https://cran.r-project.org/web/packages/future/index.html) package, unified parallel and distributed processing in R for everyone.

### Why are there so many parallelisation packages?

R's parallel computing ecosystem has grown over more than two decades, and the variety reflects a mix of historical accident, ecosystem-specific needs, and evolving design ideas:

* **Historical core.** The base R `parallel` package (introduced in R 2.14.0 in 2011) was created by merging two earlier packages: `snow` (Simple Network of Workstations, 1999) for socket-based clusters, and `multicore` (2009) for forked workers on Unix. Most other packages either build on `parallel` or replace it.
* **Different programming styles.** Some users prefer imperative loops (`foreach`), some prefer apply-style functions (`*apply`, `bplapply`, `future_lapply`), and tidyverse users prefer functional mapping (`map`, `future_map`). Each style has at least one parallelisation package tailored to it.
* **Different ecosystems.** `BiocParallel` was written for [Bioconductor](https://bioconductor.org/), where workflows commonly run on HPC clusters with schedulers like SLURM or SGE, and where robust error handling and logging matter. `furrr` was written for the tidyverse. `doParallel` was written by Revolution Analytics (now Microsoft) as a backend for `foreach`.
* **Different backends.** Forking (Unix only), socket clusters (cross-platform), MPI, batch job schedulers, and cloud workers all need different machinery underneath. Some packages (`future`, `BiocParallel`) try to abstract this away; others expose it directly.
* **Progressive features.** Newer packages typically add capabilities the older ones lack: progress bars (`pbapply`), automatic detection of variables that need to be exported (`future`), reproducible parallel RNG, structured error handling, and so on.

A useful mental model is that there are really only **two parallelisation mechanisms** in R — forking and socket clusters — and most of the packages above are different *frontends* on top of those mechanisms.

### Forking vs socket clusters

Almost every package in this notebook is ultimately doing one of these two things:

* **Forking** copies the parent R process using the operating system's `fork()` call. Workers share memory with the parent (copy-on-write), so objects in the parent environment are automatically available to every worker — no explicit export step is needed. Forking starts up quickly and avoids data transfer cost, but it is **only available on Unix-like systems** (Linux and macOS) and can be unsafe inside multi-threaded host processes such as the RStudio GUI.
* **Socket clusters** spawn fresh R processes that communicate with the parent over network sockets. Each worker starts with an empty environment, so any objects, packages, or functions it needs must be explicitly transferred (`clusterExport`, `clusterEvalQ`, or — for `future`-based tools — automatic globals detection). Socket clusters are slower to start and have higher communication overhead, but they work on **all platforms including Windows**.

When you see `mclapply`, `MulticoreParam`, or `plan(multicore)`, that is forking. When you see `makeCluster`, `SnowParam`, `parLapply`, or `plan(multisession)`, that is a socket cluster.

### Frontends and backends

A second useful distinction is between *frontends* — the API you write your code against — and *backends* — what actually runs the work:

| Frontend                                                  | Typical backends                                                            |
| --------------------------------------------------------- | --------------------------------------------------------------------------- |
| `parallel::mclapply` / `parLapply`                        | fork / socket (built in)                                                    |
| `foreach::%dopar%`                                        | `doParallel`, `doMC`, `doSNOW`, `doFuture`, `doMPI`, ...                    |
| `BiocParallel::bplapply`                                  | `MulticoreParam`, `SnowParam`, `BatchtoolsParam`, ...                       |
| `future.apply::future_lapply`, `furrr::future_map`        | any `future` plan: `sequential`, `multicore`, `multisession`, `cluster`     |

The same loop body can usually be moved between frontends with little change, but the *backend* you pick determines startup cost, memory behaviour, OS portability, and how variables are shared with the workers.

## Timing processes with `system.time`

From `?proc.time`:

>The "user time" is the CPU time charged for the execution of user instructions of the calling process.
>
>The "system time" is the CPU time charged for execution by the system on behalf of the calling process.

Elapsed time is the amount of time that has elapsed/passed. The `user` and `system` time while sleeping is close to zero because the CPU is idly waiting and not executing anything.

```{r sleep}
system.time(
  Sys.sleep(5)
)
```

More information is provided on [Stack Overflow](https://stackoverflow.com/questions/5688949/what-are-user-and-system-times-measuring-in-r-system-timeexp-output):

>"User CPU time" gives the CPU time spent by the current process (i.e., the current R session and outside the kernel)
>
>"System CPU time" gives the CPU time spent by the kernel (the operating system) on behalf of the current process. The operating system is used for things like opening files, doing input or output, starting other processes, and looking at the system clock: operations that involve resources that many processes must share.

## Testing data

Create a list of 100 data frames each with 5,000 observations across 100 variables.

```{r create_data}
create_df <- function(n, m, seed = 1984){
  set.seed(seed)
  as.data.frame(
    matrix(
      data = rnorm(n = n * m),
      nrow = n,
      ncol = m
    )
  )
}

my_list <- lapply(1:100, function(x) create_df(5000, 100, x))
length(my_list)
```

## Number of threads used

This is a parameterised notebook; the number of threads used for the code examples is `r params$threads`.

```{r num_threads}
params$threads
```

## `parallel`

The `parallel` package ships with base R and is the foundation that almost every other package in this notebook builds on. It exposes both parallelisation mechanisms directly: forking (via `mclapply`, `mcmapply`, etc.) and socket clusters (via `makeCluster`, `parLapply`, `parSapply`, etc.). It is low-level and unopinionated — there is no progress reporting, no automatic globals detection, and error handling is bare-bones — but it has zero dependencies and is always available.

Load the `parallel` package.

```{r load_parallel}
library(parallel)
```

Create a summary of each variable in each data frame without parallelisation.

```{r lapply_summary}
system.time(
  my_sum <- lapply(my_list, summary)
)
```

The `mclapply` function can be used to process a list in parallel. Note that this function uses forking, which is not available on Windows.

```{r mclapply}
system.time(
  my_sum_mc <- mclapply(my_list, summary, mc.cores = params$threads)
)
```

Compare the two summaries.

```{r compare_my_sum}
identical(my_sum, my_sum_mc)
```

Another way to run the jobs in parallel is via sockets. For Windows users, you will need to use this method for parallelisation. In addition, you need to use the `parLapply` function instead of `mclapply`.

```{r parlapply}
cl <- makeCluster(params$threads)
system.time(
  my_sum_sock <- parLapply(cl, my_list, summary)
)
stopCluster(cl)

identical(my_sum_mc, my_sum_sock)
```

Note that forking is faster.

### Worker environment

If you run the code below:

```{r no_export, eval=FALSE}
cl <- makeCluster(params$threads)
system.time(
  test <- parLapply(cl, seq_len(params$threads), function(x){
    class(my_list)
  })
)
stopCluster(cl)
```

you will get the following error:

```
Error in checkForRemoteErrors(val) : 
  4 nodes produced errors; first error: object 'my_list' not found
```

This is because each worker is using a different environment. To make the `my_list` object available to each worker, we use the `clusterExport()` function.

```{r cluster_export}
cl <- makeCluster(params$threads)
clusterExport(cl, list("my_list"))
system.time(
  test2 <- parSapply(cl, seq_len(params$threads), function(x){
    class(my_list)
  })
)
stopCluster(cl)

test2
```

## `pbapply`

`pbapply` is primarily a progress-bar package: it provides drop-in replacements for `*apply` (`pblapply`, `pbsapply`, etc.) that show a progress bar in interactive sessions. Parallel support is layered on top — internally it dispatches to `parLapply` when given a cluster object and to `mclapply` when given an integer worker count — so you get progress reporting "for free" without changing parallelisation backend. Reach for it when a long-running job benefits from a visible progress indicator.

Parallelisation with a progress bar! From the help page of `pblapply`:

>Parallel processing can be enabled through the cl argument. parLapply is called when cl is a 'cluster' object, mclapply is called when cl is an integer. Showing the progress bar increases the communication overhead between the main process and nodes / child processes compared to the parallel equivalents of the functions without the progress bar. The functions fall back to their original equivalents when the progress bar is disabled (i.e. getOption("pboptions")$type == "none" or dopb() is FALSE). This is the default when interactive() if FALSE (i.e. called from command line R script).

```{r pbapply_socket}
library(pbapply)
cl <- makeCluster(params$threads)
system.time(
  my_sum_pb <- pblapply(my_list, summary, cl = cl)
)
stopCluster(cl)

identical(my_sum_mc, my_sum_pb)
```

Use `mclapply`.

```{r pbapply_forking}
system.time(
  my_sum_pb_fork <- pblapply(my_list, summary, cl = params$threads)
)

identical(my_sum_pb, my_sum_pb_fork)
```

## `doParallel`

`doParallel` is a *backend* for the [`foreach`](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html) frontend. `foreach` looks like a `for` loop but returns a value (like `lapply`) and — crucially — is parallel-backend agnostic: the same `foreach(...) %dopar% { ... }` block can be run sequentially, on a forked or socket cluster (via `doParallel`), on an MPI cluster (`doMPI`), or via the `future` framework (`doFuture`), depending on which backend you register. Reach for `foreach` when your loop body is more complex than a single function application (e.g. multiple result accumulators, custom `.combine` strategies, or nested loops with `%:%`).

Load the `doParallel` package.

```{r load_doparallel}
library(doParallel)
```

Using `foreach`.

```{r foreach_dopar}
cl <- makeCluster(params$threads)
registerDoParallel(cl)

system.time(
  my_sum_dopar <- foreach(l = my_list) %dopar% {
    summary(l)
  }
)
stopCluster(cl)

identical(my_sum_mc, my_sum_dopar)
```

## `BiocParallel`

`BiocParallel` is Bioconductor's unified parallelisation interface, designed for the kinds of workloads common in genomics: long-running jobs over large objects, where workers can fail mid-run and where logs and checkpoints matter. The frontend is a small set of `bp*` functions (`bplapply`, `bpmapply`, `bpiterate`, ...) that all take a `BPPARAM` argument describing the backend:

* `MulticoreParam` — forking on Unix (falls back to `SnowParam` on Windows).
* `SnowParam` — socket cluster; can be `type = "PSOCK"` (default) or `type = "FORK"`.
* `SerialParam` — no parallelism, useful for debugging.
* `BatchtoolsParam` — submit jobs to HPC schedulers (SLURM, SGE, LSF, Torque, ...) via the [`batchtools`](https://mllg.github.io/batchtools/) package.

Reach for `BiocParallel` when you are working inside the Bioconductor ecosystem, when you need structured error handling and per-task logging, or when you want the same code to run on a laptop and a cluster by changing only the `BPPARAM`.

Load `BiocParallel`.

```{r load_biocparallel}
library(BiocParallel)
```

Using `bplapply`.

```{r bplapply}
param <- SnowParam(workers = params$threads, type = "SOCK")
system.time(
  my_sum_bp <- bplapply(my_list, summary, BPPARAM = param)
)

identical(my_sum_mc, my_sum_bp)
```

Forking.

```{r bplapply_fork}
param <- SnowParam(workers = params$threads, type = "FORK")
system.time(
  my_sum_bp_fork <- bplapply(my_list, summary, BPPARAM = param)
)

identical(my_sum_bp, my_sum_bp_fork)
```

Using `MulticoreParam`.

```{r bplapply_multicore}
param <- MulticoreParam(workers = params$threads, progressbar = FALSE)
system.time(
  my_sum_bp_mc <- bplapply(my_list, summary, BPPARAM = param)
)

identical(my_sum_bp_fork, my_sum_bp_mc)
```

## `furrr`

`furrr` is a tidyverse-flavoured parallelisation package: it provides `future_map`, `future_map2`, `future_pmap`, `future_walk`, etc., which are drop-in replacements for the corresponding `purrr` functions. Under the hood it uses the [`future`](https://cran.r-project.org/web/packages/future/index.html) framework, which means switching between sequential, multicore, multisession, and cluster execution is just a matter of calling `plan()` — your `future_map(...)` code does not change. `furrr` also handles globals (variables captured from the enclosing environment) and packages automatically, so you rarely need an explicit `clusterExport`-style step.

Load required libraries.

```{r rrr}
library(furrr)
library(purrr)
```

Map without parallelisation.

```{r map}
system.time(
  my_sum_pur <- map(my_list, summary)
)

identical(my_sum_mc, my_sum_pur)
```

Map with parallelisation.

```{r future_map}
plan(multisession, workers = params$threads)
system.time(
  my_sum_fur <- future_map(my_list, summary)
)

identical(my_sum_pur, my_sum_fur)
```

## `future`

The `future` package, by Henrik Bengtsson, is a unified abstraction for parallel and distributed computing in R. The central idea is the *future*: a placeholder for a value that may still be computing somewhere — possibly in another process, possibly on another machine. Code that uses futures is written once and the choice of *where* to run it is made separately, by calling `plan()`:

* `plan(sequential)` — run in the current process (no parallelism).
* `plan(multicore)` — fork on Unix; not available on Windows or inside RStudio.
* `plan(multisession)` — socket cluster of background R sessions; works everywhere.
* `plan(cluster)` — explicit `parallel::makeCluster` cluster, including remote machines over SSH.
* `plan(remote)` / `plan(batchtools_*)` — remote workers, HPC schedulers, etc.

`future` itself is mostly a low-level engine; in practice you usually use one of its frontends — `future.apply` (`*apply`-style), `furrr` (`purrr`-style), or `doFuture` (as a `foreach` backend). The companion package [`future.apply`](https://future.apply.futureverse.org/) used below provides parallel versions of `lapply`, `sapply`, `vapply`, `mapply`, and `apply`.

Load required libraries.

```{r future}
library(future)
library(future.apply)
```

Map with parallelisation using `future_lapply()`.

```{r future_lapply}
plan(multisession, workers = params$threads)
system.time(
  my_sum_future_lapply <- future_lapply(my_list, summary)
)

identical(my_sum, my_sum_future_lapply)
```

## Summary

So, which package should you use? A few rules of thumb:

* **Stick with what your ecosystem uses.** If you're working in Bioconductor, use `BiocParallel`; if you're already writing tidyverse / `purrr` code, use `furrr`. The cost of consistency with surrounding code is usually worth more than a small performance difference.
* **For a quick parallel `lapply`, reach for base `parallel`.** It's already installed, has no dependencies, and `mclapply` is a one-line change from `lapply` on Unix.
* **For complex loop bodies, use `foreach` + `doParallel`** (or `doFuture`). `foreach` shines when you need custom result combination (`.combine`), nested loops (`%:%`), or to accumulate multiple outputs per iteration.
* **For portability across laptop / server / HPC, use `future`.** Writing your code once against `future_lapply` / `future_map` and switching backends with `plan()` is the most flexible option, and the `future` framework also handles globals, packages, and reproducible RNG (`future.seed = TRUE`) for you.
* **Add `pbapply` when you want a progress bar** on an otherwise straightforward parallel `*apply`.

Two cross-cutting points worth remembering:

* **Forking is faster than sockets, but Unix-only.** Forked workers share memory copy-on-write, so there is no startup data transfer and no need to export variables. On Windows (or inside some GUI host processes) you have no choice but to use a socket cluster.
* **Parallel is not always faster.** For very fast per-element work, the cost of starting workers and shipping data to them can dominate. If `lapply` already runs in milliseconds, parallelising it will usually make things slower.
