---
title: "Analysing 20k cells"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# https://stackoverflow.com/questions/30237310/setting-work-directory-in-knitr-using-opts-chunksetroot-dir-doesnt-wor
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(patchwork)
library(harmony)
library(Seurat)
library(BPCells)
library(SingleR)
library(celldex)
```

## Data

Read HDF5 file paths into a list.

```{r hdf5_files}
hdf5_files <- list.files(path = "data", pattern = "5k_Human", full.names = TRUE)
hdf5_files
```

Read raw counts into a list of matrices.

```{r mats}
mats <- purrr::map(seq_along(hdf5_files), function(x){
  my_mat <- BPCells::open_matrix_10x_hdf5(hdf5_files[x])
  colnames(my_mat) <- paste0('donor', x, '_', colnames(my_mat))
  my_mat
})
str(mats, max.level = 1)
```

Get gene names.

```{r mat_tmp}
mat_tmp <- Seurat::Read10X_h5(hdf5_files[1])
my_genes <- row.names(mat_tmp)
rm(mat_tmp)
gc()
```


Combine the matrices.

```{r combined_matrix}
combined_matrix <- do.call(cbind, mats)
row.names(combined_matrix) <- my_genes
combined_matrix
```

## Seurat object

Create Seurat object from the list of matrices.

```{r seurat_object}
pbmc20k <- CreateSeuratObject(
  counts = combined_matrix,
  min.cells = 3,
  min.features = 200
)
pbmc20k
```

Counts layer.

```{r pbmc20k_counts_layer}
pbmc20k@assays$RNA$counts
```

Donor information in `orig.ident`.

```{r metadata}
head(pbmc20k@meta.data)
```

Use {BPCells} to store the matrix in the Seurat object to on-disk matrices. Note, that this is only possible for V5 assays.

```{r write_matrix}
BPCells::write_matrix_dir(
  mat = pbmc20k@assays$RNA$counts,
  dir = 'data/pbmc20k',
  overwrite = TRUE
)

pbmc20k.mat <- open_matrix_dir(dir = "data/pbmc20k")

pbmc20k@assays$RNA$counts <- pbmc20k.mat
pbmc20k@assays$RNA$counts
```

## Seurat workflow

Process with the Seurat 4 workflow.

```{r seurat_wf_v4}
options(future.globals.maxSize = 2 * 1024^3)

fixed_PrepDR5 <- function(object, features = NULL, layer = 'scale.data', verbose = TRUE) {
  layer <- layer[1L]
  olayer <- layer
  layer <- SeuratObject::Layers(object = object, search = layer)
  if (is.null(layer)) {
    abort(paste0("No layer matching pattern '", olayer, "' not found. Please run ScaleData and retry"))
  }
  data.use <- SeuratObject::LayerData(object = object, layer = layer)
  features <- features %||% VariableFeatures(object = object)
  if (!length(x = features)) {
    stop("No variable features, run FindVariableFeatures() or provide a vector of features", call. = FALSE)
  }
  if (is(data.use, "IterableMatrix")) {
    features.var <- BPCells::matrix_stats(matrix=data.use, row_stats="variance")$row_stats["variance",]
  } else {
    features.var <- apply(X = data.use, MARGIN = 1L, FUN = var)
  }
  features.keep <- features[features.var > 0]
  if (!length(x = features.keep)) {
    stop("None of the requested features have any variance", call. = FALSE)
  } else if (length(x = features.keep) < length(x = features)) {
    exclude <- setdiff(x = features, y = features.keep)
    if (isTRUE(x = verbose)) {
      warning(
        "The following ",
        length(x = exclude),
        " features requested have zero variance; running reduction without them: ",
        paste(exclude, collapse = ', '),
        call. = FALSE,
        immediate. = TRUE
      )
    }
  }
  features <- features.keep
  features <- features[!is.na(x = features)]
  features.use <- features[features %in% rownames(data.use)]
  if(!isTRUE(all.equal(features, features.use))) {
    missing_features <- setdiff(features, features.use)
    if(length(missing_features) > 0) {
    warning_message <- paste("The following features were not available: ",
                             paste(missing_features, collapse = ", "),
                             ".", sep = "")
    warning(warning_message, immediate. = TRUE)
    }
  }
  data.use <- data.use[features.use, ]
  return(data.use)
}

assignInNamespace('PrepDR5', fixed_PrepDR5, 'Seurat')

seurat_wf_v4 <- function(seurat_obj, scale_factor = 1e4, num_features = 2000, num_pcs = 30, cluster_res = 0.5, debug_flag = FALSE){
  
  seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = scale_factor, verbose = debug_flag)
  seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = 'vst', nfeatures = num_features, verbose = debug_flag)
  seurat_obj <- ScaleData(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunPCA(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunHarmony(seurat_obj, "orig.ident")
  seurat_obj <- RunUMAP(seurat_obj, reduction = "harmony",  dims = 1:num_pcs, verbose = debug_flag)
  
  seurat_obj
}

pbmc20k <- seurat_wf_v4(pbmc20k)
```

Normalised and scaled data are stored as `IterableMatrix` objects.

```{r iterable_matrices}
pbmc20k@assays$RNA$data
pbmc20k@assays$RNA$scale.data
```

UMAP.

```{r umap}
DimPlot(pbmc20k, reduction = "umap", group.by = "orig.ident", pt.size = .1)
```

## Annotation

Annotate using {SingleR}.

```{r singler}
monaco_immune <- fetchReference("monaco_immune", "2024-02-26")
monaco_immune

pbmc20k.anno <- SingleR(
  test=as(pbmc20k@assays$RNA$data, "sparseMatrix"),
  ref=monaco_immune,
  labels=colData(monaco_immune)$label.main
)

head(pbmc20k.anno)
```

Add annotations to metadata.

```{r add_anno}
cbind(
  pbmc20k@meta.data,
  as.data.frame(pbmc20k.anno)
) -> pbmc20k@meta.data
```

UMAP with annotations.

```{r umap_with_labels}
DimPlot(pbmc20k, reduction = "umap", group.by = "labels", pt.size = .1, label = TRUE, repel = TRUE)
```

## Saving and loading

If you save your object and load it in in the future, Seurat will access the on-disk matrices by their path, which is stored in the assay level data. To make it easy to ensure these are saved in the same place, we provide new functionality to the `SaveSeuratRds()` function. In this function, you specify your filename. The pointer to the path in the Seurat object will change to the current directory.

This also makes it easy to share your Seurat objects with BPCells matrices by sharing a folder that contains both the object and the BPCells directory.

Make sure you use a different directory than where the on-disk matrices are stored or they will be recursively copied.

```{r export_seurat}
output_dir <- "data/pbmc20k_seurat"
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

SaveSeuratRds(
  object = pbmc20k,
  file = paste0(output_dir, "/pbmc20k.rds")
)
list.files(output_dir)
list.files(paste0(output_dir, "/pbmc20k"))
```

Need to use `LoadSeuratRds()` to load or else none of the layers will be imported.

```{r import_seurat}
pbmc20k_import <- LoadSeuratRds(paste0(output_dir, '/pbmc20k.rds'))
pbmc20k_import
pbmc20k_import@assays$RNA$counts
```
