---
title: "Making a heatmap in R with the pheatmap package"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pheatmap)
library(dendextend)
library(grid)
library(gridExtra)
```

## Example data

Load data and subset for demonstration purposes.

```{r example_data}
example_file <- "https://davetang.org/file/TagSeqExample.tab"
data <- read.delim(example_file, header = TRUE, row.names = "gene")
data_subset <- as.matrix(data[rowSums(data)>50000,])
dim(data_subset)
```

## Default heatmap

Default heatmap using `pheatmap`.

```{r pheatmap_default, fig.width=7, fig.height=7}
pheatmap(data_subset)
```

Manually scale the rows of the dataset.

```{r pheatmap_default_own_scale, fig.width=7, fig.height=7}
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

data_subset_norm <- t(apply(data_subset, 1, cal_z_score))
pheatmap(data_subset_norm)
```

Using `scale = "row"` produces the same heatmap when we manually scaled the data.

```{r pheatmap_scale, fig.width=7, fig.height=7}
pheatmap(data_subset, scale = "row")
```

## Adding a title

Add a title using `main`.

```{r pheatmap_title_main, fig.width=7, fig.height=7}
pheatmap(data_subset, main = "My title")
```

Add a title using `textGrob`; you will need the `grid` and `gridExtra` packages.

```{r pheatmap_title_textgrob, fig.width=7, fig.height=7}
my_title <- textGrob("My title", gp = gpar(fontsize = 21, fontface = "bold"))
one <- pheatmap(data_subset, silent = TRUE)
grid.arrange(grobs = list(my_title, one[[4]]), heights = c(0.1, 1))
```

## Two heatmaps

Use `grid.arrange` to arrange multiple heatmaps.

```{r two_heatmaps}
one <- pheatmap(data_subset, silent = TRUE)
two <- pheatmap(data_subset, silent = TRUE)

grid.arrange(grobs = list(one[[4]], two[[4]]))
```

## Dendrogram

Dendrogram results from pheatmap().

```{r pheatmap_result, fig.height=8, fig.width=10}
par(mar = c(3.1, 2.1, 1.1, 5.1))

my_heatmap <- pheatmap(data_subset, silent = TRUE)
names(my_heatmap)

my_heatmap$tree_row %>%
  as.dendrogram() %>%
  plot(horiz = TRUE)
```

Reproduce the gene dendrogram.

```{r gene_dendrogram, fig.height=8, fig.width=10}
par(mar = c(3.1, 2.1, 1.1, 5.1))

my_hclust_gene <- hclust(dist(data_subset), method = "complete")
my_hclust_gene$height

as.dendrogram(my_hclust_gene) %>%
  plot(horiz = TRUE)
```

Obtaining the gene IDs as per the order of the dendrogram (from top to bottom).

```{r heatmap_gene}
rev(row.names(data_subset)[my_hclust_gene$order])
```

Reproduce the sample dendrogram.

```{r sample_dendrogram}
my_hclust_sample <- hclust(dist(t(data_subset)), method = "complete")

as.dendrogram(my_hclust_sample) %>%
  plot()
```

## Heatmap annotations

Add row and column annotations.

```{r annotation, fig.height=7, fig.width=7}
my_gene_col <- cutree(tree = as.dendrogram(my_hclust_gene), k = 2)
my_gene_col <- data.frame(cluster = ifelse(test = my_gene_col == 1, yes = "cluster 1", no = "cluster 2"))

my_sample_col <- data.frame(sample = rep(c("tumour", "normal"), c(4,2)))
row.names(my_sample_col) <- colnames(data_subset)

set.seed(1984)
my_random <- as.factor(sample(x = 1:2, size = nrow(my_gene_col), replace = TRUE))
my_gene_col$random <- my_random

pheatmap(data_subset, annotation_row = my_gene_col, annotation_col = my_sample_col)
```

Define your own column order by modifying `data_subset` and setting `cluster_cols` to FALSE.

```{r own_col_order}
my_col_order <- c("N2", "T1a", "N1", "T1b", "T2", "T3")

pheatmap(
  data_subset[, my_col_order],
  annotation_col = my_sample_col,
  cluster_cols = FALSE
)
```

## Gene clusters

Adding more clusters.

```{r more_clusters, fig.height=7, fig.width=7}
my_gene_col <- cutree(tree = as.dendrogram(my_hclust_gene), k = 6)
my_gene_col <- data.frame(cluster = paste0("cluster ", my_gene_col), row.names = names(my_gene_col))

my_sample_col <- data.frame(sample = rep(c("tumour", "normal"), c(4,2)))
row.names(my_sample_col) <- colnames(data_subset)

set.seed(1984)
my_random <- as.factor(sample(x = 1:2, size = nrow(my_gene_col), replace = TRUE))
my_gene_col$random <- my_random

pheatmap(data_subset, annotation_row = my_gene_col, annotation_col = my_sample_col)
```

Change annotation colours and ordering.

```{r annotation_2}
my_gene_col <- cutree(tree = as.dendrogram(my_hclust_gene), k = 2)
my_gene_col <- data.frame(cluster = ifelse(test = my_gene_col == 1, yes = "cluster1", no = "cluster2"))

my_sample_col <- data.frame(sample = rep(c("tumour", "normal"), c(4,2)))
row.names(my_sample_col) <- colnames(data_subset)

# change order
my_sample_col$sample <- factor(my_sample_col$sample, levels = c("normal", "tumour"))

set.seed(1984)
my_random <- as.factor(sample(x = c("random1", "random2"), size = nrow(my_gene_col), replace = TRUE))
my_gene_col$random <- my_random

my_colour = list(
    sample = c(normal = "#5977ff", tumour = "#f74747"),
    random = c(random1 = "#82ed82", random2 = "#9e82ed"),
    cluster = c(cluster1 = "#e89829", cluster2 = "#cc4ee0")
)

p <- pheatmap(data_subset,
              annotation_colors = my_colour,
              annotation_row = my_gene_col,
              annotation_col = my_sample_col,
              cellheight = 7,
              cellwidth = 18)

save_pheatmap_png <- function(x, filename, width=1200, height=1000, res = 150) {
  png(filename, width = width, height = height, res = res)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}

# not run
# save_pheatmap_png(p, "heatmap_colour.png")
```

## Breaks

Introduce breaks by cutting the dendrogram.

```{r break, fig.height=7, fig.width=7}
pheatmap(
  data_subset,
  annotation_row = my_gene_col,
  annotation_col = my_sample_col,
  cutree_rows = 2,
  cutree_cols = 2
)
```
