---
title: "Network graph using R Plotly"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

Plotly can be used to plot a [Network Graph in R](https://plotly.com/r/network-graphs/). The documentation does not work, so here's my implementation.

## Dependencies

```{r install_packages, message=FALSE, warning=FALSE}
packages <- c('plotly', 'igraph', 'igraphdata', 'sna')

sapply(packages, function(x){
  y <- require(x, character.only = TRUE)
  if(y == FALSE){
    install.packages(x, quiet = TRUE)
    library(x, character.only = TRUE)
  }
  as.character(packageVersion(x))
})
```

## Example

Load [Zachary's karate club](https://en.wikipedia.org/wiki/Zachary%27s_karate_club). Network description from Wikipedia:

>A social network of a karate club was studied by Wayne W. Zachary for a period of three years from 1970 to 1972. The network captures 34 members of a karate club, documenting links between pairs of members who interacted outside the club. During the study a conflict arose between the administrator "John A" and instructor "Mr. Hi" (pseudonyms), which led to the split of the club into two. Half of the members formed a new club around Mr. Hi; members from the other part found a new instructor or gave up karate. Based on collected data Zachary correctly assigned all but one member of the club to the groups they actually joined after the split.

```{r karate}
data(karate, package="igraphdata")
karate
```

Upgrade graph.

```{r upgrade_karate}
G <- upgrade_graph(karate)
str(G)
```

This is how the graph is supposed to look, plotted using `igraph`.

```{r plot_igraph}
set.seed(1984)
L <- layout_nicely(G)
plot.igraph(G, layout = L)
```

The colour can be obtained from `vertex_attr`, which can query vertex attributes of a graph.

```{r vertex_attr}
vertex_attr(G, 'color')
```

The function `layout_nicely`:

>This function tries to choose an appropriate graph layout algorithm for the graph, automatically, based on a simple algorithm.

```{r check_out_layout}
head(L)
```

We can get the name of the vertices using `V`.

```{r vertices}
vs <- V(G)
vs
```

The edge list shows the connections.

```{r edge_list}
el <- as.data.frame(get.edgelist(G))
head(el)
```

Create the network with just the nodes using our layout `L`.

```{r nodes}
network <- plot_ly(
  x = ~L[, 1],
  y = ~L[, 2],
  mode = "markers",
  text = vs$label,
  hoverinfo = "text",
  type = "scatter",
  size = I(42),
  color = as.character(vertex_attr(G, 'color')),
  colors = c('orange', 'skyblue'),
  showlegend=FALSE
)
network
```

The graph above lacks the edges, which we will manually create. For example, these two nodes need to connect.

```{r eg1}
el[1, ]
```

The layout contains the coordinates of the nodes but is not named.

```{r tail_layout}
tail(L)
```

We can get the names using `names`.

```{r vertice_names}
my_layout <- L
row.names(my_layout) <- names(V(G))
tail(my_layout)
```

To get the (x, y) coordinates, we just subset `my_layout`.

```{r get_xy}
get_xy <- function(x){
  my_layout[x, ]
}

get_xy('Mr Hi')
```

Get the (x, y) coordinates of every node in the edge list.

```{r get_xy_el}
xy1 <- t(
  apply(el, 1, function(x){
    get_xy(x[1])
  })
)

xy2 <- t(
  apply(el, 1, function(x){
    get_xy(x[2])
  })
)

head(xy1)
```

Build the list of lines that will be used to connect the nodes.

```{r build_lines}
my_line <- list(
  type = "line",
  line = list(color = "#030303", width = 0.3),
  xref = "x",
  yref = "y"
)

my_lines <- lapply(seq_along(xy1[, 1]), function(x){
  c(
    my_line,
    x0 = xy1[x, 1],
    y0 = xy1[x, 2],
    x1 = xy2[x, 1],
    y1 = xy2[x, 2]
  )
})
```

Plot the graph using `layout` to modify the default layout.

```{r layout_graph}
axis <- list(title = "", showgrid = FALSE, showticklabels = FALSE, zeroline = FALSE)

layout(
  network,
  title = "Zachary's karate club Network",
  shapes = my_lines,
  xaxis = axis,
  yaxis = axis
)
```

## Plot igraph with Plotly

Incorporate all the steps above into one function but without colours.

```{r plotly_igraph}
plotly_igraph <- function(G, L = layout_nicely(G), my_title = NULL){
  vs <- V(G)
  el <- as.data.frame(get.edgelist(G))
  network <- plot_ly(
    x = ~L[, 1],
    y = ~L[, 2],
    mode = "markers",
    text = names(vs),
    hoverinfo = "text",
    type = "scatter",
    size = I(42),
    showlegend=FALSE
  )
  row.names(L) <- names(vs)
  get_xy <- function(x){
    L[x, ]
  }
  xy1 <- t(
    apply(el, 1, function(x){
      get_xy(x[1])
    })
  )
  
  xy2 <- t(
    apply(el, 1, function(x){
      get_xy(x[2])
    })
  )
  
  my_line <- list(
    type = "line",
    line = list(color = "#030303", width = 0.3),
    xref = "x",
    yref = "y"
  )
  
  my_lines <- lapply(seq_along(xy1[, 1]), function(x){
    c(
      my_line,
      x0 = xy1[x, 1],
      y0 = xy1[x, 2],
      x1 = xy2[x, 1],
      y1 = xy2[x, 2]
    )
  })
  
  axis <- list(title = "", showgrid = FALSE, showticklabels = FALSE, zeroline = FALSE)

  layout(
    network,
    title = my_title,
    shapes = my_lines,
    xaxis = axis,
    yaxis = axis
  )
  
}

set.seed(1984)
plotly_igraph(G, my_title = "Zachary's karate club Network")
```

## Graphviz dot file

The [read.dot](https://search.r-project.org/CRAN/refmans/sna/html/read.dot.html) function can read network information from a Graphviz's DOT format file, returning an adjacency matrix.

```{r read_dot}
adj_mat <- read.dot("data/rnaseq_variant_calling.dot")
dim(adj_mat)
```

The adjacency matrix is directional. `CALL_markduplicates` -> `CALL_split_n_cigar_reads` and not the other way around.

```{r check_adj_mat}
adj_mat[1:2, 1:2]
```

[Convert matrix to adjacency list](adj_list.html).

```{r adj_list}
as_tibble(adj_mat, rownames = "parent") |>
  pivot_longer(-parent, names_to='child') |>
  filter(value == 1) |>
  select(-value) |>
  mutate(parent = str_trim(parent, 'both'), child = str_trim(child, 'both')) -> adj_list

head(adj_list)
```

Create network.

```{r plot_net}
net <- graph.data.frame(adj_list, directed = TRUE)
set.seed(1984)
plot(net, layout = layout_nicely)
```

Need to make the lines into arrows.

```{r plotly_igraph_net}
set.seed(1984)
plotly_igraph(net, L = layout_nicely(net))
```
