---
title: "The tidyr pivot_longer and pivot_wider functions using base R"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
```

# Introduction

I use [tidyverse packages](https://www.tidyverse.org/) a lot and I prefer them
over base R functions, especially when it comes to plotting. However, sometimes
I want to write an R script with no dependencies. This is known as using base R
functions, i.e. using functions that come with R. Theoretically this means that
anyone with R can run the script, if I don't use any features introduced in
later versions of R, like the base R pipe.

The `pivot_longer` and `pivot_wider` functions in the [tidyr
package](https://tidyr.tidyverse.org/) converts data into long and wide format,
respectively. You may already be familiar with wide data; one example of wide
data is a gene expression data, where gene expression for a gene is measured and
stored in different tissues.

```{r gene_exp}
gene_exp <- read.delim(
  file = "https://davetang.org/file/TagSeqExample.tab",
  header = TRUE 
)

head(gene_exp)
```

Long data looks like this.

```{r gene_exp_long}
tidyr::pivot_longer(
  data = gene_exp,
  cols = -gene,
  names_to = "sample",
  values_to = "count"
) -> gene_exp_long

head(gene_exp_long)
```

There are advantages to using wide and long format but I typically convert my
wide data to long format for use with [ggplot2](https://ggplot2.tidyverse.org/).

```{r gene_exp_long_plot}
library(ggplot2)

ggplot(gene_exp_long[1:(6*20), ], aes(gene, count, fill = sample)) +
  geom_col(position = position_dodge()) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank()) +
  NULL
```

Converting long data back to wide data can be done using `pivot_wider`.

```{r pivot_wider}
tidyr::pivot_wider(
  data = gene_exp_long,
  id_cols = gene,
  names_from = sample,
  values_from = count
)
```

How do we do this using base R?

# Reshape

The documentation for `reshape()` describes the function as:

> This function reshapes a data frame between "wide" format (with repeated
> measurements in separate columns of the same row) and "long" format (with the
> repeated measurements in separate rows).

The documentation also shows how `reshape()` is typically used:

* Typical usage for converting from long to wide format:

```
# reshape(data, direction = "wide",
#         idvar = "___", timevar = "___", # mandatory
#         v.names = c(___),    # time-varying variables
#         varying = list(___)) # auto-generated if missing
```

* Typical usage for converting from wide to long format:

```
# reshape(data, direction = "long",
#         varying = c(___), # vector 
#         sep)              # to help guess 'v.names' and 'times'
```

Wide to long using `reshape`.

```{r reshape_long}
reshape(
  data = gene_exp,
  direction = "long",
  varying = colnames(gene_exp)[-1],
  v.names = "count",
  times = colnames(gene_exp)[-1],
  timevar = "sample"
) -> out

# order by gene like pivot_longer
out <- out[order(out$gene), ]

# remove row names
row.names(out) <- NULL

# remove id column
out$id <- NULL

head(out)

table(out$count == gene_exp_long$count)
```

We achieved the same* result using `reshape` but with a bit more typing. (*I
simply compared the count values above instead of using `identical` or
`all.equal` because `reshape` adds attributes to the object that make it
different to the `pivot_longer` object.)

The arguments for `varying` and `times` should be the column names of the data
frame minus the variable to keep constant. `v.names` corresponds to `values_to`
and `timevar` corresponds to `names_to` in `pivot_longer`.

Long to wide using `reshape`.

```{r reshape_wide}
reshape(
  data = out,
  direction = "wide",
  idvar = "gene",
  timevar = "sample",
  v.names = "count"
) -> out2

colnames(out2) <- sub("^count\\.", "", colnames(out2))
head(gene_exp)
head(out2)
```

# Conclusions

R is a statistical language and the design/implementation of functions, their
arguments, and documentation reflect this. I'm not a statistician and reading
the documentation for base R functions is difficult for me. This is one reason
why I prefer the Tidyverse.

However, as I mentioned in the introduction, there are times when I want an R
script to have little to no dependencies. In one of my scripts, I need to
convert data back to wide format and used `pivot_wider`. But now I can use the
base R function `reshape`.

## Further reading

* [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
