---
title: "Using the R seqinr package"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

The [seqinr package](https://cran.r-project.org/web/packages/seqinr/index.html) provides many useful functions for working with biological sequences in R. We will use data from [NONCODE](http://noncode.org/download.php) to demonstrate some features of `seqinr`.

```
wget 'http://www.noncode.org/datadownload/ncrna_NONCODE[v3.0].fasta.tar.gz'
tar xzf ncrna_NONCODE\[v3.0\].fasta.tar.gz
mv ncrna_NONCODE\[v3.0\].fasta ncrna_noncode_v3.fa
cat ncrna_noncode_v3.fa | grep "^>" | wc -l
411553
```

Show last couple of entries.

```{bash}
cat data/ncrna_noncode_v3.fa | grep "^>" | tail -3
```

Install (if missing) and load `seqinr`.

```{r seqinr_package, message=FALSE, warning=FALSE}
if(!require("seqinr", quietly = TRUE)){
  install.packages("seqinr")
}
library("seqinr")
```

The `read.fasta` function is used to load a FASTA file and we will use it to load `ncrna_noncode_v3.fa`.

```{r read.fasta}
ncrna <- read.fasta("data/ncrna_noncode_v3.fa")
length(ncrna)
```

The entries are saved in a list.

```{r ncrna_class}
class(ncrna)
```

Each list item is named after the first annotation in the FASTA file.

```{r ncrna_names}
head(names(ncrna))
```

Check the first entry, which is stored in index 2, as the first entry is a fake FASTA entry that contains some information on the annotations stored in the FASTA file.

```{r entry_2}
ncrna[[2]]
```

Count nucleotides.

```{r count_nuc}
count(ncrna[[2]], 1)
```

Count di-nucleotides.

```{r count_di_nuc}
count(ncrna[[2]], 2)
```

GC percent.

```{r gc_percent}
GC(ncrna[[2]])
```

Store entire FASTA headers.

```{r store_header}
my_header <- getAnnot(ncrna)
```

Create data frame using FASTA header.

```{r my_df}
head(my_header[-1])

my_split <- lapply(my_header[-1], function(x){
  str_split(x, " \\| ", simplify = TRUE)
})

my_df <- do.call(rbind.data.frame, my_split)

names(my_df) <- as.vector(str_split(my_header[[1]], " \\| ", simplify = TRUE))

my_df %>%
  rename(nncid = ">nncid") %>%
  mutate(nncid = sub("^>", "", nncid)) -> my_df

head(my_df)
```

Organisms with the most ncRNAs.

```{r organism_ncrna}
my_df %>%
  group_by(organism) %>%
  summarise(tally = n()) %>%
  arrange(desc(tally)) %>%
  head()
```

Class with the most ncRNAs.

```{r class_ncrna}
my_df %>%
  group_by(class) %>%
  summarise(tally = n()) %>%
  arrange(desc(tally)) %>%
  head()
```

Find all human piwi-interacting RNAs (piRNAs) and store their `nncid`.

```{r nncid_human_pirna}
my_df %>%
  filter(class == "piRNA", organism =="Homo sapiens") %>%
  pull(nncid) -> nncid_human_pirna

length(nncid_human_pirna)
```

Create `pirna_human` for storing human piRNAs.

```{r pirna_human}
pirna_human <- ncrna[nncid_human_pirna]
getSequence(pirna_human[[1]], as.string = TRUE)
```

Report number of sequences with N's and remove them.

```{r num_n}
my_n <- grepl('n', unlist(getSequence(pirna_human, as.string = TRUE)))
pirna_human <- pirna_human[!my_n]
table(my_n)
```

Save `pirna_human` as a FASTA file (not run).

```{r write_fasta, eval=FALSE}
my_anno <- getAnnot(pirna_human)
my_anno <- lapply(my_anno, function(x) sub("^>", "", x))
write.fasta(sequences = pirna_human, names = my_anno, file.out = "human_pirna.fa")
```

piRNAs typically start with U/T.

```{r first_base}
prop.table(table(sapply(pirna_human, function(x) x[1])))
```
In addition, piRNAs typically have an A at the tenth base and the proportion below is slightly higher than 0.25 if we expect an equal distribution.

```{r tenth_base}
prop.table(table(sapply(pirna_human, function(x) x[10])))
```

Length distribution of piRNAs.

```{r length_dist}
table(getLength(pirna_human))
```

piRNAs are typically 26-31 nucleotide long, as observed below.

```{r plot_length_dist}
as.data.frame(table(getLength(pirna_human))) %>%
  ggplot(., aes(Var1, Freq)) +
    geom_col() +
    labs(x = "piRNA length", y = "Frequency") +
    theme_bw()
```

10th base proportion for different lengths.

```{r tenth_base_per_length}
my_len <- 26:32
my_prop <- sapply(my_len, function(x){
  wanted <- getLength(pirna_human) == x
  prop.table(table(sapply(pirna_human[wanted], function(x) x[10])))
})

as.data.frame(t(my_prop), row.names = my_len)
```

Frequencies of nucleotides along every position.

```{r freq_nuc_along}
my_lens <- 26:32
for (my_len in my_lens){
  wanted <- getLength(pirna_human) == my_len
  my_seq <- getSequence(pirna_human[wanted])
  my_freq <- apply(do.call(rbind, my_seq), 2, function(x) prop.table(table(x)))
  
  as.data.frame(my_freq) %>%
    mutate(nuc = c('A', 'C', 'G', 'T')) %>%
    select(nuc, everything()) %>%
    pivot_longer(!nuc, names_to = "Position", values_to = "Frequency") %>%
    mutate(Position = as.integer(sub("^V", "", Position))) -> my_freq_df
  
  ggplot(my_freq_df, aes(Position, Frequency, colour = nuc)) +
    geom_line() +
    geom_point() +
    theme_bw() +
    ggtitle(paste0("Nucleotide frequencies along ", my_len, " nt human piRNAs")) -> p
  
  print(p)
}
```

Please refer to the [seqinr manual](https://cran.r-project.org/web/packages/seqinr/seqinr.pdf) for further information.
