---
title: "Ordering Objects using Seriation in R"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Installation

Install stable CRAN version.

```{r install}
if(! "seriation" %in% installed.packages()[, 1]){
  install.packages("seriation", repos = c("https://mhahsler.r-universe.dev", "https://cloud.r-project.org/"))
}

packageVersion("seriation")
```

## Getting started

Use the [example dataset](https://github.com/mhahsler/seriation#usage) `SupremeCourt`, which:

>Contains a (a subset of the) decisions for the stable 8-yr period 1995-2002 of the second Rehnquist Supreme Court. Decisions are aggregated to the joint probability for disagreement between judges.

```{r supreme_court}
library(seriation)
data("SupremeCourt")
SupremeCourt
```

Convert to distance matrix.

```{r distance_matrix}
d <- as.dist(SupremeCourt)
d
```

Perform the default seriation method to reorder the objects.

```{r default_seriation}
my_order <- seriate(d)
get_order(my_order)
```

Plot heatmap.

```{r heatmap}
p1 <- ggpimage(d, upper_tri = TRUE) +
  ggtitle("Judges (original alphabetical order)")

p2 <- ggpimage(d, my_order, upper_tri = TRUE) +
  ggtitle("Judges (reordered by seriation)")

p1 + p2 & scale_fill_gradientn(colours = c("darkgrey", "skyblue"))
```

Return linear configuration where more similar objects are located closer to each other.

```{r get_config}
get_config(my_order)
```

Plot linear configuration.

```{r plot_config}
plot_config(my_order)
```

## Heatmaps with seriation

The `Wood` dataset consists of:

>A data matrix containing a sample of the normalized gene expression data for 6 locations in the stem of Popla trees published in the study by Herzberg et al (2001). The sample of 136 genes selected by Caraux and Pinloche (2005).

```{r wood}
data("Wood")
dim(Wood)
```

Methods of interest for heatmaps are dendrogram leaf order-based methods applied to rows and columns. This is done using `method = "heatmap"`. The actual seriation method can be passed on as parameter seriaton_method, but it has a suitable default if it is omitted.

```{r wood_order}
wood_hc_complete <- seriate(Wood, method = "Heatmap", seriation_method = "HC_complete")
wood_olo_complete <- seriate(Wood, method = "Heatmap", seriation_method = "OLO_complete")
get_order(wood_hc_complete)
```

Ordering.

```{r comp_order}
get_order(wood_hc_complete, 2)
get_order(wood_olo_complete, 2)
```

Heatmap.

```{r wood_heatmap}
p1 <- ggpimage(Wood) +
  ggtitle("Wood (no order)") +
  theme(legend.position = "none")
  
p2 <- ggpimage(Wood, wood_hc_complete) +
  ggtitle("Wood (HC complete)") +
  theme(legend.position = "none")

p3 <- ggpimage(Wood, wood_olo_complete) +
  ggtitle("Wood (OLO complete)")

p1 + p2 + p3 & scale_fill_gradientn(colours = c("skyblue", "red"))
```
