---
title: "Saving Seurat objects"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install_bpcells, eval=FALSE}
if ("BPCells" %in% row.names(installed.packages()) == FALSE){
  remotes::install_github("bnprks/BPCells/r")
}
```

```{r load_libraries}
suppressPackageStartupMessages(library(BPCells))
suppressPackageStartupMessages(library(Seurat))
```

# Load Data

Load from my server.

```{r pbmc3k}
pbmc3k <- readRDS(url("https://davetang.org/file/pbmc3k_seurat.rds", "rb"))
pbmc3k
```

# Use BPCells

Sparse matrix.

```{r sparse_matrix}
class(pbmc3k@assays$RNA$counts)
```

Write a matrix directory and load the matrix using {BPCells}.

```{r bpcells_matrix}
my_outdir <- "data/pbmc3k_bpcells_mat"
if(!dir.exists(my_outdir)){
  BPCells::write_matrix_dir(
    mat = pbmc3k@assays$RNA$counts,
    dir = my_outdir
  )
}

# Now that we have the matrix on disk, we can load it
pbmc3k.mat <- open_matrix_dir(dir = my_outdir)
pbmc3k.mat
```

Create a Seurat object. 

```{r create_seurat_obj}
pbmc3k_bpcells <- CreateSeuratObject(
  counts = pbmc3k.mat,
  project = 'pbmc3k',
  min.cells = 3,
  min.features = 200
)

pbmc3k_bpcells@assays$RNA$counts
```

# Seurat version 4

Mitochondrial percent.

```{r mito_genes}
mito.genes <- grep(pattern = "^MT-", x = rownames(x = pbmc3k_bpcells@assays$RNA), ignore.case = TRUE, value = TRUE)

pbmc3k_bpcells[["percent.mt"]] <- PercentageFeatureSet(pbmc3k_bpcells, features = mito.genes)
VlnPlot(pbmc3k_bpcells, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, layer = "counts")
```

Save original data into `RAW` assay before filtering.

```{r raw_assay}
# build a new independent assay
raw_counts <- GetAssayData(pbmc3k_bpcells, layer = "counts", assay = "RNA")
raw_assay <- CreateAssayObject(counts = raw_counts)
pbmc3k_bpcells[["RAW"]] <- raw_assay

Seurat::Assays(pbmc3k_bpcells)
Seurat::DefaultAssay(pbmc3k_bpcells)
```

Filter.

```{r filter_mito}
pbmc3k_bpcells <- subset(pbmc3k_bpcells, subset = percent.mt < 15)
pbmc3k_bpcells
```

Unfortunately, the `RAW` assay becomes filtered as well.

```{r check_raw}
dim(pbmc3k_bpcells@assays$RNA$counts)
dim(pbmc3k_bpcells@assays$RAW$counts)
```

Seurat workflow.

```{r seurat_workflow}
debug_flag <- FALSE
start_time <- Sys.time()

pbmc3k_bpcells <- NormalizeData(pbmc3k_bpcells, normalization.method = "LogNormalize")
pbmc3k_bpcells <- FindVariableFeatures(pbmc3k_bpcells, selection.method = 'vst', nfeatures = 2000, verbose = debug_flag)
pbmc3k_bpcells <- ScaleData(pbmc3k_bpcells, verbose = debug_flag)
pbmc3k_bpcells <- RunPCA(pbmc3k_bpcells, verbose = debug_flag)
pbmc3k_bpcells <- RunUMAP(pbmc3k_bpcells, dims = 1:30, verbose = debug_flag)
pbmc3k_bpcells <- FindNeighbors(pbmc3k_bpcells, dims = 1:30, verbose = debug_flag)
pbmc3k_bpcells <- FindClusters(pbmc3k_bpcells, resolution = 0.5, verbose = debug_flag)
pbmc3k_bpcells

end_time <- Sys.time()
end_time - start_time
```

Counts.

```{r counts}
pbmc3k_bpcells@assays$RNA$counts
```

Data.

```{r data}
pbmc3k_bpcells@assays$RNA$data
```

Scale data.

```{r scale_data}
pbmc3k_bpcells@assays$RNA$scale.data
```

# Plots

```{r dim_plot}
DimPlot(pbmc3k_bpcells)
```

# Add miscellaneous data

[Get and set miscellaneous data](https://satijalab.github.io/seurat-object/reference/Misc.html).

```{r misc}
Misc(pbmc3k_bpcells)
```

Set and output.

```{r misc_set_and_get}
Misc(pbmc3k_bpcells, slot = "seed") <- 1984
Misc(pbmc3k_bpcells, slot = "author") <- "Davo"
Misc(pbmc3k_bpcells)
```

Get specific slot.

```{r author}
Misc(pbmc3k_bpcells, slot = "author")
```

# Exporting

Save.

```{r save_rds}
saveRDS(object = pbmc3k_bpcells, file = "pbmc3k_save_rds.rds")
```

Load.

```{r read_rds}
pbmc3k_read_rds <- readRDS("pbmc3k_save_rds.rds")
pbmc3k_read_rds@assays$RNA$counts
pbmc3k_read_rds@assays$RNA$data
pbmc3k_read_rds@assays$RNA$scale.data
pbmc3k_read_rds
```

Get miscellaneous data.

```{r author_after_loading}
Misc(pbmc3k_read_rds)
```

Check `RAW` assay.

```{r check_raw_after_loading}
pbmc3k_read_rds@assays$RAW$counts
dim(pbmc3k_read_rds@assays$RAW$counts)
dim(pbmc3k_read_rds@assays$RNA$counts)
```
