---
title: "Getting started with Seurat"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This post follows the Peripheral Blood Mononuclear Cells (PBMCs) [tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial) for 2,700 single cells. It was written while I was going through the tutorial and contains my notes. The dataset for this tutorial can be downloaded from the [10X Genomics dataset page](https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/pbmc3k) but it is also hosted on Amazon (see below). The PBMCs, which are primary cells with relatively small amounts of RNA (around 1pg RNA/cell), come from a healthy donor. There were 2,700 cells detected and sequencing was performed on an Illumina NextSeq 500 with around 69,000 reads per cell. To get started [install Seurat](https://satijalab.org/seurat/articles/install.html) by using install.packages().

```{r install_seurat, eval=FALSE}
install.packages("Seurat")
```

If you get the warning:

>‘SeuratObject’ was built under R 4.3.0 but the current version is 4.3.2; it is recomended that you
reinstall ‘SeuratObject’ as the ABI for R may have changed

re-install the `SeuratObject` package using a repository that has an updated copy. The same goes for the `htmltools` package.

```{r install_seurat_obj, eval=FALSE}
install.packages("SeuratObject", repos = "https://cran.ism.ac.jp/")
install.packages("htmltools", repos = "https://cran.ism.ac.jp/")
packageVersion("SeuratObject")
packageVersion("htmltools")
```

Load `Seurat`.

```{r load_seurat}
library("Seurat")
packageVersion("Seurat")
```

## Data

To follow the tutorial, you need the 10X data.

```{bash, eval=FALSE}
mkdir -p data/pbmc3k && cd data/pbmc3k
wget -c https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz
tar -xzf pbmc3k_filtered_gene_bc_matrices.tar.gz
```

The extracted files.

```{bash}
ls -1 data/pbmc3k/filtered_gene_bc_matrices/hg19
```

`matrix.mtx` is a [MatrixMarket](https://math.nist.gov/MatrixMarket/formats.html) file. It has the following properties:

* Only non-zero entries are stored in the file
* Comments start with a `%`, like LaTeX
* The first line indicates the total number of rows, columns, and entries
* The following lines after the first provide a row and column number and the value at that coordinate

```{bash}
head data/pbmc3k/filtered_gene_bc_matrices/hg19/matrix.mtx
```

## Seurat
