---
title: "Re-running Seurat"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# https://stackoverflow.com/questions/30237310/setting-work-directory-in-knitr-using-opts-chunksetroot-dir-doesnt-wor
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(Seurat)
```

If I re-run Seurat in the same manner with the same dataset, will I get identical results?

## Seurat object

Import [raw pbmc3k dataset](pbmc3k.html) from my server.

```{r seurat_obj}
seurat_obj <- readRDS(url("https://davetang.org/file/pbmc3k_seurat.rds", "rb"))
seurat_obj
```

Filter.

```{r pbmc3k}
pbmc3k <- CreateSeuratObject(
  counts = seurat_obj@assays$RNA$counts,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)
pbmc3k
```

## Seurat workflows

[Seurat workflows](seurat_v4_vs_v5.html) as functions.

```{r seurat_wfs}
seurat_wf_v4 <- function(seurat_obj, scale_factor = 1e4, num_features = 2000, num_pcs = 30, cluster_res = 0.5, debug_flag = FALSE){
  
  seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = scale_factor, verbose = debug_flag)
  seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = 'vst', nfeatures = num_features, verbose = debug_flag)
  seurat_obj <- ScaleData(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunPCA(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindClusters(seurat_obj, resolution = cluster_res, verbose = debug_flag)
  
  seurat_obj
}

seurat_wf_v5 <- function(seurat_obj, scale_factor = 1e4, num_features = 2000, num_pcs = 30, cluster_res = 0.5, debug_flag = FALSE){
  
  seurat_obj <- SCTransform(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunPCA(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindClusters(seurat_obj, resolution = cluster_res, verbose = debug_flag)
  
  seurat_obj
}
```

## First run

Process pbmc3k using the Seurat version 5 workflow.

```{r first_run}
pbmc3k_v5_1 <- seurat_wf_v5(pbmc3k)
pbmc3k_v5_1
```

## Second run

Process pbmc3k using the Seurat version 5 workflow again.

```{r second_run}
pbmc3k_v5_2 <- seurat_wf_v5(pbmc3k)
pbmc3k_v5_2
```

## Compare first and second runs

Compare UMAPs.

```{r compare_umaps}
DimPlot(pbmc3k_v5_1) + DimPlot(pbmc3k_v5_2)
```

Looks the same but let's double check.

```{r compare_umap_cell_embeddings}
identical(
  pbmc3k_v5_1@reductions$umap@cell.embeddings,
  pbmc3k_v5_2@reductions$umap@cell.embeddings
)
```

Compare clustering.

```{r compare_clustering}
identical(
  row.names(pbmc3k_v5_1@meta.data),
  row.names(pbmc3k_v5_2@meta.data)
)

identical(
  pbmc3k_v5_1@meta.data$seurat_clusters,
  pbmc3k_v5_2@meta.data$seurat_clusters
)
```

## Reorder

Use the same dataset but re-order the genes in the count matrix randomly.

```{r pbmc3k_reorder_genes}
my_mat <- seurat_obj@assays$RNA$counts
set.seed(1984)
row_order <- sample(rownames(my_mat))
my_mat <- my_mat[row_order, ]

stopifnot(all(rownames(my_mat) %in% rownames(seurat_obj@assays$RNA$counts)))
stopifnot(all(colnames(my_mat) %in% colnames(seurat_obj@assays$RNA$counts)))

pbmc3k_reorder_genes <- CreateSeuratObject(
  counts = my_mat,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

stopifnot(all(colnames(pbmc3k_reorder_genes@assays$RNA$counts) %in% colnames(pbmc3k@assays$RNA$counts)))
stopifnot(all(rownames(pbmc3k_reorder_genes@assays$RNA$counts) %in% rownames(pbmc3k@assays$RNA$counts)))

pbmc3k_reorder_genes
```

Use the same dataset but re-order the barcodes in the count matrix randomly.

```{r pbmc3k_reorder_cells}
my_mat <- seurat_obj@assays$RNA$counts
set.seed(1984)
col_order <- sample(colnames(my_mat))
my_mat <- my_mat[, col_order]

stopifnot(all(rownames(my_mat) %in% rownames(seurat_obj@assays$RNA$counts)))
stopifnot(all(colnames(my_mat) %in% colnames(seurat_obj@assays$RNA$counts)))

pbmc3k_reorder_cells <- CreateSeuratObject(
  counts = my_mat,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

stopifnot(all(colnames(pbmc3k_reorder_cells@assays$RNA$counts) %in% colnames(pbmc3k@assays$RNA$counts)))
stopifnot(all(rownames(pbmc3k_reorder_cells@assays$RNA$counts) %in% rownames(pbmc3k@assays$RNA$counts)))

pbmc3k_reorder_cells
```

Process the re-ordered pbmc3k dataset using the Seurat version 5 workflow again.

```{r seurat_v5_reordered}
pbmc3k_v5_genes <- seurat_wf_v5(pbmc3k_reorder_genes)
pbmc3k_v5_cells <- seurat_wf_v5(pbmc3k_reorder_cells)

pbmc3k_v5_1
pbmc3k_v5_cells
```

## Compare third run

Compare UMAPs.

```{r compare_umaps_3}
(DimPlot(pbmc3k_v5_1) + ggtitle("Re-ordered genes")) + DimPlot(pbmc3k_v5_genes)
(DimPlot(pbmc3k_v5_1) + ggtitle("Re-ordered cells")) + DimPlot(pbmc3k_v5_cells)
```

Compare clustering.

```{r compare_clustering_3}
compare_clustering <- function(obj1, obj2){
  idx <- match(row.names(obj1@meta.data), row.names(obj2@meta.data))
  stopifnot(row.names(obj1@meta.data) == row.names(obj2@meta.data)[idx])
  table(
    obj1@meta.data$seurat_clusters,
    obj2@meta.data$seurat_clusters[idx]
  )
}

compare_clustering(pbmc3k_v5_1, pbmc3k_v5_genes)
compare_clustering(pbmc3k_v5_1, pbmc3k_v5_cells)
compare_clustering(pbmc3k_v5_genes, pbmc3k_v5_cells)
```

### Use Seurat version 4

What if we used version 4?

```{r compare_seurat_version_4}
pbmc3k_v4_1 <- seurat_wf_v4(pbmc3k)
pbmc3k_v4_genes <- seurat_wf_v4(pbmc3k_reorder_genes)
pbmc3k_v4_cells <- seurat_wf_v4(pbmc3k_reorder_cells)

compare_clustering(pbmc3k_v4_1, pbmc3k_v4_genes)
compare_clustering(pbmc3k_v4_1, pbmc3k_v4_cells)
```

Re-ordering genes results in the same results using Seurat version 4 but different results when re-ordering cells.

### Same order

Since it seems a different order of barcodes/cells generates slightly different results, what if we used the same order? (The easiest way is to simply create a new Seurat object with the new order. Originally, I had tried to re-order an existing Seurat object but ended up creating an invalid Seurat object.)

```{r compare_same_order}
pbmc3k <- CreateSeuratObject(
  counts = seurat_obj@assays$RNA$counts,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

set.seed(1941)
rs <- sample(rownames(pbmc3k@assays$RNA$counts))
cs <- sample(colnames(pbmc3k@assays$RNA$counts))

pbmc3k_c <- CreateSeuratObject(
  counts = pbmc3k@assays$RNA$counts[rs, cs],
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

pbmc3k_d <- CreateSeuratObject(
  counts = pbmc3k_reorder_cells@assays$RNA$counts[rs, cs],
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

pbmc3k_v4_c <- seurat_wf_v4(pbmc3k_c)
pbmc3k_v4_d <- seurat_wf_v4(pbmc3k_d)

stopifnot(row.names(pbmc3k_v4_c@meta.data) == row.names(pbmc3k_v4_d@meta.data))
stopifnot(row.names(pbmc3k_v4_c@reductions$pca@cell.embeddings) == row.names(pbmc3k_v4_d@reductions$pca@cell.embeddings))
stopifnot(row.names(pbmc3k_v4_c@reductions$umap@cell.embeddings) == row.names(pbmc3k_v4_d@reductions$umap@cell.embeddings))

DimPlot(pbmc3k_v4_c, reduction = "pca") + DimPlot(pbmc3k_v4_d, reduction = "pca")
identical(
  pbmc3k_v4_c@reductions$pca@cell.embeddings,
  pbmc3k_v4_d@reductions$pca@cell.embeddings
)

DimPlot(pbmc3k_v4_c) + DimPlot(pbmc3k_v4_d)
identical(
  pbmc3k_v4_c@reductions$umap@cell.embeddings,
  pbmc3k_v4_d@reductions$umap@cell.embeddings
)

table(
  pbmc3k_v4_c@meta.data$seurat_clusters,
  pbmc3k_v4_d@meta.data$seurat_clusters
)
```

## Summary

Seurat version 5:

* Re-running Seurat on the same object produced the same UMAP and clustering results.
* Re-running Seurat on the same data but shuffled genes produced different results.
* Re-running Seurat on the same data but shuffled barcodes/cells produced different results.

[Tim](https://github.com/satijalab/seurat/issues/4192#issuecomment-797637532) answered with respect to re-ordering cells:

> The graph-based clustering algorithm (Louvain, SLM, Leiden, etc.) is non-deterministic. Identical results across different runs of the algorithm are obtained by setting the same random seed. Changing the order of the cells will change the order that nodes are visited during the local moving phase of the algorithm and will potentially change the final cluster identities of cells.

However it is concerning that re-ordering genes produced different results when using Seurat version 5.

Seurat version 4:

* Re-running Seurat on the same data but shuffled genes produced the same result.
* Re-running Seurat on the same data but shuffled barcodes/cells produced different results.

Differences can be mitigated if the same order of cells is used.
