---
title: "Re-running Seurat"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# https://stackoverflow.com/questions/30237310/setting-work-directory-in-knitr-using-opts-chunksetroot-dir-doesnt-wor
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(Seurat)
```

If I re-run Seurat in the same manner with the same dataset, will I get identical results?

## Seurat object

Import [raw pbmc3k dataset](pbmc3k.html) from my server.

```{r seurat_obj}
seurat_obj <- readRDS(url("https://davetang.org/file/pbmc3k_seurat.rds", "rb"))
seurat_obj
```

Filter.

```{r pbmc3k}
pbmc3k <- CreateSeuratObject(
  counts = seurat_obj@assays$RNA$counts,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)
pbmc3k
```

## Seurat workflows

[Seurat version workflows](seurat_v4_vs_v5.html) as functions.

```{r seurat_wfs}
seurat_wf_v4 <- function(seurat_obj, scale_factor = 1e4, num_features = 2000, num_pcs = 30, cluster_res = 0.5, debug_flag = FALSE){
  
  seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = scale_factor, verbose = debug_flag)
  seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = 'vst', nfeatures = num_features, verbose = debug_flag)
  seurat_obj <- ScaleData(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunPCA(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindClusters(seurat_obj, resolution = cluster_res, verbose = debug_flag)
  
  seurat_obj
}

seurat_wf_v5 <- function(seurat_obj, scale_factor = 1e4, num_features = 2000, num_pcs = 30, cluster_res = 0.5, debug_flag = FALSE){
  
  seurat_obj <- SCTransform(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunPCA(seurat_obj, verbose = debug_flag)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs, verbose = debug_flag)
  seurat_obj <- FindClusters(seurat_obj, resolution = cluster_res, verbose = debug_flag)
  
  seurat_obj
}
```

## First run

Process pbmc3k using the Seurat version 5 workflow.

```{r first_run}
pbmc3k_1 <- seurat_wf_v5(pbmc3k)
pbmc3k_1
```

## Second run

Process pbmc3k using the Seurat version 5 workflow again.

```{r second_run}
pbmc3k_2 <- seurat_wf_v5(pbmc3k)
pbmc3k_2
```

## Compare first and second runs

Compare UMAPs.

```{r compare_umaps}
DimPlot(pbmc3k_1) + DimPlot(pbmc3k_1)
```

Looks the same but let's double check.

```{r compare_umap_cell_embeddings}
identical(
  pbmc3k_1@reductions$umap@cell.embeddings,
  pbmc3k_2@reductions$umap@cell.embeddings
)
```

Compare clustering.

```{r compare_clustering}
identical(
  row.names(pbmc3k_1@meta.data),
  row.names(pbmc3k_2@meta.data)
)

identical(
  pbmc3k_1@meta.data$seurat_clusters,
  pbmc3k_2@meta.data$seurat_clusters
)
```

## Third run

Use the same dataset but re-order the count matrix randomly.

```{r pbmc3k_reordered}
my_mat <- seurat_obj@assays$RNA$counts
set.seed(1984)
col_order <- sample(colnames(my_mat))
row_order <- sample(rownames(my_mat))
my_mat <- my_mat[row_order, col_order]

pbmc3k_reordered <- CreateSeuratObject(
  counts = my_mat,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

stopifnot(all(colnames(pbmc3k_reordered@assays$RNA$counts) %in% colnames(pbmc3k@assays$RNA$counts)))
stopifnot(all(rownames(pbmc3k_reordered@assays$RNA$counts) %in% rownames(pbmc3k@assays$RNA$counts)))

pbmc3k_reordered
```

Process the re-ordered pbmc3k dataset using the Seurat version 5 workflow again.

```{r third_run}
pbmc3k_3 <- seurat_wf_v5(pbmc3k_reordered)
pbmc3k_3
```

## Compare third run

Compare UMAPs.

```{r compare_umaps_3}
DimPlot(pbmc3k_1) + DimPlot(pbmc3k_3)
```

Compare clustering.

```{r compare_clustering_3}
idx <- match(row.names(pbmc3k_1@meta.data), row.names(pbmc3k_3@meta.data))

stopifnot(row.names(pbmc3k_1@meta.data) == row.names(pbmc3k_3@meta.data)[idx])

table(
  pbmc3k_1@meta.data$seurat_clusters,
  pbmc3k_3@meta.data$seurat_clusters[idx]
)
```

### Use Seurat version 4

What if we used version 4?

```{r compare_seurat_version_4}
pbmc3k_a <- seurat_wf_v4(pbmc3k)
pbmc3k_b <- seurat_wf_v4(pbmc3k_reordered)

idx <- match(row.names(pbmc3k_a@meta.data), row.names(pbmc3k_b@meta.data))
stopifnot(row.names(pbmc3k_a@meta.data) == row.names(pbmc3k_b@meta.data)[idx])

table(
  pbmc3k_a@meta.data$seurat_clusters,
  pbmc3k_b@meta.data$seurat_clusters[idx]
)
```

### Same order

Since it seems a different order of genes and barcodes generates slightly different results, what if we used the same order? (The easiest way is to simply create a new Seurat object with the new order. Originally, I had tried to re-order an existing Seurat object but ended up creating an invalid Seurat object.)

```{r compare_same_order}
pbmc3k <- CreateSeuratObject(
  counts = seurat_obj@assays$RNA$counts,
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

set.seed(1941)
rs <- sample(rownames(pbmc3k@assays$RNA$counts))
cs <- sample(colnames(pbmc3k@assays$RNA$counts))

pbmc3k_c <- CreateSeuratObject(
  counts = pbmc3k@assays$RNA$counts[rs, cs],
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

pbmc3k_d <- CreateSeuratObject(
  counts = pbmc3k_reordered@assays$RNA$counts[rs, cs],
  min.cells = 3,
  min.features = 200,
  project = "pbmc3k"
)

pbmc3k_c <- seurat_wf_v4(pbmc3k_c)
pbmc3k_d <- seurat_wf_v4(pbmc3k_d)

stopifnot(row.names(pbmc3k_c@meta.data) == row.names(pbmc3k_d@meta.data))
stopifnot(row.names(pbmc3k_c@reductions$pca@cell.embeddings) == row.names(pbmc3k_d@reductions$pca@cell.embeddings))
stopifnot(row.names(pbmc3k_c@reductions$umap@cell.embeddings) == row.names(pbmc3k_d@reductions$umap@cell.embeddings))

DimPlot(pbmc3k_c, reduction = "pca") + DimPlot(pbmc3k_d, reduction = "pca")
identical(
  pbmc3k_c@reductions$pca@cell.embeddings,
  pbmc3k_d@reductions$pca@cell.embeddings
)

DimPlot(pbmc3k_c) + DimPlot(pbmc3k_d)
identical(
  pbmc3k_c@reductions$umap@cell.embeddings,
  pbmc3k_d@reductions$umap@cell.embeddings
)

table(
  pbmc3k_c@meta.data$seurat_clusters,
  pbmc3k_d@meta.data$seurat_clusters
)
```

## Summary

* Re-running Seurat on the same object produced the same UMAP and clustering results.
* Re-running Seurat on the same data but shuffled produced different results.

As answered by [Tim](https://github.com/satijalab/seurat/issues/4192#issuecomment-797637532):

> The graph-based clustering algorithm (Louvain, SLM, Leiden, etc.) is non-deterministic. Identical results across different runs of the algorithm are obtained by setting the same random seed. Changing the order of the cells will change the order that nodes are visited during the local moving phase of the algorithm and will potentially change the final cluster identities of cells.
