---
title: "Removing ambient RNA contamination with SoupX"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Introduction

In droplet-based single-cell RNA-seq experiments such as those produced by the 10x Genomics Chromium platform, every droplet that is sequenced contains some amount of cell-free mRNA floating in the input solution. These cell-free transcripts are released by cells that lysed prior to encapsulation and are then captured along with the mRNA from the live cell that the droplet is supposed to represent. The result is that the count matrix for any given cell is a mixture of two things: transcripts that genuinely originated in that cell, and a low-level background of transcripts that came from the ambient solution — the so-called "soup".

The [SoupX](https://github.com/constantAmateur/SoupX) package (Young MD, Behjati S. *SoupX removes ambient RNA contamination from droplet-based single-cell RNA sequencing data.* [GigaScience 9(12):giaa151, 2020](https://doi.org/10.1093/gigascience/giaa151)) provides a method for estimating and removing this ambient RNA contamination. The intuition behind SoupX is that we can use empty droplets — droplets that did not contain a cell, but which still received some volume of the input solution — to characterise the soup directly. Once we know what the soup looks like, we can ask, for each real cell, what fraction of its observed counts are best explained by soup contamination and subtract that contribution out.

The SoupX workflow has three logical steps:

1. **Estimate the soup profile.** Use the empty droplets in the raw 10x output to compute the relative abundance of every gene in the ambient solution.
2. **Estimate the contamination fraction (`rho`).** For each cell, work out what fraction of its counts came from the soup. `rho = 0` means the cell has no contamination; `rho = 1` means every UMI came from the soup. SoupX provides an automatic method (`autoEstCont`) and a manual method (using genes that you know cannot be expressed in particular cell types).
3. **Correct the count matrix.** Subtract the expected soup contribution from every cell's counts, distributing the removed counts across genes in proportion to how strongly each gene is implicated as soup.

The original [SoupX PBMC tutorial](https://cran.r-project.org/web/packages/SoupX/vignettes/pbmcTutorial.html) walks through this workflow on a small PBMC dataset that ships with the package. The tutorial uses `load10X()` against a Cell Ranger `outs/` directory, which automatically attaches Cell Ranger's clustering and dimension reduction; the cluster labels and UMAP coordinates needed by `setClusters()` and `setDR()` are therefore already present.

This notebook reproduces the same workflow on a publicly available 10x Genomics PBMC dataset (`SC3pv3_GEX_Human_PBMC`, distributed with Cell Ranger 7.0.1) using only the raw and filtered feature-barcode matrices in HDF5 format. Because we don't have a full Cell Ranger `outs/` directory, we cannot use `load10X()` — we load the two matrices manually, build the `SoupChannel` ourselves, and run a short Seurat pipeline to produce the clustering and UMAP that the SoupX tutorial would otherwise have inherited from Cell Ranger. The data files are tracked outside the repository; download them with `script/download.sh` before knitting this notebook.

## Packages

We need three packages: SoupX itself, Seurat for loading the 10x HDF5 files and doing a quick clustering pass (SoupX strongly recommends supplying clustering information), and ggplot2 for visualisations. We also load `future` so that Seurat's parallel-aware functions (such as `ScaleData`, `FindMarkers`, and `JackStraw`) can run on multiple workers, and `RhpcBLASctl` so we can give the BLAS-bound steps (`RunPCA`, distance computations, etc.) the same number of threads.

```{r packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(SoupX)
  library(Seurat)
  library(ggplot2)
  library(Matrix)
  library(future)
  library(RhpcBLASctl)
  library(hdf5r)
})
```

## Parallelisation

We will use four threads anywhere in this notebook that supports parallel execution. Seurat picks up parallelism from the active `future` plan, so a single `plan(multisession, workers = 4)` is enough to opt every `future`-aware Seurat function into using four background R sessions. A handful of Seurat / uwot / igraph functions read their thread count from a function argument or from the BLAS thread pool instead of from `future`; we set those explicitly when we get to them. SoupX itself does not support parallelism — `autoEstCont()` and `adjustCounts()` are single-threaded — so those steps are unaffected.

```{r parallel_setup}
plan(multisession, workers = 4)

options(future.globals.maxSize = 2 * 1024^3)

RhpcBLASctl::blas_set_num_threads(4)
RhpcBLASctl::omp_set_num_threads(4)
```

## Loading the 10x data

SoupX needs two count matrices to characterise the soup:

* The **table of counts (`toc`)** — the filtered matrix containing only droplets that Cell Ranger has classified as cells. This is what we will eventually correct.
* The **table of droplets (`tod`)** — the raw matrix containing every barcode that received any UMI at all, including the empty droplets. SoupX uses the empty droplets in this matrix to estimate the soup profile.

For 10x Cell Ranger output the convenience function `SoupX::load10X()` will read both matrices from a Cell Ranger `outs/` directory and additionally pull in any clustering and dimension-reduction information that Cell Ranger produced. We do not have access to a full `outs/` directory here — only the two HDF5 files — so we will load the matrices manually with `Seurat::Read10X_h5()` and assemble the `SoupChannel` ourselves.

```{r load_h5}
tod <- Seurat::Read10X_h5("data/SC3pv3_GEX_Human_PBMC_raw_feature_bc_matrix.h5")
toc <- Seurat::Read10X_h5("data/SC3pv3_GEX_Human_PBMC_filtered_feature_bc_matrix.h5")

dim(tod)
dim(toc)
```

The raw matrix has the same number of rows (genes) as the filtered matrix but many more columns (barcodes) because it includes all the empty droplets.

## Constructing the SoupChannel

A `SoupChannel` is the central data structure in SoupX: it is a list-like object that holds the filtered count matrix, the soup profile, and any clustering or dimension-reduction information you provide. It represents a single 10x channel — for multi-sample experiments you would build one `SoupChannel` per sample.

When you call `SoupChannel(tod, toc)`, the soup profile is calculated automatically by summing counts across the empty droplets in `tod`, and the raw droplet table is then dropped from memory. If you want to inspect the soup profile yourself or assemble it differently, you can pass `calcSoupProfile = FALSE` and call `estimateSoup()` manually later.

```{r soup_channel}
sc <- SoupChannel(tod, toc)
sc
```

The printout tells us how many genes and how many cells the channel covers. Note that "cells" here means every barcode in the filtered matrix — SoupX trusts Cell Ranger's cell calls.

## Clustering the cells

It is technically possible to run SoupX without clustering information, but the [SoupX vignette](https://cran.r-project.org/web/packages/SoupX/vignettes/pbmcTutorial.html) is emphatic that you will get far better results with clustering than without:

> While it is possible to run SoupX without clustering information, you will get far better results if some basic clustering is provided. Therefore, it is **strongly** recommended that you provide some clustering information to SoupX.

The reason is that the automatic contamination estimation works by looking for genes that are highly expressed in the soup but are clearly absent from particular cell populations. To do that, SoupX needs to know which cells belong to which population. SoupX is not sensitive to the exact clustering used — the default Seurat clustering is fine — so we will run a standard Seurat pipeline at low resolution to get cluster labels and a UMAP embedding to visualise things.

`ScaleData` is `future`-aware and will fan out across the four workers we configured above; `RunPCA` is dominated by BLAS calls and benefits from the four BLAS threads; `RunUMAP` uses [uwot](https://github.com/jlmelville/uwot) under the hood and exposes a `n.threads` argument that we set explicitly to 4. The remaining Seurat steps (`NormalizeData`, `FindVariableFeatures`, `FindNeighbors`, `FindClusters`) are either fast or single-threaded and are unaffected by the parallel configuration.

```{r seurat_pipeline, message=FALSE, warning=FALSE}
srat <- CreateSeuratObject(counts = toc)
srat <- NormalizeData(srat, verbose = FALSE)
srat <- FindVariableFeatures(srat, verbose = FALSE)
srat <- ScaleData(srat, verbose = FALSE)
srat <- RunPCA(srat, verbose = FALSE)
srat <- FindNeighbors(srat, dims = 1:30, verbose = FALSE)
srat <- FindClusters(srat, resolution = 0.5, verbose = FALSE)
srat <- RunUMAP(srat, dims = 1:30, n.threads = 4, verbose = FALSE)
```

Attach the cluster assignments and the UMAP coordinates to the `SoupChannel`. `setClusters()` expects a named vector mapping cell barcode to cluster label; `setDR()` expects a two-column data frame whose row names match the cell barcodes.

```{r set_clusters_dr}
sc <- setClusters(sc, setNames(Idents(srat), colnames(srat)))
sc <- setDR(sc, Embeddings(srat, "umap")[colnames(sc$toc), ])
```

A quick sanity-check plot of the clustering. SoupX does not need this plot for any computation, but it is useful for orienting yourself before looking at the soup-related diagnostics.

```{r umap_plot}
dr_cols <- sc$DR
dd <- data.frame(
  UMAP_1 = sc$metaData[[dr_cols[1]]],
  UMAP_2 = sc$metaData[[dr_cols[2]]],
  cluster = factor(sc$metaData$clusters)
)
ggplot(dd, aes(UMAP_1, UMAP_2, colour = cluster)) +
  geom_point(size = 0.3) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  ggtitle("PBMC clusters") +
  theme_minimal()
```

## Visualising the soup

Before estimating contamination fractions, it is informative to look at the soup profile directly. The soup profile is a probability distribution over genes telling us what the ambient solution looks like; a handful of genes typically dominate.

```{r soup_top_genes}
head(sc$soupProfile[order(sc$soupProfile$est, decreasing = TRUE), ], 20)
```

The genes at the top of this list are highly expressed in the soup, which usually means they are highly expressed in some abundant cell population whose mRNA leaked into the supernatant. In PBMC data you would typically expect to see haemoglobin genes (`HBB`, `HBA1`, `HBA2`) — released by lysed red blood cells contaminating the prep — and immunoglobulin genes (`IGKC`, `IGLC*`, `IGHA*`) — released by lysed plasma cells. These are exactly the genes whose expression in non-erythroid, non-plasma-cell clusters is most clearly contamination, which makes them ideal markers for inspecting the correction.

You can also visualise where the counts of a soup-implicated gene fall on the UMAP. A gene that is purely soup-derived will appear weakly and uniformly across all clusters; a gene with genuine cell-type-specific expression on top of soup contamination will be strong in its cell type and weak elsewhere.

```{r plot_marker_distribution, fig.width=8, fig.height=4}
plotMarkerDistribution(sc)
```

For each candidate gene, `plotMarkerDistribution()` shows the per-cell expression alongside a horizontal line indicating the level the soup alone would predict; genes that sit far above that line in some cells and at it in others are good candidates for either bimodal "expressed vs soup-only" classification (used by `autoEstCont`) or manual contamination estimation via `estimateNonExpressingCells()`.

## Estimating the contamination fraction

From SoupX 1.3.0 onwards, the recommended approach is to let `autoEstCont()` estimate the contamination fraction automatically. The function looks for sets of genes whose expression pattern across clusters is consistent with them being soup-only in some clusters, and uses those clusters to back out a contamination fraction `rho` for every cell.

```{r auto_est_cont, fig.width=6, fig.height=4}
sc <- autoEstCont(sc)
```

The diagnostic plot shows the distribution of estimated contamination fractions and the genes that the algorithm used. A typical PBMC dataset has a global `rho` around 0.05–0.10 (i.e. 5–10% of UMIs are soup), but values vary substantially between datasets and even between cells within a dataset. If `autoEstCont()` warns you that the contamination fraction is unusually high or that few marker genes were found, fall back to manual estimation with `estimateNonExpressingCells()` and `calculateContaminationFraction()` using genes that you know should not be expressed in particular cell types — see the SoupX vignette for the manual workflow.

## Correcting the count matrix

With `rho` estimated for every cell, `adjustCounts()` produces the corrected count matrix by subtracting the expected soup contribution from every cell. Counts are removed preferentially from genes that look most soup-like, in a way that respects the discrete-counting nature of the data.

```{r adjust_counts}
out <- adjustCounts(sc)
```

By default `adjustCounts()` returns non-integer counts. Most downstream tools — including Seurat — accept these without complaint, but if you have a pipeline that requires strictly integer counts you can pass `roundToInt = TRUE` (or wrap the call in `round()`).

```{r adjust_counts_int, eval=FALSE}
out <- adjustCounts(sc, roundToInt = TRUE)
```

## Sanity checking the correction

Two genes are particularly useful for sanity-checking the result on PBMC data:

* **Haemoglobin genes (`HBB`, `HBA1`, `HBA2`)** — released by lysed red blood cells. They should appear in the soup but should be essentially absent from every cluster after correction, because no PBMC type genuinely expresses them.
* **`IGKC` (immunoglobulin kappa constant)** — released by lysed plasma cells. It should be strongly expressed in the B-cell / plasma-cell cluster and largely removed from every other cluster.

The function `plotChangeMap()` overlays, on the dimension reduction we attached with `setDR()`, the fraction of observed counts that SoupX attributed to soup for a given gene in each cell — that is, `(observed - corrected) / observed`, the default `dataType = "soupFrac"`. A gene that was almost entirely soup will show a high soup fraction across the whole map; a gene with real cell-type-specific expression will show a high soup fraction in the clusters where it should not be expressed and a low soup fraction in the cluster where it genuinely is.

```{r change_map_hb, fig.width=6, fig.height=5}
plotChangeMap(sc, out, "HBB")
```

```{r change_map_igkc, fig.width=6, fig.height=5}
plotChangeMap(sc, out, "IGKC")
```

We can also tabulate the global before/after counts for these genes, summed across all cells, to see at a glance how much SoupX removed.

```{r before_after_table}
genes <- intersect(c("HBB", "HBA1", "HBA2", "IGKC", "IGLC2", "IGHA1"), rownames(toc))
data.frame(
  gene    = genes,
  before  = rowSums(toc[genes, ]),
  after   = rowSums(out[genes, ]),
  removed = rowSums(toc[genes, ]) - rowSums(out[genes, ])
)
```

## Using the corrected matrix downstream

The corrected matrix `out` can be used as a drop-in replacement for the original counts in any downstream workflow. For example, to feed it back into Seurat:

```{r downstream}
srat_clean <- CreateSeuratObject(counts = out)
srat_clean
```

You would then re-run normalisation, feature selection, dimension reduction, and clustering on the cleaned matrix. In most PBMC datasets the broad cluster structure is unchanged, but contamination-driven artefacts — for example, clusters that appeared to express haemoglobin or immunoglobulin genes purely because of soup — disappear, and downstream differential expression becomes substantially cleaner.

## Saving the corrected matrix

We save the corrected counts together with the per-cell contamination estimate and cluster label so that another notebook can compare these results to other ambient-RNA correction methods on the same dataset. See [ambient.html](ambient.html) for the side-by-side comparison with DecontX.

```{r save_output}
saveRDS(
  list(
    counts = out,
    meta   = data.frame(
      barcode = colnames(out),
      rho     = sc$metaData[colnames(out), "rho"],
      cluster = sc$metaData[colnames(out), "clusters"],
      row.names = colnames(out)
    )
  ),
  "output/soupx_corrected.rds"
)
```

## Cleaning up

Reset the `future` plan so the background R sessions are released.

```{r reset_plan}
plan(sequential)
```

## Session info

```{r session_info}
sessionInfo()
```
