---
title: "Split single column of annotations into multiple columns"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use [importbio](https://github.com/davetang/importbio) to load a VCF file (or load any file that has a single column containing multiple annotations that are consistently defined).

```{r importvcf}
library(importbio)
my_vcf <- importvcf("https://raw.githubusercontent.com/davetang/learning_vcf_file/master/eg/Pfeiffer.vcf")

head(my_vcf)
```

We will be functions within the `tidyverse`.

```{r tidyverse}
library(tidyverse)
```

Use `separate_rows` to create a new row for each annotation. The `mutate` call is to remove a trailing semicolon, if it exists.

```{r separate_rows}
library(tidyverse)
my_vcf %>%
  mutate(info = sub(pattern = ";$", replacement = "", x = .data$info)) %>%
  separate_rows(info, sep = ";\\s*") %>%
  head()
```

Use `separate` to split key-value annotation pair (separated by an equal sign) into two columns. Missing values will have a NA value.

```{r separate, warning=FALSE, message=FALSE}
my_vcf %>%
  mutate(info = sub(pattern = ";$", replacement = "", x = .data$info)) %>%
  separate_rows(info, sep = ";\\s*") %>%
  separate(info, c('key', 'value'), sep = "=") %>%
  head(10)
```

Finally, use `pivot_wider` to convert the data back into wide format. In the example below, I show three new columns created from splitting the `info` column.

```{r pivot_wider}
my_vcf %>%
  mutate(info = sub(pattern = ";$", replacement = "", x = .data$info)) %>%
  separate_rows(info, sep = ";\\s*") %>%
  separate(info, c('key', 'value'), sep = "=") %>%
  distinct() %>% # remove duplication annotations, if any
  filter(key == "AC" | key == "DP" | key == "MQ") %>%
  pivot_wider(id_cols = vid, names_from = key, values_from = value)
```

## Summary

1. Use `separate_rows` to split a single column with multiple annotations into rows
2. Use `separate` to split a key-value annotation pair into two columns
3. Use `pivot_wider` to convert the table in long format back to wide format
