---
title: "Trace"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

`?trace`:

> Interactive Tracing and Debugging of Calls to a Function or Method
>
> A call to trace allows you to insert debugging code (e.g., a call to browser or recover) at chosen places in any function. A call to untrace cancels the tracing. Specified methods can be traced the same way, without tracing all calls to the generic function. Trace code (tracer) can be any R expression. Tracing can be temporarily turned on or off globally by calling tracingState.

`trace()` is a built-in R function that lets you temporarily insert code into another function. It's useful for:

* Debugging
* Logging when a function is called
* Peeking into function arguments
* Finding out who called what, and when

It works even on functions inside packages without editing package source code.

Usage:

> trace(what, tracer, exit, at, print, signature,
      where = topenv(parent.frame()), edit = FALSE)
      
You can get started by setting two arguments:

* `what`: the function you want to trace
* `tracer`: the code you want to run when that function is called

## See when a function is called

The example below will print "sq was called!" when `sq()` is called. After printing, the function will operate as usual.

```{r sq}
sq <- function(x){
  x^2
}

trace("sq", tracer = quote(print("sq was called!")))

sq(4)
```

## See what arguments were passed

You can inspect the function's arguments inside the tracer by using `args()` or directly printing them.

```{r sq_args}
trace("sq", tracer = quote({
  print(paste("x is", x))
}))

sq(5)
```

## Trace a base R function

You can also trace functions from packages.

```{r mean}
trace("mean", tracer = quote(print("mean() was called!")))

mean(c(1, 2, 3))
```

Calling `mean()` from another function.

```{r my_mean}
trace("mean", tracer = quote(print("mean() was called!")))

my_mean <- function(x){
  mean(x)
}

my_mean(1:10)
```

Calling `base::mean()`.

```{r base_mean}
trace("mean", tracer = quote(print("mean() was called!")))

base::mean(1:10)
```

## Trace a function inside a package namespace

`trace("kmeans")` with a namespace.

```{r kmeans}
trace("kmeans", tracer = quote(print("stats::kmeans was called")), where = asNamespace("stats"))

k <- kmeans(iris[, 1:4], centers = 3)
```

Calling `stats::kmeans()`.

```{r stats_kmeans}
k <- stats::kmeans(iris[, 1:4], centers = 3)
```

## Removing a trace

When done debugging, remove the trace.

```{r untrace}
untrace("sq")
untrace("mean")
untrace("kmeans", where = asNamespace("stats"))
```
