---
title: "zellkonverter"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# https://stackoverflow.com/questions/30237310/setting-work-directory-in-knitr-using-opts-chunksetroot-dir-doesnt-wor
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

From [Conversion Between scRNA-seq Objects](https://www.bioconductor.org/packages/release/bioc/html/zellkonverter.html):

> Provides methods to convert between Python AnnData objects and SingleCellExperiment objects. These are primarily intended for use by downstream Bioconductor packages that wrap Python methods for single-cell data analysis. It also includes functions to read and write H5AD files used for saving AnnData objects to disk.

## Dependencies

Install packages using `BiocManager::install()`.

```{r install_dependencies, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("zellkonverter")
```

Load libraries.

```{r load_dependencies}
suppressPackageStartupMessages(library(zellkonverter))
```

## Data

Download data.

```{r pbmc3k_file}
pbmc3k_url <- 'https://davetang.org/file/pbmc3k/pbmc3k.h5ad'
pbmc3k_file <- paste0('data/', basename(pbmc3k_url))

if(!file.exists(pbmc3k_file)){
  download.file(url = pbmc3k_url, destfile = pbmc3k_file)
}
```

Read as `SingleCellExperiment` using `readH5AD()`, which uses [basilisk](https://www.bioconductor.org/packages/release/bioc/html/basilisk.html):

> + /home/rstudio/.cache/R/basilisk/1.20.0/0/bin/conda create --yes --prefix /home/rstudio/.cache/R/basilisk/1.20.0/zellkonverter/1.18.0/zellkonverterAnnDataEnv-0.11.4 'python=3.13.2' --quiet -c conda-forge --override-channels

```{r sce, error=TRUE}
sce <- readH5AD(pbmc3k_file)
sce
```
